---
title: "`admiral_df`: dataset diagnostics for admiral"
subtitle: "Issue #3160 proof of concept — what is implemented, how it works, and what persists"
author: "admiral development"
date: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    toc-expand: 2
    number-sections: true
    code-fold: show
    code-tools: true
    code-overflow: wrap
    df-print: kable
    theme:
      light: flatly
      dark: darkly
execute:
  warning: true
  message: true
  collapse: true
---

```{r}
#| label: setup
#| include: false
suppressMessages({
  library(dplyr)
  library(tidyr)
  library(tibble)
  library(cli)
  library(rlang)
})
# plain-text cli output so the rendered document is readable without ANSI codes
options(cli.num_colors = 1)
devtools::load_all(quiet = TRUE)
```

# What this document is

This is a developer-facing walkthrough of the `admiral_df` proof of concept for
[issue #3160](https://github.com/pharmaverse/admiral/issues/3160). Every code
chunk below **executes against the real functions** in `R/admiral_df.R` — re-render
this document to re-verify the behaviour it describes.

The core idea: a clinical programmer who has just run a derivation is asking two
questions —

1. **Did I create what I expected?** (counts, parameters, visits, treatment arms)
2. **Did I break anything?** (duplicated baselines, duplicated occurrence flags,
   records that shouldn't exist)

The second question is where the value is, and it is what a generic
`str()`/`skimr`-style summary cannot answer. The POC answers both with a single
call: `summary()` on a tagged dataset.

::: {.callout-note title="Design principle: diagnostics, not validation"}
Every check in this system is **reported, never signalled**. If `summary()` ever
errored, people would stop calling it, and it would become useless as a browsing
tool. It is deliberately not a replacement for `{diffdf}` or a study's QC
scripts.
:::

# The `admiral_df` class

## Two independent pieces of state

The whole system rests on two pieces of metadata attached to an ordinary
tibble, and it is essential to understand that they **travel independently**
(this becomes the crux of @sec-persistence):

| State | What it is | What it does |
|---|---|---|
| `admiral_df` **class** | Prepended to the class vector | Makes `summary()` dispatch to `summary.admiral_df()` instead of `summary.data.frame()` |
| `admiral_keys` **attribute** | Character vector of variable names | Declares the intended record structure — the variables the dataset should have exactly one record per |

`as_admiral_df()` is the low-level tagger. It is idempotent, cheap, and
`NULL`-safe (a `NULL` dataset passes through unchanged, which lets derivations
call it unconditionally on early returns):

```{r}
advs <- tribble(
  ~STUDYID,  ~USUBJID,      ~PARAMCD, ~AVAL, ~AVISIT,
  "PILOT01", "01-701-1015", "DIABP",     51, "BASELINE",
  "PILOT01", "01-701-1015", "SYSBP",    121, "BASELINE",
  "PILOT01", "01-701-1028", "DIABP",     79, "BASELINE",
  "PILOT01", "01-701-1028", "SYSBP",    130, "BASELINE"
)

tagged <- as_admiral_df(advs, keys = c("USUBJID", "PARAMCD"))
class(tagged)
attr(tagged, "admiral_keys")
```

Note the subtlety in the `keys` argument: `keys = NULL` (the default) **leaves
any existing attribute untouched** rather than clearing it. Re-tagging with
`as_admiral_df(x)` therefore restores the class but keeps whatever attribute was
already riding along — including a stale one. This is deliberate (it makes
re-tagging after `summarise()` cheap) but has consequences discussed in
@sec-persistence.

## Type detection: `get_admiral_df_type()`

`summary()` tailors its output to the ADaM structure of the dataset, detected
purely from the columns present. The precedence order matters because the
signals overlap (both BDS and TTE contain `PARAMCD`):

1. **`TTE`** — has `PARAMCD` + `CNSR` + `STARTDT`
2. **`OCCDS`** — no `PARAMCD`, but has a `--DECOD`/`--TERM` variable and/or
   `TRTEMFL`/`AOCCxxFL` (regex `"(DECOD|TERM)$|^AOCC.*FL$|^TRTEMFL$"`)
3. **`BDS`** — has `PARAMCD` + (`AVAL` or `AVALC`)
4. **`ADSL`** — no `PARAMCD`, and either one row per subject key **or** typical
   subject-level variables (`TRT01P`/`TRTSDT`)
5. **`other`** — none of the above

```{r}
c(
  bds   = get_admiral_df_type(advs),
  adsl  = get_admiral_df_type(tibble(USUBJID = "1", TRT01P = "Placebo")),
  tte   = get_admiral_df_type(tibble(PARAMCD = "OS", CNSR = 0, STARTDT = Sys.Date())),
  occds = get_admiral_df_type(tibble(USUBJID = "1", AEDECOD = "HEADACHE", ASEQ = 1L)),
  other = get_admiral_df_type(tibble(x = 1))
)
```

The ADSL fallback is worth spelling out: `is_adsl_structure()` accepts a
dataset as subject-level either **structurally** (already unique per
`get_admiral_option("subject_keys")`) or **semantically** (carries `TRT01P` or
`TRTSDT`). The semantic route matters because a *broken* ADSL — one with a
duplicated subject — must still be classified as ADSL, or the diagnostic could
never report the duplication.

# How the keys are resolved {#sec-keys}

This is the heart of the design, and the part with the most moving pieces. The
record-structure check needs to know *which* variables the dataset should be
unique by, and there are **three pathways** for that knowledge to arrive,
resolved in strict precedence order:

```{mermaid}
flowchart TD
    A["summary(x, keys = ?)"] --> B{"keys argument<br/>supplied?"}
    B -- "yes" --> S["<b>supplied</b><br/>character vector, or a Metacore<br/>object resolved via get_admiral_keys()<br/><i>transient — this call only</i>"]
    B -- "no" --> C{"'admiral_keys'<br/>attribute set?"}
    C -- "yes" --> D["<b>declared</b><br/><i>persistent — travels with the object</i>"]
    C -- "no (NULL)" --> I["<b>inferred</b><br/>infer_admiral_keys()<br/><i>heuristic, recomputed every call</i>"]
    D --> D1["set by metacore spec:<br/>set_admiral_keys(x, spec, 'ADVS')"]
    D --> D2["set by the user:<br/>as_admiral_df(x, keys = c(...))"]
    D --> D3["set by a derivation:<br/>derive_param_computed() stamps<br/>c(vars2chr(by_vars), 'PARAMCD')"]
```

The layering — *explicit argument overrides persistent metadata overrides
heuristic* — is a well-worn R idiom (`print(x, digits =)` over
`getOption("digits")`). Each pathway is examined below.

## Pathway 1 — declared via a metacore specification

`{metacore}` specs already know the intended keys of every dataset: the
`key_seq` column of `ds_vars`. Two functions expose this:

- **`get_admiral_keys(spec, "ADSL")`** — a thin wrapper around
  `metacore::get_keys()` that returns the key variable names as a plain
  character vector, in key order. It adds a `dataset_name = NULL` convenience
  for single-dataset specs (e.g. after `metacore::select_dataset()`), and warns
  when the spec defines no keys. Internally it must splice the dataset name in
  with `rlang::inject(... !!dataset_name)`, because `get_keys()` captures its
  argument via NSE and would otherwise filter for the literal symbol.
- **`set_admiral_keys(dataset, spec, "ADSL")`** — attaches those keys as the
  `admiral_keys` attribute, stores the dataset name in `admiral_ds_name` (shown
  in the summary heading), and tags the class.

`{metacore}` is a *suggested* dependency: it is only required at the moment
these functions are called (`rlang::check_installed()` prompts otherwise).

```{r}
#| eval: !expr 'requireNamespace("metacore", quietly = TRUE)'
load(metacore::metacore_example("pilot_ADaM.rda")) # loads `metacore`
spec <- metacore

get_admiral_keys(spec, "ADSL")

adsl_wip <- tibble(
  STUDYID = "PILOT01",
  USUBJID = c("01-701-1015", "01-701-1028"),
  TRT01P  = c("Placebo", "Xanomeline"),
  TRTSDT  = as.Date(c("2024-01-10", "2024-01-11"))
)

adsl_spec <- set_admiral_keys(adsl_wip, spec, "ADSL")
class(adsl_spec)
attr(adsl_spec, "admiral_keys")
attr(adsl_spec, "admiral_ds_name")
```

The attribute is exactly what the spec declares — nothing is inferred. `summary()`
then reports the structure **without** any annotation, because declared keys are
the trusted default:

```{r}
#| eval: !expr 'requireNamespace("metacore", quietly = TRUE)'
summary(adsl_spec)
```

::: {.callout-warning title="One documented asymmetry"}
`summary(x, keys = spec)` is **transient** while `set_admiral_keys(x, spec)` is
**persistent** — same spec, different lifetime. This mirrors
argument-vs-options semantics, but it is the one place a user could plausibly
expect the call to have tagged the object when it did not.
:::

## Pathway 2 — declared by the user or by a derivation

The user can stamp keys directly (`as_admiral_df(x, keys =)`), and — this is
the original #3160 story — **derivations stamp their own output**.
`derive_param_computed()` ends with:

```r
# the output has one record per `by_vars` + `PARAMCD`; record that so
# `summary()` can check the structure exactly instead of inferring it
as_admiral_df(
  bind_rows(dataset, hori_data),
  keys = c(vars2chr(by_vars), "PARAMCD")
)
```

The function that *creates* a dataset knows its intended structure better than
anyone downstream, and persisting that knowledge is what makes the
zero-ceremony post-derivation check work — including three pipeline steps
later:

```{r}
map <- derive_param_computed(
  advs,
  by_vars = exprs(USUBJID, AVISIT),
  parameters = c("SYSBP", "DIABP"),
  set_values_to = exprs(
    AVAL = (AVAL.SYSBP + 2 * AVAL.DIABP) / 3,
    PARAMCD = "MAP"
  )
)
attr(map, "admiral_keys")
summary(map)
```

Note the structure line has **no** `(inferred)` tag: the keys came from the
derivation's `by_vars`. And when a derivation genuinely goes wrong — say a
record is added twice — the same line becomes a finding:

```{r}
dup <- as_admiral_df(
  bind_rows(map, filter(map, PARAMCD == "MAP")[1, ]),
  keys = c("USUBJID", "AVISIT", "PARAMCD")
)
summary(dup)
```

## Pathway 3 — inferred from type and data

When no keys were supplied or declared, `infer_admiral_keys()` guesses. This is
the fallback, always annotated `(inferred)` in the output, and it is more
careful than a naive guess:

**Step 1 — semantic core by type.** ADSL starts from the subject keys; BDS and
TTE start from subject keys + `PARAMCD`; anything `other` gives up
(`character(0)`, no structure line at all).

**Step 2 — minimal expansion.** `minimal_unique_key()` adds standard ADaM key
variables one at a time, in precedence order (`AVISITN`, `AVISIT`, `ATPTN`,
`ATPT`, `ADTM`, `ADT`, `APERIOD`, `APERIODC`, `ASPID`), **only as far as needed**
to make the key unique. A single-visit dataset therefore reports the minimal key
without `AVISIT` padding; a multi-visit one automatically expands:

```{r}
# single visit: AVISIT is present but NOT needed for uniqueness
summary(as_admiral_df(advs))
```

```{r}
advs_2visits <- tribble(
  ~STUDYID,  ~USUBJID,      ~PARAMCD, ~AVAL, ~AVISIT,
  "PILOT01", "01-701-1015", "DIABP",     51, "BASELINE",
  "PILOT01", "01-701-1015", "DIABP",     50, "WEEK 2",
  "PILOT01", "01-701-1015", "SYSBP",    121, "BASELINE",
  "PILOT01", "01-701-1015", "SYSBP",    119, "WEEK 2"
)
# two visits: the key expands automatically until unique
summary(as_admiral_df(advs_2visits))
```

**Step 3 — the surrogate-key guardrail.** For BDS/TTE the candidate list
deliberately excludes sequence variables (`ASEQ`, `SRCSEQ`, ...). If `ASEQ`
were allowed in, uniqueness would be reached *trivially* — every row has its own
sequence number — and genuine duplicates would be masked:

```{r}
advs_aseq <- tribble(
  ~STUDYID,  ~USUBJID,      ~PARAMCD, ~AVAL, ~AVISIT,    ~ASEQ,
  "PILOT01", "01-701-1015", "MAP",       74, "BASELINE",  1L,
  "PILOT01", "01-701-1015", "MAP",       75, "BASELINE",  2L  # dup grain, new ASEQ
)
# ASEQ is excluded from inference, so the duplicate is still found
summary(as_admiral_df(advs_aseq))
```

**The OCCDS exception.** OCCDS has no analysis-value structure to fall back on:
multiple records per subject and term are legitimate (a subject can have the
same AE twice). There, `ASEQ` *is* the genuine record key, so it is **required**
rather than treated as a surrogate — and if it is absent, inference refuses to
guess and warns instead:

```{r}
adae <- tribble(
  ~STUDYID,  ~USUBJID,      ~AEDECOD,   ~TRTEMFL, ~ASEQ,
  "PILOT01", "01-701-1015", "HEADACHE", "Y",      1L,
  "PILOT01", "01-701-1015", "NAUSEA",   "Y",      2L,
  "PILOT01", "01-701-1028", "HEADACHE", "Y",      1L
)
summary(as_admiral_df(adae))          # grain = USUBJID + ASEQ
summary(as_admiral_df(select(adae, -ASEQ)))  # cannot check -> warning
```

## Why keep both the argument and the attribute?

Both the `keys =` argument and the `admiral_keys` attribute exist, which looks
redundant at first. The assessment (see `keys_design_notes.md`) is that they
solve **two different problems**:

- The **attribute is the producer's channel** — the derivation or the spec
  records the *intended* structure once, and every later `summary()` checks it
  with zero ceremony. Dropping it would kill the derivation tagging that
  motivated #3160.
- The **argument is the consumer's channel** — "is this dataset unique by *the
  keys I say*", explicit at the call site, no mutation of the object, and the
  escape hatch when the declared keys are wrong.

The genuine risk of dual pathways is comprehension ("why did summary check
*these* keys?"), and that is addressed by provenance: `key_source` is stored in
the returned object, and the printout annotates `(supplied)` and `(inferred)`
while leaving the trusted declared case unmarked.

# What `summary()` reports per type

Beyond the shared header (type, subjects, observations, structure check),
each type gets a tailored section, built up **additively**: an element is only
computed when the variables it needs exist and are not entirely `NA`, so a
mid-derivation dataset is summarized as far as it goes, and nothing it cannot
speak to is mentioned.

Most of what a summary reports is **facts** — counts, breakdowns, per-parameter
statistics, denominators. It also carries three **checks**, one for BDS, one
for OCCDS, and one for the ADSL comparison. Following the core team review on
2026-08-03, the checks are deliberately confined to a single family:

> A check belongs in {admiral} only if it is a **uniqueness or
> referential-integrity statement about keys and key-like flags** — the same
> kind of statement as the record structure check, applied to a subset of the
> records or across two datasets.

Data-quality checking beyond that (value domains, cross-variable consistency,
missingness, date ordering) is the job of a separate *ADaM checks* package.
{admiral} derives; it does not adjudicate whether the study's data is good.
Twelve such checks existed in the prototype and were removed.

The check-printing logic is shared (`print_summary_checks()`): a failing check
gets its own bullet with the offending count; passing checks collapse into
**one** confirmation line that still names each of them, so a clean dataset
does not cost a screen of zeroes but you can still see what was verified. A
check that could not run is *not mentioned at all* — not run is different from
passed.

::: {.panel-tabset}

## ADSL

ADSL is **facts only**. `summarize_adsl()` collects the denominators a
programmer checks first — treatment breakdown (preferring `TRT01P`, falling
back to `TRT01A` then `ARM`), population flag counts, the subject funnel
(randomized / treated / completed), treatment duration, end-of-study status,
deaths, demographics.

```{r}
adsl <- as_admiral_df(admiral_adsl, keys = "USUBJID")
summary(adsl)
```

It is the one type with no check section at all, and it is the clearest
illustration of where the 2026-08-03 line falls. The prototype checked `ARM`
against `ACTARM` here, and on the CDISC pilot data it flagged 12 subjects who
were planned for Xanomeline High Dose and actually took Low Dose. That is a
real dose deviation — but whether it is a finding or a documented feature of
the study is a clinical judgement about the study's data, not a statement about
whether a derivation produced the structure it intended. The same reasoning
retired the `SAFFL`-without-`TRTSDT`, `TRTEDT < TRTSDT`, and flag-domain
checks. All four are on the hand-over list for the ADaM checks package.

What remains is still enough to catch a lot at a glance: an arm with an
implausible N, a funnel where `treated` falls far below `randomized`, an
`EOSSTT` breakdown that does not add up.

And a partially built ADSL: every unavailable item is dropped rather than
reported as missing, so the summary stays readable mid-derivation.

```{r}
partial_adsl <- admiral_adsl %>%
  select(STUDYID, USUBJID, AGE, SEX, ARM) %>%
  as_admiral_df(keys = "USUBJID")
summary(partial_adsl)
```

## BDS

`summarize_bds()`'s centrepiece is the **per-parameter table** — records,
subjects, visits, missing `AVAL`, min/median/max per `PARAMCD` — replacing the
flat parameter list. Most BDS defects (empty parameter, missing-value spike,
absurd min/max) are visible as a row. The statistics expressions are assembled
conditionally (`stats$subjects <- quote(n_distinct(USUBJID))` only when
`USUBJID` exists) and spliced into a single `summarise(!!!stats)`, so the table
adapts to whatever columns are present. It also reports derived records by
`DTYPE` — exactly what the derivations added — and one check:

- 🚩 more than one `ABLFL == "Y"` per subject/parameter (and `BASETYPE`/`ATPT`
  where present) — the classic silent defect that corrupts `BASE`/`CHG`/`PCHG`

The baseline flag is expected to be unique within a subject and parameter, so
this is a uniqueness statement on a subset of the records — the record
structure check by another name, which is why it survived the 2026-08-03 trim
when the BDS value checks (`CHG` without `BASE`, `AVISIT` ↔ `AVISITN`, one
`PARAM` per `PARAMCD`) did not.

```{r}
adlb <- as_admiral_df(admiral_adlb)
summary(adlb)
```

A duplicated baseline injected. It changes no record count, no subject count,
and no record structure — the check is the only thing that tells the two
summaries apart:

```{r}
dup_base_row <- head(
  which(
    admiral_adlb$PARAMCD == "ALT" & is.na(admiral_adlb$ABLFL) &
      is.na(admiral_adlb$DTYPE) & admiral_adlb$USUBJID == admiral_adlb$USUBJID[1]
  ),
  1
)

broken_adlb <- admiral_adlb %>%
  mutate(ABLFL = if_else(row_number() == dup_base_row, "Y", ABLFL)) %>%
  as_admiral_df()

summary(broken_adlb)
```

And the original #3160 use case — summarizing a derivation's output directly.
The new `ASTALT` row in the parameter table shows what was added, with its
value range as a sanity check:

```{r}
ratio <- admiral_adlb %>%
  filter(is.na(DTYPE)) %>%
  derive_param_computed(
    by_vars = exprs(USUBJID, AVISIT, AVISITN),
    parameters = c("AST", "ALT"),
    set_values_to = exprs(
      AVAL = AVAL.AST / AVAL.ALT,
      PARAMCD = "ASTALT",
      PARAM = "AST/ALT ratio"
    )
  )
summary(ratio)
```

## OCCDS

`summarize_occds()` reports distinct terms per coding level
(`--BODSYS`/`--DECOD`/`--TERM`), treatment-emergent vs total records (the
denominator shift — most AE outputs are restricted to `TRTEMFL == "Y"`),
severity/grade (preferring the analysis version `ASEV` over `AESEV`, and
severity over toxicity grade), serious count, and one check:

- 🚩 **occurrence flag integrity** — the grouping level is derived from the
  flag's infix (the part between `AOCC` and `FL`): `S` → per subject and body
  system, `P` → per subject and dictionary term, anything else (`AOCCFL`,
  `AOCC02FL`, ...) → per subject. Only *more than one* `"Y"` per group is a
  defect — zero is legitimate, because the flags are typically restricted to
  the emergent records. This is the classic OCCDS defect: a duplicated flag
  double-counts the subject in every "n (%) of subjects" table downstream, and
  the record-structure check cannot see it because the record count is
  unchanged.

`APERIOD` joins the grouping for every flag when the dataset has it. Vaccine
studies and other multi-period designs flag the first occurrence once per
subject *per period*, which is correct and must not be reported as a
duplicate — and the flag name cannot express it, since `AOCC01FL` is a sponsor
numbering rather than a level. In a single-period study `APERIOD` is constant,
so nothing changes.

Like the BDS baseline check, this is a uniqueness statement about a flag within
a group, which is what kept it on 2026-08-03 when the OCCDS record-level date
checks (missing `ASTDT`, treatment-emergent records starting before `TRTSDT`)
were removed.

```{r}
#| eval: !expr 'requireNamespace("pharmaversesdtm", quietly = TRUE)'
adae_full <- pharmaversesdtm::ae %>%
  convert_blanks_to_na() %>%
  derive_vars_merged(
    dataset_add = admiral_adsl,
    new_vars = exprs(TRTSDT, TRTEDT, TRT01A),
    by_vars = exprs(STUDYID, USUBJID)
  ) %>%
  derive_vars_dt(new_vars_prefix = "AST", dtc = AESTDTC, highest_imputation = "M") %>%
  derive_vars_dt(new_vars_prefix = "AEN", dtc = AEENDTC) %>%
  derive_var_trtemfl(
    start_date = ASTDT, end_date = AENDT,
    trt_start_date = TRTSDT, trt_end_date = TRTEDT
  ) %>%
  mutate(ASEQ = AESEQ) %>%
  restrict_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      by_vars = exprs(USUBJID), order = exprs(ASTDT, AESEQ),
      new_var = AOCCFL, mode = "first"
    ),
    filter = TRTEMFL == "Y"
  ) %>%
  restrict_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      by_vars = exprs(USUBJID, AEBODSYS), order = exprs(ASTDT, AESEQ),
      new_var = AOCCSFL, mode = "first"
    ),
    filter = TRTEMFL == "Y"
  ) %>%
  as_admiral_df(keys = c("USUBJID", "ASEQ"))

summary(adae_full)
```

```{r}
#| eval: !expr 'requireNamespace("pharmaversesdtm", quietly = TRUE)'
dup_row <- head(which(adae_full$TRTEMFL == "Y" & is.na(adae_full$AOCCFL)), 1)

broken_adae <- adae_full %>%
  mutate(AOCCFL = if_else(row_number() == dup_row, "Y", AOCCFL)) %>%
  as_admiral_df(keys = c("USUBJID", "ASEQ"))

summary(broken_adae)
```

## TTE

TTE currently gets the shared header plus events vs censored (`CNSR == 0` vs
otherwise). The design notes flag that this should become **per-`PARAMCD`**
counts (each parameter is a separate endpoint, so a single total is misleading
with more than one endpoint) — that increment is not yet implemented.

```{r}
adtte <- tribble(
  ~STUDYID,  ~USUBJID,      ~PARAMCD, ~CNSR, ~STARTDT,
  "PILOT01", "01-701-1015", "OS",         0, as.Date("2024-01-10"),
  "PILOT01", "01-701-1028", "OS",         1, as.Date("2024-01-11")
)
summary(as_admiral_df(adtte))
```

:::

# Comparing a dataset against ADSL {#sec-vs-adsl}

For non-ADSL types, `summary()` accepts an **optional** `adsl =` argument:

```r
summary(object, keys = NULL, adsl = NULL, ...)
```

ADSL is the subject-level source of truth, and a record-level dataset can be
wrong *relative to it* in ways it can never reveal on its own. Three things
only ADSL can provide:

1. **The full subject universe** — a child dataset cannot know who is
   *missing* from it, nor supply the denominator for any "n (%) of subjects"
   statement.
2. **The authoritative subject-level values** — `TRTSDT`, `TRT01A`, `SAFFL` in
   a child dataset are *copies* (via `derive_vars_merged()`), and copies go
   stale when ADSL is refreshed after the child was built.
3. **Subject-level events the child must respect** — records dated after death
   are defects the child dataset cannot detect alone.

When `adsl` is supplied, `summarize_vs_adsl()` adds a comparison section;
when it is not, the section is absent entirely — not run is different from
passed, as everywhere else. Everything is one `USUBJID`-level join plus
vectorized comparisons, so it stays cheap on large datasets.

## Facts: coverage and denominators

The facts are the denominators programmers open ADSL for: overall coverage,
the safety-population denominator, and per-arm coverage (an empty or thin arm
is the shape of every downstream incidence table):

```{r}
adsl_ref <- tribble(
  ~USUBJID,      ~SAFFL, ~TRT01A,      ~TRTSDT,               ~DTHDT,
  "01-701-1015", "Y",    "Placebo",    as.Date("2024-01-01"), as.Date(NA),
  "01-701-1028", "Y",    "Xanomeline", as.Date("2024-01-02"), as.Date(NA),
  "01-701-1034", "N",    "Placebo",    as.Date(NA),           as.Date(NA)
)

summary(as_admiral_df(advs), adsl = adsl_ref)
```

## The red flag: orphan subjects

One check runs through the same `print_summary_checks()` machinery as the
per-type ones — a failure gets its own bullet, a pass collapses into one line,
and if `USUBJID` is missing on either side it is not mentioned:

- 🚩 **Orphan subjects** — in the dataset but not in ADSL (wrong ADSL version,
  screen failures or test subjects left in, or `USUBJID` formatting drift —
  which *presents* as orphans, so the failure bullet includes example IDs and
  the full vector is stored in `$vs_adsl$orphans`).

A subject key with nothing to point at is a referential-integrity failure: the
cross-dataset form of the record structure check, and always a defect. That is
why it stayed on 2026-08-03 while the *value* comparisons implemented alongside
it did not — stale shared variables (`TRTSDT`, `TRT01A`, demographics and the
rest compared per subject against ADSL), records dated after `DTHDT`, and
treatment-emergent records for subjects ADSL says were never treated. All three
are strong findings and all three belong in the ADaM checks package, where the
`adsl =` argument's shape is the obvious thing to reuse.

```{r}
adae_small <- as_admiral_df(
  tribble(
    ~USUBJID,      ~ASEQ, ~AEDECOD,    ~TRTEMFL, ~ASTDT,                ~TRTSDT,
    "01-701-1015",    1L, "HEADACHE",  "Y",      as.Date("2024-01-10"), as.Date("2024-01-01"),
    "01-701-1028",    1L, "NAUSEA",    "Y",      as.Date("2024-03-10"), as.Date("2024-01-02"),
    "01-701-1034",    1L, "FATIGUE",   "Y",      as.Date("2024-01-20"), as.Date(NA),
    # an orphan subject: not in ADSL at all
    "01-701-9999",    1L, "DIZZINESS", "Y",      as.Date("2024-01-15"), as.Date(NA)
  ),
  keys = c("USUBJID", "ASEQ")
)

summary(adae_small, adsl = adsl_ref)
```

Note the OCCDS summary also gained the **incidence line** ("subjects with a
treatment-emergent record: n of N"), preferring the `SAFFL == "Y"` denominator
and falling back to all of ADSL — the number every AE table leads with, which
the child dataset alone cannot compute because subjects with zero occurrences
are not in it.

## Argument semantics

Two deliberate boundaries, mirroring the design-notes decisions:

- **Malformed `adsl` aborts.** The data checks never signal, but *argument*
  validation may: an `adsl` without `USUBJID`, or with more than one record
  per subject, makes every comparison ambiguous — that is caller error, and
  `summary()` errors informatively rather than reporting nonsense.
- **`adsl` on an ADSL is ignored with a warning.** Comparing an ADSL against
  itself is meaningless; comparing two ADSL versions is a different feature
  (`{diffdf}` territory).

```{r}
#| error: true
summary(
  as_admiral_df(advs),
  adsl = tibble(USUBJID = c("1", "1"), SAFFL = "Y")  # duplicate subject
)
```

# Persistence: when does the tag survive a pipeline? {#sec-persistence}

This section matters most for the go-forward decision. The class and the
attribute are ordinary R metadata, and their survival through a pipeline is a
property of **dplyr, not of admiral** — the behaviour below was verified with
dplyr 1.2.1 / tidyr 1.3.2 and can change with dplyr releases.

## The mechanism: `dplyr_reconstruct()`

Most dplyr verbs rebuild their result with `dplyr::dplyr_reconstruct()`, whose
`data.frame` method copies **all** attributes from the template object —
including a custom class and `admiral_keys`. This is the opposite of the common
assumption that attributes never survive dplyr. The tag is lost only when:

1. an operation builds a **new** data frame from scratch (`summarise()`,
   pivots, `nest()`), or
2. the template it reconstructs from is a different, *untagged* object
   (`bind_rows(plain, tagged)`, joins where the tagged dataset is `y`).

```{r}
#| code-fold: true
#| code-summary: "helper: tag_status() — records what is left of the tag"
tag_status <- function(label, data) {
  tibble(
    operation = label,
    class_kept = inherits(data, "admiral_df"),
    keys_kept = !is.null(attr(data, "admiral_keys"))
  )
}

advs_tagged <- as_admiral_df(advs, keys = c("USUBJID", "PARAMCD"))

# a genuinely plain row: neither the class nor the attribute
plain_row <- tribble(
  ~STUDYID,  ~USUBJID,      ~PARAMCD, ~AVAL, ~AVISIT,
  "PILOT01", "01-701-1015", "MAP",       74, "BASELINE"
)
```

## Survives, drops, and the treacherous middle

```{r}
#| code-fold: true
#| code-summary: "code: run every operation and collect the results"
survives <- bind_rows(
  tag_status("mutate()", mutate(advs_tagged, AVALC = as.character(AVAL))),
  tag_status("filter()", filter(advs_tagged, AVAL > 60)),
  tag_status("arrange()", arrange(advs_tagged, desc(AVAL))),
  tag_status("select()", select(advs_tagged, USUBJID, PARAMCD, AVAL)),
  tag_status("slice() / head() / [i, ]", slice(advs_tagged, 1:2)),
  tag_status("distinct()", distinct(advs_tagged, USUBJID, PARAMCD, AVAL)),
  tag_status("count()", count(advs_tagged, USUBJID)),
  tag_status("bind_cols()", bind_cols(advs_tagged, tibble(FLAG = rep("Y", 4)))),
  tag_status("bind_rows(tagged, plain)", bind_rows(advs_tagged, plain_row)),
  tag_status(
    "left_join(tagged, plain)",
    left_join(advs_tagged, tibble(USUBJID = "01-701-1015", TRT01P = "Placebo"),
      by = "USUBJID"
    )
  ),
  tag_status("tidyr::fill()", tidyr::fill(advs_tagged, AVAL)),
  tag_status("saveRDS() -> readRDS()", {
    f <- tempfile(fileext = ".rds")
    saveRDS(advs_tagged, f)
    out <- readRDS(f)
    unlink(f)
    out
  })
)

drops_both <- bind_rows(
  tag_status("summarise()", summarise(advs_tagged, N = n(), .by = USUBJID)),
  tag_status("group_by() |> ungroup()", ungroup(group_by(advs_tagged, USUBJID))),
  tag_status(
    "tidyr::pivot_wider()",
    pivot_wider(advs_tagged, names_from = PARAMCD, values_from = AVAL)
  ),
  tag_status("tidyr::pivot_longer()", pivot_longer(advs_tagged, AVAL)),
  tag_status("nest() |> unnest()", unnest(nest(advs_tagged, data = AVAL), data)),
  tag_status("tidyr::complete()", complete(advs_tagged, USUBJID, PARAMCD)),
  tag_status("bind_rows(plain, tagged)", bind_rows(plain_row, advs_tagged)),
  tag_status(
    "left_join(plain, tagged)",
    left_join(tibble(USUBJID = c("01-701-1015", "01-701-1028")), advs_tagged,
      by = "USUBJID"
    )
  )
)

partial <- bind_rows(
  tag_status("group_by()", group_by(advs_tagged, USUBJID)),
  tag_status("rowwise()", rowwise(advs_tagged)),
  tag_status("as_tibble()", as_tibble(advs_tagged)),
  tag_status("as.data.frame()", as.data.frame(advs_tagged)),
  tag_status("tidyr::drop_na()", drop_na(advs_tagged))
)
```

::: {.panel-tabset}

### ✅ Keeps both

The everyday derivation pipeline keeps its tag end to end:

```{r}
#| echo: false
survives
```

### ❌ Drops both

Operations that build a new frame from scratch — and, critically, **argument
order**: `bind_rows()` reconstructs from its *first* argument and joins from
`x`, so the same data with the arguments swapped loses the tag:

```{r}
#| echo: false
drops_both
```

### ⚠️ Partial drops

The two halves are **not always lost together**. These keep `admiral_keys` but
drop the class:

```{r}
#| echo: false
partial
```

Two distinct failure modes fall out of this:

- **Class gone, attribute kept** (`as_tibble()`, `as.data.frame()`, and —
  *inside* the grouped block — `group_by()`): `summary()` dispatches to
  `summary.data.frame()`, so the admiral diagnostic disappears **without any
  error**. This is the dangerous case, because re-tagging with a bare
  `as_admiral_df(x)` restores the class but silently promotes the possibly-stale
  orphaned attribute back to "declared".
- **`group_by()` specifically** replaces the class vector outright, so
  `inherits(x, "admiral_df")` is already `FALSE` inside the grouped block;
  `ungroup()` then removes `admiral_keys` as well.

:::

## Stale keys: preserved, and therefore able to go wrong

The keys are stored as plain variable names, so `rename()` and `select()`
preserve the attribute **without updating it**:

```{r}
renamed <- rename(advs_tagged, SUBJID = USUBJID)
colnames(renamed)
attr(renamed, "admiral_keys")   # still says USUBJID
```

`summary()` refuses to let this pass silently. A declared (or supplied) key
that is absent from the dataset triggers a warning explaining exactly how the
check was weakened — without it, the partially stale case would check the grain
on `PARAMCD` alone and report the verdict as a declared one, inventing
duplicates here and just as easily hiding real ones with different data:

```{r}
summary(renamed)                      # partially stale: checked on PARAMCD only
```

```{r}
res <- summary(select(renamed, -PARAMCD))  # fully stale: nothing left to check
print(res)
```

Inferred keys never trigger this warning — they are derived from the columns
actually present.

## The three states of `admiral_keys`

A subtlety that any go-forward design must handle: the attribute has **three
distinct states** that print almost identically but behave differently in
`summary.admiral_df()`:

| State | `summary()` behaviour |
|---|---|
| `NULL` (no attribute) | Falls back to `infer_admiral_keys()` — structure still checked, marked `(inferred)` |
| `character(0)` (attribute set but empty) | Inference fallback is **skipped** *and* there is no key to check — **no structure check at all, silently**. `set_admiral_keys()` produces exactly this when the spec defines no `key_seq`. |
| variable names | Checked as declared — but only for the names still present (see stale keys above) |

Beware also that `all(NULL %in% cols)` is `TRUE` in R (`all(logical(0))`), so a
naive "are the keys present?" check reports success for a dataset that declares
no keys at all.

## `check_admiral_df()`: auditing the tag

Because the class and the attribute travel independently,
`inherits(x, "admiral_df")` alone is not evidence that the structure will be
checked — and `admiral_keys` alone is not evidence that `summary()` will even
dispatch. `check_admiral_df()` is the exported reporter for exactly this: it
mirrors the key resolution of `summary.admiral_df()` (including the
`NULL`-vs-`character(0)` distinction) without emitting any of its warnings, and
returns a programmable `admiral_df_check` object.

```{r}
check_admiral_df(advs_tagged)          # fully tagged
```

```{r}
check_admiral_df(renamed)              # declared keys gone stale
```

```{r}
check_admiral_df(as_tibble(advs_tagged))  # the orphaned-attribute case
```

The result is usable programmatically, e.g. as a guard in a pipeline:

```{r}
check_admiral_df(advs_tagged)$structure_checked
```

## Safe patterns

Distilled from the above, the guidance for pipelines today:

1. **Tag last** — run `as_admiral_df()` / `set_admiral_keys()` after reshaping
   is done, not before.
2. **Re-tag after** `summarise()`, `ungroup()`, `pivot_*()` and friends —
   `as_admiral_df()` is idempotent and cheap.
3. **Keep the tagged dataset first** — first argument of `bind_rows()`, the `x`
   of a join.
4. **Re-tag with explicit `keys`** after `rename()` of a key variable, and
   after anything that orphaned the attribute — a bare `as_admiral_df(x)`
   restores the class but keeps the stale attribute.

::: {.callout-important title="Why persistence is the go-forward question"}
The value proposition of the declared-keys pathway — *tag once, check anywhere*
— depends entirely on the attribute actually surviving the pipelines users
write. The evidence above says it survives the everyday derivation verbs but
not reshaping, and it can go stale or become orphaned in ways that fail
silently unless mitigated. The current mitigations are the stale-key warning
in `summary()` and `check_admiral_df()`. Options for going further include:
having *every* `derive_*` function re-tag its output (extending the
`derive_param_computed()` pattern, which would make the tag self-healing across
admiral pipelines); implementing `dplyr` group/rowwise S3 methods for
`admiral_df` to survive `group_by()`; or accepting the current behaviour and
documenting the safe patterns. This decision shapes how much users can trust
the declared pathway versus how often they fall back to supplied or inferred
keys.
:::

# Current status and open questions

**Implemented** (this branch, `3160_poc_liam`):

- `as_admiral_df()`, `get_admiral_df_type()`, `is_adsl_structure()`
- `infer_admiral_keys()` + `minimal_unique_key()` with the surrogate-key
  guardrail and the OCCDS/`ASEQ` special case
- `get_admiral_keys()` / `set_admiral_keys()` — the metacore pathway
- `summary.admiral_df()` with the supplied > declared > inferred precedence,
  `rlang::check_dots_empty()`, stale-key warnings, and `key_source` provenance
- Per-type content: ADSL (facts only), BDS (per-parameter table, `DTYPE`
  breakdown + the baseline uniqueness check), OCCDS (terms, emergent, severity
  + the occurrence flag check), TTE (events/censored)
- `check_admiral_df()` + printers; `derive_param_computed()` tags its output
- The `adsl =` argument (@sec-vs-adsl): coverage/denominator facts, OCCDS
  incidence, and the orphan subject check (`summarize_vs_adsl()`)

**Decided 2026-08-03** (core team review): data-quality checking is out of
scope for {admiral} and belongs in a separate ADaM checks package. Twelve of
the prototype's fifteen checks were removed; the three that remain are
uniqueness or referential-integrity statements about keys and key-like flags.
The removed ones are catalogued in `summary_admiral_df_design_notes.md` and
`adsl_comparison_design_notes.md` as the hand-over specification, and remain
recoverable from this branch's history.

**Open questions** (from the design notes):

1. **TTE per-parameter counts** — the current single events/censored total is
   misleading with multiple endpoints, and TTE is the one type with no
   per-type section. The largest remaining gap.
2. **ADSL linkage, next increments** — the fact half: a `population =`
   argument for denominators, per-parameter coverage in the BDS table, and a
   session-wide `set_admiral_options(adsl =)` channel.
3. **`metacore =` as a full argument** — beyond keys: mandatory variables from
   `ds_vars` not yet present, expected vs actual `PARAMCD` from `value_spec`.
4. **Verbosity** — the full wishlist runs past one screen; likely a
   `print(x, detail =)` gate with everything always computed into the object.
5. **Persistence strategy** — see @sec-persistence; this is the decision that
   determines how trustworthy the declared pathway is in real pipelines.
