---
title: "Get Started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Main Idea

The main idea of `{admiral}` is that an ADaM dataset is built by a sequence of
derivations. Each derivation adds one or more variables or parameters to the
processed dataset. This modular approach makes it easy to adjust code by adding,
removing, or modifying derivations. Each derivation is a function call. Consider
for example the following script which creates a (very simple) ADSL dataset.

```{r, message=FALSE, warning=FALSE}
library(dplyr)
library(lubridate)
library(admiral)
library(admiraltest)

# read in SDTM datasets
data("admiral_dm")
data("admiral_ds")
data("admiral_ex")

dm <- admiral_dm
ds <- admiral_ds
ex <- admiral_ex
```

```{r, eval=TRUE}
# derive treatment variables
adsl <- dm %>%
  mutate(TRT01P = ARM, TRT01A = ACTARM) %>%
  derive_vars_merged_dtm(
    dataset_add = ex,
    filter_add = nchar(EXSTDTC) >= 10,
    new_vars_prefix = "TRTS",
    dtc = EXSTDTC,
    order = vars(TRTSDTM, EXSEQ),
    mode = "first",
    by_vars = vars(STUDYID, USUBJID)
  ) %>%
  derive_vars_merged_dtm(
    dataset_add = ex,
    filter_add = nchar(EXENDTC) >= 10,
    new_vars_prefix = "TRTE",
    dtc = EXENDTC,
    time_imputation = "last",
    flag_imputation = "none",
    order = vars(TRTEDTM, EXSEQ),
    mode = "last",
    by_vars = vars(STUDYID, USUBJID)
  ) %>%
  derive_vars_dtm_to_dt(vars(TRTSDTM, TRTEDTM)) %>%
  derive_var_trtdurd()
```

```{r, eval=TRUE, echo=FALSE}
dataset_vignette(
  adsl,
  display_vars = vars(
    STUDYID, USUBJID, TRT01P, TRT01A,
    TRTSDTM, TRTEDTM, TRTSDT, TRTEDT,
    TRTDURD
  )
)
```

# Derivations

The most important functions in `{admiral}` are the
[derivations](../reference/index.html#section-derivations). These functions
start with `derive_`. The first parameter of these functions expects the
input dataset. This allows us to string together derivations using the `%>%`
operator.

Functions which derive a dedicated variable start with `derive_var_` followed by
the variable name, e.g., `derive_var_trtdurd()` derives the `TRTDURD` variable.

Functions which derive a dedicated parameter start with `derive_param_` followed by
the parameter name, e.g., `derive_param_bmi()` derives the `BMI` parameter.

# Input and Output

It is expected that the input dataset is not grouped. Otherwise an error is
issued.

The input dataset should not include variables starting with `temp_`. These
variable names are reserved for temporary variables used within the derivation
and are removed from the output dataset. If the input dataset contains such
variables, an error is issued.

It is expected all variable names are uppercase in the input dataset and new
variables will be returned in uppercase. It is expected all `SUPPxx.QNAM` 
values are uppercase to conform with this convention when using the 
`derive_vars_suppqual()` to adhere to the uppercase requirement.

The output dataset is ungrouped. The observations are not ordered in a dedicated
way. In particular, the order of the observations of the input dataset may not
be preserved.

# Computations

[Computations](../reference/index.html#section-computations) expect vectors as
input and return a vector. Usually these computation functions can not be used with `%>%`. These functions can be used in expressions like
`convert_dtc_to_dt()` in the derivation of `FINLABDT` in the
example below:

```{r, eval=TRUE}
# derive final lab visit date
ds_final_lab_visit <- ds %>%
  filter(DSDECOD == "FINAL LAB VISIT") %>%
  transmute(USUBJID, FINLABDT = convert_dtc_to_dt(DSSTDTC))

# derive treatment variables
adsl <- dm %>%
  # merge on final lab visit date
  derive_vars_merged(
    dataset_add = ds_final_lab_visit,
    by_vars = vars(USUBJID)
  )
```

```{r, eval=TRUE, echo=FALSE}
dataset_vignette(
  adsl,
  display_vars = vars(STUDYID, USUBJID, FINLABDT)
)
```

# Parameters

For parameters which expect variable names or expressions of variable names, 
symbols or expressions must be specified rather than strings.

- For parameters which expect a single variable name, the name can be specified
without quotes and quotation, e.g. `new_var = TEMPBL`

- For parameters which expect one or more variable names, a list of symbols is
expected, e.g. `by_vars = vars(PARAMCD, AVISIT)` 

- For parameters which expect a single expression, the expression needs to be
passed "as is", e.g. `filter = PARAMCD == "TEMP"`

- For parameters which expect one or more expressions, a list of expressions is
expected, e.g. `order = vars(AVISIT, desc(AESEV))`

# Handling of Missing Values

When using the `{haven}` package to read SAS datasets into R, SAS-style character missing values, i.e. `""`, are *not* converted into proper R `NA` values. Rather they are kept as is. This is problematic for any downstream data processing as R handles `""` just as any other string. Thus, before any data manipulation is being performed SAS blanks should be converted to R `NA`s using `{admiral}`'s `convert_blanks_to_na()` function, e.g.

```r
dm <- haven::read_sas("dm.sas7bdat") %>% 
  convert_blanks_to_na()
```

Note that any logical operator being applied to an `NA` value *always* returns `NA` rather than `TRUE` or `FALSE`.

```{r}
visits <- c("Baseline", NA, "Screening", "Week 1 Day 7")
visits != "Baseline"
```

The only exception is `is.na()` which returns `TRUE` if the input is `NA`.

```{r}
is.na(visits)
```

Thus, to filter all visits which are not `"Baseline"` the following condition would need to be used.

```{r}
visits != "Baseline" | is.na(visits)
```

Also note that most aggregation functions, like `mean()` or `max()`, also return `NA` if any element of the input vector is missing.

```{r}
mean(c(1, NA, 2))
```

To avoid this behavior one has to explicitly set `na.rm = TRUE`.

```{r}
mean(c(1, NA, 2), na.rm = TRUE)
```

This is very important to keep in mind when using `{admiral}`'s aggregation functions such as `derive_summary_records()`.

# Validation

All functions are reviewed and tested to ensure that they work as described in
the documentation. They are **not validated** yet.

Although `{admiral}` follows CDISC standards, it does not claim that the dataset
resulting from calling `{admiral}` functions is ADaM compliant. This has to be
ensured by the user.

# Starting a Script

For the ADaM data structures, an overview of the flow and example function calls
for the most common steps are provided by the following vignettes:

- [Creating ADSL](adsl.html)
- [Creating an OCCDS ADaM](occds.html)
- [Creating a BDS Finding ADaM](bds_finding.html)

`{admiral}` also provides template R scripts as a starting point. They can be
created by calling `use_ad_template()`, e.g.,

```r
use_ad_template(
  adam_name = "adsl",
  save_path = "./ad_adsl.R"
)
```

A list of all available templates can be obtained by `list_all_templates()`:

```{r}
list_all_templates()
```

# Support

Support is provided via the [admiral Slack
channel](https://app.slack.com/client/T028PB489D3/C02M8KN8269).

# See also

- [Template scripts](https://github.com/pharmaverse/admiral/tree/main/inst/templates)

- [Programming Strategy](programming_strategy.html)
