---
title: "Simple ADSL & ADVS Walkthrough"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simple ADSL & ADVS Walkthrough}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(admiraldev)
```

# Main Idea

The main idea of `{admiral}` is that an ADaM dataset is built by a sequence of
derivations. Each derivation adds one or more variables or parameters to the
processed dataset. This modular approach makes it easy to adjust code by adding,
removing, or modifying derivations. Each derivation is a function call. Consider
for example the following scripts which create a (very simple) [ADSL](#adsl_example) dataset or
[ADVS](#advs_example) dataset.

# ADSL example {#adsl_example}

## Load Packages and Example Datasets

First, we will load our packages and example datasets to help with our `ADSL` creation. The `{dplyr}` and `{lubridate}` packages are `{tidyverse}` packages and used heavily throughout this script.  The `{admiral}` package also leverages the `{pharmaversesdtm}` package for example SDTM datasets which are from the CDISC Pilot Study.

```{r, message=FALSE, warning=FALSE}
library(dplyr, warn.conflicts = FALSE)
library(lubridate)
library(admiral)
library(pharmaversesdtm)

# Read in SDTM datasets
data("dm")
data("ds")
data("ex")

dm <- convert_blanks_to_na(dm)
ds <- convert_blanks_to_na(ds)
ex <- convert_blanks_to_na(ex)
```

## Derive Treatment Variables (`TRT0xP`, `TRT0xA`) {#treatmentvargs}

The mapping of the treatment variables is left to the ADaM programmer. An example mapping may be:

```{r eval=TRUE}
adsl <- dm %>%
  mutate(TRT01P = ARM, TRT01A = ACTARM)
```

## Derive/Impute Numeric Treatment Date/Time and Flag Variables (`TRTSDTM`, `TRTEDTM`, `TRTSTMF`, `TRTETMF`) {#trtdatetimegs}

The function `derive_vars_dtm()` can be used to convert the DTC variables from
`EX` to numeric datetime variable and impute missing components. The function
call returns the original data frame with the column `EXSTDTM`, `EXENDTM` and
corresponding time imputation  flag variables `EXSTTMF` and  `EXENTMF` added to
the end of the dataframe. Exposure observations with an incomplete date are
ignored. We impute missing time to be 23:59:59 using `time_imputation = "last"`.
The required imputation flags are determined automatically by the function. Here
only time imputation flags are derived because time is imputed but date is not
imputed.

Don't be intimidated by the number of arguments! We try to make our arguments
self-explanatory, e.g. the `new_vars_prefix` places `EXST` at the start of the
`--DTM` variable and `time_imputation = "last"` appends 23:59:59.  However, this
is not always possible to make every argument self-explanatory.  If you click on
the function, `derive_vars_dtm()`, you can bring up the reference documentation
and learn more about each argument.

```{r, eval=TRUE}
# Derive treatment variables
## Impute time of exposure dates (creates numeric datetime and time imputation flag variables)
ex_ext <- ex %>%
  derive_vars_dtm(
    dtc = EXSTDTC,
    new_vars_prefix = "EXST"
  ) %>%
  derive_vars_dtm(
    dtc = EXENDTC,
    new_vars_prefix = "EXEN",
    time_imputation = "last"
  )

## Derive variables for first/last treatment date and time imputation flags
adsl <- adsl %>%
  derive_vars_merged(
    dataset_add = ex_ext,
    filter_add = !is.na(EXSTDTM),
    new_vars = exprs(TRTSDTM = EXSTDTM, TRTSTMF = EXSTTMF),
    order = exprs(EXSTDTM, EXSEQ),
    mode = "first",
    by_vars = exprs(STUDYID, USUBJID)
  ) %>%
  derive_vars_merged(
    dataset_add = ex_ext,
    filter_add = !is.na(EXENDTM),
    new_vars = exprs(TRTEDTM = EXENDTM, TRTETMF = EXENTMF),
    order = exprs(EXENDTM, EXSEQ),
    mode = "last",
    by_vars = exprs(STUDYID, USUBJID)
  )
```

## Derive Date Variables from Date/Time Variables (`TRTSDT`, `TRTEDT`) {#trtdategs}

The datetime variables returned can be converted to dates using the `derive_vars_dtm_to_dt()` 
function.

```{r eval=TRUE}
adsl <- adsl %>%
  derive_vars_dtm_to_dt(source_vars = exprs(TRTSDTM, TRTEDTM))
```

## Derive Treatment Duration (`TRTDURD`) {#trtdurgs}

Now, that `TRTSDT` and `TRTEDT` are derived, the function `derive_var_trtdurd()`
can be used to calculate the Treatment duration (`TRTDURD`).  Notice the lack of inputs.  The function defaults are set to `TRTSDT` and `TRTEDT`.  Clicking on `derive_var_trtdurd()` will bring up the reference documentation where you can see the **default** arguments.  

```{r eval=TRUE}
adsl <- adsl %>%
  derive_var_trtdurd()
```

Amazing! With one `{dplyr}` function and four `{admiral}` functions we
successfully created nine new variables for our `ADSL` dataset. Let's take a
look at all our newly derived variables.

```{r, eval=TRUE, echo=FALSE}
dataset_vignette(
  adsl,
  display_vars = exprs(
    STUDYID, USUBJID, TRT01P, TRT01A,
    TRTSDTM, TRTEDTM, TRTSDT, TRTEDT,
    TRTSTMF, TRTETMF, TRTDURD
  )
)
```

**Note:** We only display variables that were derived.  A user running this code will have additional adsl variables displayed. You can use the **Choose Columns to Display** button to add more variables into the table.  

# Derivations

The most important functions in `{admiral}` are the
[derivations](../reference/index.html#derivations-for-adding-variables).
Derivations add variables or observations to the input dataset. Existing
variables and observations of the input dataset are not changed. Derivation
functions start with `derive_`. The first argument of these functions expects
the input dataset. This allows us to string together derivations using the `%>%`
operator.

Functions which derive a dedicated variable start with `derive_var_` followed by
the variable name, e.g., `derive_var_trtdurd()` derives the `TRTDURD` variable.

Functions which can derive multiple variables start with `derive_vars_` followed
by the variable name, e.g., `derive_vars_dtm()` can derive both the `TRTSDTM`
and `TRTSTMF` variables.

Functions which derive a dedicated parameter start with `derive_param_` followed by
the parameter name, e.g., `derive_param_bmi()` derives the `BMI` parameter.

# Input and Output

It is expected that the input dataset is not grouped. Otherwise an error is
issued.

The output dataset is ungrouped. The observations are not ordered in a dedicated
way. In particular, the order of the observations of the input dataset may not
be preserved.

# Computations

[Computations](../reference/index.html#computation-functions-for-vectors) expect vectors as
input and return a vector. Usually these computation functions can not be used with `%>%`. These functions can be used in expressions like
`convert_dtc_to_dt()` in the derivation of `FINLABDT` in the
example below:

```{r, eval=TRUE}
# Add the date of the final lab visit to ADSL
adsl <- dm %>%
  derive_vars_merged(
    dataset_add = ds,
    by_vars = exprs(USUBJID),
    new_vars = exprs(FINLABDT = convert_dtc_to_dt(DSSTDTC)),
    filter_add = DSDECOD == "FINAL LAB VISIT"
  )
```

```{r, eval=TRUE, echo=FALSE}
dataset_vignette(
  adsl,
  display_vars = exprs(STUDYID, USUBJID, FINLABDT)
)
```
# ADVS example {#advs_example}

## Load Packages and Example Datasets

Similarly to ADSL's example, we will load our packages and example datasets to
help with our `ADVS` creation. The datasets required are `VS` and `ADSL`.

```{r message=FALSE}
library(admiral)
library(dplyr, warn.conflicts = FALSE)
library(pharmaversesdtm)
library(lubridate)
library(stringr)
library(tibble)

data("admiral_adsl")
data("vs")

adsl <- admiral_adsl
vs <- convert_blanks_to_na(vs)
```
```{r echo=FALSE}
vs <- filter(vs, USUBJID %in% c("01-701-1015", "01-701-1023", "01-703-1086", "01-703-1096", "01-707-1037", "01-716-1024"))
```

# Starting a Script

For the ADaM data structures, an overview of the flow and example function calls
for the most common steps are provided by the following vignettes:

- [Creating ADSL](adsl.html)
- [Creating an OCCDS ADaM](occds.html)
- [Creating a BDS Finding ADaM](bds_finding.html)

`{admiral}` also provides template R scripts as a starting point. They can be
created by calling `use_ad_template()`, e.g.,

```r
use_ad_template(
  adam_name = "adsl",
  save_path = "./ad_adsl.R"
)
```

A list of all available templates can be obtained by `list_all_templates()`:

```{r}
library(admiral)
list_all_templates()
```

# Support

Support is provided via the [admiral Slack
channel](https://app.slack.com/client/T028PB489D3/C02M8KN8269).

# See also

- [Template scripts](https://github.com/pharmaverse/admiral/tree/main/inst/templates)

- [Programming Strategy](https://pharmaverse.github.io/admiraldev/articles/programming_strategy.html)
