---
title: "Get Started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(admiral)
```

# Main Idea

The main idea of admiral is that an ADaM dataset is build by a sequence of
derivations. Each derivation adds one or more variables or parameters to the
processed dataset. This modular approach makes it easy to adjust code by adding,
removing, or modifying derivations. Each derivation is a function call. Consider
for example the following script which creates a (very simple) ADSL dataset.

```r
library(dplyr)
library(lubridate)
library(admiral)

# read in SDTM datasets
data("dm")
data("ds")
data("ex")

# derive treatment variables
adsl <- dm %>%
  mutate(TRT01P = ARMCD, TRT01A = ACTARMCD) %>%
  derive_var_trtsdtm(dataset_ex = ex) %>%
  mutate(TRTSDT = date(TRTSDTM)) %>%
  derive_var_trtedtm(dataset_ex = ex) %>%
  mutate(TRTEDT = date(TRTEDTM)) %>%
  derive_var_trtdurd()
```

# Derivations

The most important functions in admiral are the
[derivations](../reference/index.html#section-derivations). These functions
start with `derive_`. The first parameter of these functions expects the
input dataset. This allows to string together derivations using the `%>%`
operator.

Functions which derive a dedicated variable start with `derive_var_` followed by
the variable name, e.g., `derive_var_trtdurd()` derives the `TRTDURD` variable.

Functions which derive a dedicated parameter start with `derive_param_` followed by
the parameter name, e.g., `derive_param_os()` derives the `OS` parameter.

# Input and Output

It is expected that the input dataset is not grouped. Otherwise an error is
issued.

The input dataset should not include variables starting with `temp_`. These
variable names are reserved for temporary variable used within the derivation
and are removed from the output dataset. If the input dataset contains such
variables, an error is issued.

It is expected all variable names are uppercase in the input dataset and new
variables will be returned in uppercase. It is expected all `SUPPxx.QNAM` 
values are uppercase to conform with this convention when using the 
`derive_vars_suppqual()` to adhere to the uppercase requirement.

The output dataset is ungrouped. The observations are not ordered in a dedicated
way. In particular, the order of the observations of the input dataset may not
be preserved.

# Handling of Missing Values

When using the {haven} package to read SAS datasets into R, SAS-style character missing values, i.e. `""`, are *not* converted into proper R `NA` values. Rather they are kept as is. This is problematic for any downstream data processing as R handles `""` just as any other string. Thus, before any data manipulation is being performed SAS blanks should be converted to R `NA`s using {admiral}'s `convert_blanks_to_na()` function, e.g.

```r
dm <- haven::read_sas("dm.sas7bdat") %>% 
  convert_blanks_to_na()
```

Note that any logical operator being applied to an `NA` value *always* returns `NA` rather than `TRUE` or `FALSE`.

```{r}
visits <- c("Baseline", NA, "Screening", "Week 1 Day 7")
visits != "Baseline"
```

The only exception is `is.na()` which returns `TRUE` if the input is `NA`.

```{r}
is.na(visits)
```

Thus, to filter all visits which are not `"Baseline"` the following condition would need to be used.

```{r}
visits != "Baseline" | is.na(visits)
```

Also note that most aggregation functions, like `mean()` or `max()`, also return `NA` if any element of the input vector is missing.

```{r}
mean(c(1, NA, 2))
```

To avoid this behavior one has to explicitly set `na.rm = TRUE`.

```{r}
mean(c(1, NA, 2), na.rm = TRUE)
```

This is very important to keep in mind when using {admiral}'s aggregation functions such as `derive_summary_records()`.

# Computations

[Computations](../reference/index.html#section-computations) expect vectors as
input and return a vector. These function can be used in expressions like
`convert_dtc_to_dt()` and `impute_dtc()` in the derivation of `EOSDT` in the
example above.

# Parameters

For parameters which expect variable names or expressions of variable names
symbols or expressions must be specified rather than strings.

- For parameters which expects a single variable name the name can be specified
without quotes and quotation, e.g. `new_var = TEMPBL`

- For parameters which expects one or more variable names a list of symbols is
expected, e.g. `by_vars = vars(PARAMCD, AVISIT)` 

- For parameters which expects a single expression the expression needs to be
passed "as is", e.g. `filter = PARAMCD == "TEMP"`

- For parameters which expects one or more expressions a list of expressions is
expected, e.g. `order = desc(AVISIT, desc(AESEV))`

# Validation

All functions are reviewed and tested to ensure that they work as described in
the documentation. They are **not validated** yet.

Although admiral follows CDISC standards it does not claim that the dataset
resulting from calling admiral functions is ADaM compliant. This has to be
ensured by the user.

# Starting a Script

For the ADaM data structures an overview of the flow and example function calls
for the most common steps are provided by the following vignettes:

- [Creating ADSL](adsl.html)
- [Creating an OCCDS ADaM](occds.html)
- [Creating a BDS Finding ADaM](bds_finding.html)

admiral also provides template R scripts as a starting point. They can be
created by calling `use_ad_template()`, e.g.,

```r
use_ad_template(
  adam_name = "adsl",
  save_path = "./ad_adsl.R"
)
```

A list of all available templates can be obtained by `list_all_templates()`:
```{r}
list_all_templates()
```

# Support

# See also

- [Template scripts](https://github.com/Roche-GSK/admiral/tree/master/inst/templates)

- [Contributing to admiral](contributing.html)
