---
title: "Creating ADSL"
output: 
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating ADSL}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(admiraldev)
```

# Introduction

This article describes creating an `ADSL` ADaM. Examples are currently 
presented and tested using `DM`, `EX` , `AE`, `LB` and `DS` SDTM domains. However, other domains 
could be used.

**Note:** *All examples assume CDISC SDTM and/or ADaM format as input unless 
otherwise specified.*

# Programming Flow

* [Read in Data](#readdata)
* [Derive Period, Subperiod, and Phase Variables (e.g. `APxxSDT`, `APxxEDT`, ...)](#periodvars)
* [Derive Treatment Variables (`TRT0xP`, `TRT0xA`)](#treatmentvar)
* [Derive/Impute Numeric Treatment Date/Time and Duration (`TRTSDT`, `TRTEDT`, `TRTDURD`)](#trtdatetime)
* [Derive Disposition Variables](#disposition)
  * [Disposition Dates (e.g. `EOSDT`)](#disposition_date)
  * [Disposition Status (e.g. `EOSTT`)](#disposition_status)
  * [Disposition Reason(s) (e.g. `DCSREAS`, `DCSREASP`)](#disposition_reason)
  * [Randomization Date (`RANDDT`)](#randomization_date)
* [Derive Death Variables](#death)
  * [Death Date (`DTHDT`)](#death_date)
  * [Cause of Death (`DTHCAUS`)](#death_cause)
  * [Duration Relative to Death](#death_other)
* [Derive Last Known Date Alive (`LSTALVDT`)](#lstalvdt)
* [Derive Groupings and Populations](#groupings)
  * [Grouping (e.g. `AGEGR1` or `REGION1`)](#groupings_ex)
  * [Population Flags (e.g. `SAFFL`)](#popflag)
* [Derive Other Variables](#other)
* [Add Labels and Attributes](#attributes)


## Read in Data {#readdata}

To start, all data frames needed for the creation of `ADSL` should be read into
the environment. This will be a company specific process. Some of the 
data frames needed may be `DM`, `EX`, `DS`, `AE`, and `LB`.

For example purpose, the CDISC Pilot SDTM datasets---which are included in 
`{pharmaversesdtm}`---are used.

```{r, message=FALSE, warning=FALSE}
library(admiral)
library(dplyr, warn.conflicts = FALSE)
library(pharmaversesdtm)
library(lubridate)
library(stringr)

data("dm")
data("ds")
data("ex")
data("ae")
data("lb")

dm <- convert_blanks_to_na(dm)
ds <- convert_blanks_to_na(ds)
ex <- convert_blanks_to_na(ex)
ae <- convert_blanks_to_na(ae)
lb <- convert_blanks_to_na(lb)
```

The `DM` domain is used as the basis for `ADSL`:

```{r eval=TRUE}
adsl <- dm %>%
  select(-DOMAIN)
```

```{r, eval=TRUE, echo=FALSE}
dataset_vignette(
  adsl,
  display_vars = exprs(USUBJID, RFSTDTC, COUNTRY, AGE, SEX, RACE, ETHNIC, ARM, ACTARM)
)
```

## Derive Period, Subperiod, and Phase Variables (e.g. `APxxSDT`, `APxxEDT`, ...) {#periodvars}

See the ["Visit and Period Variables"
vignette](visits_periods.html#periods_adsl) for more information.

If the variables are not derived based on a period reference dataset, they may
be derived at a later point of the flow. For example, phases like "Treatment
Phase" and "Follow up" could be derived based on treatment start and end date.

## Derive Treatment Variables (`TRT0xP`, `TRT0xA`) {#treatmentvar}

The mapping of the treatment variables is left to the ADaM programmer. An
example mapping for a study without periods may be:

```{r eval=TRUE}
adsl <- dm %>%
  mutate(TRT01P = ARM, TRT01A = ACTARM)
```

For studies with periods see the ["Visit and Period Variables"
vignette](visits_periods.html#treatment_adsl).

## Derive/Impute Numeric Treatment Date/Time and Duration (`TRTSDTM`, `TRTEDTM`, `TRTDURD`) {#trtdatetime}

The function `derive_vars_merged()` can be used to derive the treatment start
and end date/times using the `ex` domain. A pre-processing step for `ex` is
required to convert the variable `EXSTDTC` and `EXSTDTC` to datetime variables
and impute missing date or time components. Conversion and imputation is done by
`derive_vars_dtm()`.

Example calls:

```{r eval=TRUE}
# impute start and end time of exposure to first and last respectively,
# do not impute date
ex_ext <- ex %>%
  derive_vars_dtm(
    dtc = EXSTDTC,
    new_vars_prefix = "EXST"
  ) %>%
  derive_vars_dtm(
    dtc = EXENDTC,
    new_vars_prefix = "EXEN",
    time_imputation = "last"
  )

adsl <- adsl %>%
  derive_vars_merged(
    dataset_add = ex_ext,
    filter_add = (EXDOSE > 0 |
      (EXDOSE == 0 &
        str_detect(EXTRT, "PLACEBO"))) & !is.na(EXSTDTM),
    new_vars = exprs(TRTSDTM = EXSTDTM, TRTSTMF = EXSTTMF),
    order = exprs(EXSTDTM, EXSEQ),
    mode = "first",
    by_vars = exprs(STUDYID, USUBJID)
  ) %>%
  derive_vars_merged(
    dataset_add = ex_ext,
    filter_add = (EXDOSE > 0 |
      (EXDOSE == 0 &
        str_detect(EXTRT, "PLACEBO"))) & !is.na(EXENDTM),
    new_vars = exprs(TRTEDTM = EXENDTM, TRTETMF = EXENTMF),
    order = exprs(EXENDTM, EXSEQ),
    mode = "last",
    by_vars = exprs(STUDYID, USUBJID)
  )
```

This call returns the original data frame with the column `TRTSDTM`, `TRTSTMF`,
`TRTEDTM`, and `TRTETMF` added. Exposure observations with incomplete date and
zero doses of non placebo treatments are ignored. Missing time parts are imputed
as first or last for start and end date respectively.

The datetime variables returned can be converted to dates using the `derive_vars_dtm_to_dt()` 
function.

```{r eval=TRUE}
adsl <- adsl %>%
  derive_vars_dtm_to_dt(source_vars = exprs(TRTSDTM, TRTEDTM))
```


Now, that `TRTSDT` and `TRTEDT` are derived, the function `derive_var_trtdurd()`
can be used to calculate the Treatment duration (`TRTDURD`).

```{r eval=TRUE}
adsl <- adsl %>%
  derive_var_trtdurd()
```

```{r, eval=TRUE, echo=FALSE}
dataset_vignette(
  adsl,
  display_vars = exprs(USUBJID, RFSTDTC, TRTSDTM, TRTSDT, TRTEDTM, TRTEDT, TRTDURD)
)
```


## Derive Disposition Variables {#disposition}

### Disposition Dates (e.g. `EOSDT`) {#disposition_date}

The functions `derive_vars_dt()` and `derive_vars_merged()` can be used to
derive a disposition date. First the character disposition date (`DS.DSSTDTC`)
is converted to a numeric date (`DSSTDT`) calling `derive_vars_dt()`. The `DS`
dataset is extended by the `DSSTDT` variable because the date is required by
other derivations, e.g., `RANDDT` as well. Then the relevant disposition date is
selected by adjusting the `filter_add` argument.

To add the End of Study date (`EOSDT`) to the input dataset, a call could be:

```{r eval=TRUE}
# convert character date to numeric date without imputation
ds_ext <- derive_vars_dt(
  ds,
  dtc = DSSTDTC,
  new_vars_prefix = "DSST"
)

adsl <- adsl %>%
  derive_vars_merged(
    dataset_add = ds_ext,
    by_vars = exprs(STUDYID, USUBJID),
    new_vars = exprs(EOSDT = DSSTDT),
    filter_add = DSCAT == "DISPOSITION EVENT" & DSDECOD != "SCREEN FAILURE"
  )
```

The `ds_ext` dataset:
```{r, eval=TRUE, echo=FALSE}
dataset_vignette(
  ds_ext,
  display_vars = exprs(USUBJID, DSCAT, DSDECOD, DSTERM, DSSTDT, DSSTDTC),
  filter = DSDECOD != "SCREEN FAILURE"
)
```

The `adsl` dataset:
```{r, eval=TRUE, echo=FALSE}
dataset_vignette(adsl, display_vars = exprs(USUBJID, EOSDT))
```

The `derive_vars_dt()` function allows to impute partial dates as well. If
imputation is needed and missing days are to be imputed to the first of the
month and missing months to the first month of the year, set `highest_imputation
= "M"`.

### Disposition Status (e.g. `EOSSTT`) {#disposition_status}

The function `derive_vars_merged()` can be used to derive the End of Study
status (`EOSSTT`) based on `DSCAT` and `DSDECOD` from `DS`. The relevant
observations are selected by adjusting the `filter_add` argument. A function
mapping `DSDECOD` values to `EOSSTT` values can be defined and used in the
`new_vars` argument. The mapping for the call below is

- `"COMPLETED"` if `DSDECOD == "COMPLETED"`
- `NA_character_` if `DSDECOD` is `"SCREEN FAILURE"`
- `"DISCONTINUED"` otherwise

Example function `format_eosstt()`:

```{r eval=TRUE}
format_eosstt <- function(x) {
  case_when(
    x %in% c("COMPLETED") ~ "COMPLETED",
    x %in% c("SCREEN FAILURE") ~ NA_character_,
    TRUE ~ "DISCONTINUED"
  )
}
```

The customized mapping function `format_eosstt()` can now be passed to the main
function. For subjects without a disposition event the end of study status is
set to `"ONGOING"` by specifying the `missing_values` argument.

```{r eval=TRUE}
adsl <- adsl %>%
  derive_vars_merged(
    dataset_add = ds,
    by_vars = exprs(STUDYID, USUBJID),
    filter_add = DSCAT == "DISPOSITION EVENT",
    new_vars = exprs(EOSSTT = format_eosstt(DSDECOD)),
    missing_values = exprs(EOSSTT = "ONGOING")
  )
```

```{r, eval=TRUE, echo=FALSE}
dataset_vignette(adsl, display_vars = exprs(USUBJID, EOSDT, EOSSTT))
```

This call would return the input dataset with the column `EOSSTT` added. 

If the derivation must be changed, the user can create his/her own function to
map `DSDECOD` to a suitable `EOSSTT` value.

### Disposition Reason(s) (e.g. `DCSREAS`, `DCSREASP`) {#disposition_reason}

The main reason for discontinuation is usually stored in `DSDECOD` while `DSTERM` provides 
additional details regarding subject’s discontinuation (e.g., description of `"OTHER"`).

The function `derive_vars_merged()` can be used to derive a disposition reason
(along with the details, if required) at a specific timepoint. The relevant
observations are selected by adjusting the `filter_add` argument.

To derive the End of Study reason(s) (`DCSREAS` and `DCSREASP`), the function
will map `DCSREAS` as `DSDECOD`, and `DCSREASP` as `DSTERM` if `DSDECOD` is not
`"COMPLETED"`, `"SCREEN FAILURE"`, or `NA`, `NA` otherwise.

```{r}
adsl <- adsl %>%
  derive_vars_merged(
    dataset_add = ds,
    by_vars = exprs(USUBJID),
    new_vars = exprs(DCSREAS = DSDECOD, DCSREASP = DSTERM),
    filter_add = DSCAT == "DISPOSITION EVENT" &
      !(DSDECOD %in% c("SCREEN FAILURE", "COMPLETED", NA))
  )
```

```{r, eval=TRUE, echo=FALSE}
dataset_vignette(adsl, display_vars = exprs(USUBJID, EOSDT, EOSSTT, DCSREAS, DCSREASP))
```
This call would return the input dataset with the column `DCSREAS` and `DCSREASP` added. 

If the derivation must be changed, the user can define that derivation in the `filter_add` argument of the function to map `DSDECOD` and  `DSTERM` to a suitable `DCSREAS`/`DCSREASP` value.

The call below maps `DCSREAS` and `DCREASP` as follows:

- `DCSREAS` as `DSDECOD` if `DSDECOD` is not `"COMPLETED"` or `NA`, `NA` otherwise
- `DCSREASP` as `DSTERM` if `DSDECOD` is equal to `OTHER`, `NA` otherwise

```{r, eval=TRUE, echo=FALSE}
adsl <- adsl %>%
  select(-DCSREAS, -DCSREASP)
```

```{r}
adsl <- adsl %>%
  derive_vars_merged(
    dataset_add = ds,
    by_vars = exprs(USUBJID),
    new_vars = exprs(DCSREAS = DSDECOD),
    filter_add = DSCAT == "DISPOSITION EVENT" &
      DSDECOD %notin% c("SCREEN FAILURE", "COMPLETED", NA)
  ) %>%
  derive_vars_merged(
    dataset_add = ds,
    by_vars = exprs(USUBJID),
    new_vars = exprs(DCSREASP = DSTERM),
    filter_add = DSCAT == "DISPOSITION EVENT" & DSDECOD %in% "OTHER"
  )
```

```{r, eval=TRUE, echo=FALSE}
dataset_vignette(adsl, display_vars = exprs(USUBJID, EOSDT, EOSSTT, DCSREAS, DCSREASP))
```

### Randomization Date (`RANDDT`) {#randomization_date}

The function `derive_vars_merged()` can be used to derive randomization date 
variable. To map Randomization Date (`RANDDT`), the call would be:

```{r eval=TRUE}
adsl <- adsl %>%
  derive_vars_merged(
    dataset_add = ds_ext,
    filter_add = DSDECOD == "RANDOMIZED",
    by_vars = exprs(STUDYID, USUBJID),
    new_vars = exprs(RANDDT = DSSTDT)
  )
```


This call would return the input dataset with the column `RANDDT` is added.

```{r, eval=TRUE, echo=FALSE}
dataset_vignette(adsl, display_vars = exprs(USUBJID, RANDDT))
```

## Derive Death Variables {#death}

### Death Date (`DTHDT`) {#death_date}

The function `derive_vars_dt()` can be used to derive `DTHDT`. This function allows 
the user to impute the date as well.

Example calls:

```{r eval=TRUE}
adsl <- adsl %>%
  derive_vars_dt(
    new_vars_prefix = "DTH",
    dtc = DTHDTC
  )
```

```{r, eval=TRUE, echo=FALSE}
dataset_vignette(adsl %>% filter(!is.na(DTHDT) | row_number() %% 50 == 0), display_vars = exprs(USUBJID, TRTEDT, DTHDTC, DTHDT, DTHFL))
```

This call would return the input dataset with the columns `DTHDT` added and, by
default, the associated date imputation flag (`DTHDTF`) populated with the
controlled terminology outlined in the ADaM IG for date imputations. If the
imputation flag is not required, the user must set the argument
`flag_imputation` to `"none"`.

If imputation is needed and the date is to be imputed to the first day of the month/year 
the call would be:

```{r eval=FALSE}
adsl <- adsl %>%
  derive_vars_dt(
    new_vars_prefix = "DTH",
    dtc = DTHDTC,
    date_imputation = "first"
  )
```

See also [Date and Time Imputation](imputation.html).

### Cause of Death  (`DTHCAUS`) {#death_cause}

The cause of death `DTHCAUS` can be derived using the function `derive_vars_extreme_event()`.

Since the cause of death could be collected/mapped in different domains (e.g. `DS`, `AE`, `DD`), it 
is important the user specifies the right source(s) to derive the cause of death from.

For example, if the date of death is collected in the AE form when the AE is Fatal, the cause of 
death would be set to the preferred term (`AEDECOD`) of that Fatal AE, while if the date of death is 
collected in  the `DS` form, the cause of death would be set to the disposition term (`DSTERM`).
To achieve this, the `event()` objects within `derive_vars_extreme_event()` must be specified and defined such that they fit the 
study requirement. 

An example call to `derive_vars_extreme_event()` would be: 

```{r eval=TRUE}
adsl <- adsl %>%
  derive_vars_extreme_event(
    by_vars = exprs(STUDYID, USUBJID),
    events = list(
      event(
        dataset_name = "ae",
        condition = AEOUT == "FATAL",
        set_values_to = exprs(DTHCAUS = AEDECOD),
      ),
      event(
        dataset_name = "ds",
        condition = DSDECOD == "DEATH" & grepl("DEATH DUE TO", DSTERM),
        set_values_to = exprs(DTHCAUS = DSTERM),
      )
    ),
    source_datasets = list(ae = ae, ds = ds),
    tmp_event_nr_var = event_nr,
    order = exprs(event_nr),
    mode = "first",
    new_vars = exprs(DTHCAUS)
  )
```

```{r, eval=TRUE, echo=FALSE}
dataset_vignette(
  adsl,
  display_vars = exprs(USUBJID, EOSDT, DTHDTC, DTHDT, DTHCAUS),
  filter = DTHFL == "Y"
)
```

The function also offers the option to add some traceability variables (e.g. `DTHDOM` would 
store the domain where the date of death is collected, and `DTHSEQ` would store the `xxSEQ` value of 
that domain). The traceability variables should be added to the `event()` calls and included in the `new_vars` parameter of `derive_vars_extreme_event()`.

```{r eval=TRUE}
adsl <- adsl %>%
  select(-DTHCAUS) %>% # remove it before deriving it again
  derive_vars_extreme_event(
    by_vars = exprs(STUDYID, USUBJID),
    events = list(
      event(
        dataset_name = "ae",
        condition = AEOUT == "FATAL",
        set_values_to = exprs(DTHCAUS = AEDECOD, DTHDOM = "AE", DTHSEQ = AESEQ),
      ),
      event(
        dataset_name = "ds",
        condition = DSDECOD == "DEATH" & grepl("DEATH DUE TO", DSTERM),
        set_values_to = exprs(DTHCAUS = DSTERM, DTHDOM = "DS", DTHSEQ = DSSEQ),
      )
    ),
    source_datasets = list(ae = ae, ds = ds),
    tmp_event_nr_var = event_nr,
    order = exprs(event_nr),
    mode = "first",
    new_vars = exprs(DTHCAUS, DTHDOM, DTHSEQ)
  )
```

```{r, eval=TRUE, echo=FALSE}
dataset_vignette(
  adsl,
  display_vars = exprs(USUBJID, TRTEDT, DTHDTC, DTHDT, DTHCAUS, DTHDOM, DTHSEQ),
  filter = DTHFL == "Y"
)
```

Following the derivation of `DTHCAUS` and related traceability variables, it is then possible to derive grouping variables such as death categories (`DTHCGRx`) using standard tidyverse code.

```{r eval=TRUE}
adsl <- adsl %>%
  mutate(DTHCGR1 = case_when(
    is.na(DTHDOM) ~ NA_character_,
    DTHDOM == "AE" ~ "ADVERSE EVENT",
    str_detect(DTHCAUS, "(PROGRESSIVE DISEASE|DISEASE RELAPSE)") ~ "PROGRESSIVE DISEASE",
    TRUE ~ "OTHER"
  ))
```

### Duration Relative to Death {#death_other}

The function `derive_vars_duration()` can be used to derive duration relative to death like the Relative 
Day of Death (`DTHADY`) or the numbers of days from last dose to death (`LDDTHELD`).

Example calls: 

- Relative Day of Death
```{r eval=TRUE}
adsl <- adsl %>%
  derive_vars_duration(
    new_var = DTHADY,
    start_date = TRTSDT,
    end_date = DTHDT
  )
```

- Elapsed Days from Last Dose to Death
```{r eval=TRUE}
adsl <- adsl %>%
  derive_vars_duration(
    new_var = LDDTHELD,
    start_date = TRTEDT,
    end_date = DTHDT,
    add_one = FALSE
  )
```

```{r, eval=TRUE, echo=FALSE}
dataset_vignette(
  adsl,
  display_vars = exprs(USUBJID, TRTEDT, DTHDTC, DTHDT, DTHCAUS, DTHADY, LDDTHELD),
  filter = DTHFL == "Y"
)
```

## Derive the Last Date Known Alive (`LSTALVDT`) {#lstalvdt}

Similarly as for the cause of death (`DTHCAUS`), the last known alive date
(`LSTALVDT`) can be derived from multiples sources using `derive_vars_extreme_event()`.

An example could be (DTC dates are converted to numeric dates imputing missing
day and month to the first):

```{r eval=TRUE}
adsl <- adsl %>%
  derive_vars_extreme_event(
    by_vars = exprs(STUDYID, USUBJID),
    events = list(
      event(
        dataset_name = "ae",
        order = exprs(AESTDTC, AESEQ),
        condition = !is.na(AESTDTC),
        set_values_to = exprs(
          LSTALVDT = convert_dtc_to_dt(AESTDTC, highest_imputation = "M"),
          seq = AESEQ
        ),
      ),
      event(
        dataset_name = "ae",
        order = exprs(AEENDTC, AESEQ),
        condition = !is.na(AEENDTC),
        set_values_to = exprs(
          LSTALVDT = convert_dtc_to_dt(AEENDTC, highest_imputation = "M"),
          seq = AESEQ
        ),
      ),
      event(
        dataset_name = "lb",
        order = exprs(LBDTC, LBSEQ),
        condition = !is.na(LBDTC),
        set_values_to = exprs(
          LSTALVDT = convert_dtc_to_dt(LBDTC, highest_imputation = "M"),
          seq = LBSEQ
        ),
      ),
      event(
        dataset_name = "adsl",
        condition = !is.na(TRTEDT),
        set_values_to = exprs(LSTALVDT = TRTEDT, seq = 0),
      )
    ),
    source_datasets = list(ae = ae, lb = lb, adsl = adsl),
    tmp_event_nr_var = event_nr,
    order = exprs(LSTALVDT, seq, event_nr),
    mode = "last",
    new_vars = exprs(LSTALVDT)
  )
```

```{r, eval=TRUE, echo=FALSE}
dataset_vignette(
  adsl,
  display_vars = exprs(USUBJID, TRTEDT, DTHDTC, LSTALVDT),
  filter = !is.na(TRTSDT)
)
```

Traceability variables can be added by specifying the variables in the `set_values_to` parameter of the `event()` function.

```{r eval=TRUE}
adsl <- adsl %>%
  select(-LSTALVDT) %>% # created in the previous call
  derive_vars_extreme_event(
    by_vars = exprs(STUDYID, USUBJID),
    events = list(
      event(
        dataset_name = "ae",
        order = exprs(AESTDTC, AESEQ),
        condition = !is.na(AESTDTC),
        set_values_to = exprs(
          LSTALVDT = convert_dtc_to_dt(AESTDTC, highest_imputation = "M"),
          LALVSEQ = AESEQ,
          LALVDOM = "AE",
          LALVVAR = "AESTDTC"
        ),
      ),
      event(
        dataset_name = "ae",
        order = exprs(AEENDTC, AESEQ),
        condition = !is.na(AEENDTC),
        set_values_to = exprs(
          LSTALVDT = convert_dtc_to_dt(AEENDTC, highest_imputation = "M"),
          LALVSEQ = AESEQ,
          LALVDOM = "AE",
          LALVVAR = "AEENDTC"
        ),
      ),
      event(
        dataset_name = "lb",
        order = exprs(LBDTC, LBSEQ),
        condition = !is.na(LBDTC),
        set_values_to = exprs(
          LSTALVDT = convert_dtc_to_dt(LBDTC, highest_imputation = "M"),
          LALVSEQ = LBSEQ,
          LALVDOM = "LB",
          LALVVAR = "LBDTC"
        ),
      ),
      event(
        dataset_name = "adsl",
        condition = !is.na(TRTEDT),
        set_values_to = exprs(LSTALVDT = TRTEDT, LALVSEQ = NA_integer_, LALVDOM = "ADSL", LALVVAR = "TRTEDTM"),
      )
    ),
    source_datasets = list(ae = ae, lb = lb, adsl = adsl),
    tmp_event_nr_var = event_nr,
    order = exprs(LSTALVDT, LALVSEQ, event_nr),
    mode = "last",
    new_vars = exprs(LSTALVDT, LALVSEQ, LALVDOM, LALVVAR)
  )
```

```{r, eval=TRUE, echo=FALSE}
dataset_vignette(
  adsl,
  display_vars = exprs(USUBJID, TRTEDT, DTHDTC, LSTALVDT, LALVDOM, LALVSEQ, LALVVAR),
  filter = !is.na(TRTSDT)
)
```
 

## Derive Groupings and Populations {#groupings}

### Grouping (e.g. `AGEGR1` or `REGION1`) {#groupings_ex}

Numeric and categorical variables (`AGE`, `RACE`, `COUNTRY`, etc.) may need to be grouped to perform
the required analysis.
`{admiral}` does not **currently** have functionality to assist with all required groupings. So, the user will often need to create his/her own function to meet his/her study requirement. 

For example, if 

- `AGEGR1` is required to categorize `AGE` into `<18`, `18-64` and `>64`, or
- `REGION1` is required to categorize `COUNTRY` in `North America`, `Rest of the World`,

the user defined functions would look like the following:

```{r eval=TRUE}
format_agegr1 <- function(var_input) {
  case_when(
    var_input < 18 ~ "<18",
    between(var_input, 18, 64) ~ "18-64",
    var_input > 64 ~ ">64",
    TRUE ~ "Missing"
  )
}

format_region1 <- function(var_input) {
  case_when(
    var_input %in% c("CAN", "USA") ~ "North America",
    !is.na(var_input) ~ "Rest of the World",
    TRUE ~ "Missing"
  )
}
```

These functions are then used in a `mutate()` statement to derive the required grouping variables:

```{r eval=TRUE}
adsl <- adsl %>%
  mutate(
    AGEGR1 = format_agegr1(AGE),
    REGION1 = format_region1(COUNTRY)
  )
```

```{r, eval=TRUE, echo=FALSE}
dataset_vignette(
  adsl,
  display_vars = exprs(USUBJID, AGE, SEX, COUNTRY, AGEGR1, REGION1)
)
```

### Population Flags (e.g. `SAFFL`) {#popflag}

Since the populations flags are mainly company/study specific no dedicated functions are
provided, but in most cases they can easily be derived using `derive_var_merged_exist_flag`.

An example of an implementation could be:

```{r eval=TRUE}
adsl <- adsl %>%
  derive_var_merged_exist_flag(
    dataset_add = ex,
    by_vars = exprs(STUDYID, USUBJID),
    new_var = SAFFL,
    condition = (EXDOSE > 0 | (EXDOSE == 0 & str_detect(EXTRT, "PLACEBO")))
  )
```

```{r, eval=TRUE, echo=FALSE}
dataset_vignette(
  adsl,
  display_vars = exprs(USUBJID, TRTSDT, ARM, ACTARM, SAFFL)
)
```


## Derive Other Variables  {#other}
The users can add specific code to cover their need for the analysis.

The following functions are helpful for many ADSL derivations:

  - `derive_vars_merged()` - Merge Variables from a Dataset to the Input Dataset
  - `derive_var_merged_exist_flag()` - Merge an Existence Flag
  - `derive_var_merged_summary()` - Merge Summary Variables

See also [Generic Functions](generic.html).

## Add Labels and Attributes {#attributes}

Adding labels and attributes for SAS transport files is supported by the
following packages:

- [metacore](https://atorus-research.github.io/metacore/): establish a common
foundation for the use of metadata within an R session.

- [metatools](https://pharmaverse.github.io/metatools/): enable the use of
metacore objects. Metatools can be used to build datasets or enhance columns in
existing datasets as well as checking datasets against the metadata.

- [xportr](https://atorus-research.github.io/xportr/): functionality to
associate all metadata information to a local R data frame, perform data set
level validation checks and convert into a [transport v5
file(xpt)](https://documentation.sas.com/doc/en/pgmsascdc/9.4_3.5/movefile/n1xbwdre0giahfn11c99yjkpi2yb.htm).

NOTE: All these packages are in the experimental phase, but the vision is to
have them associated with an End to End pipeline under the umbrella of the
[pharmaverse](https://github.com/pharmaverse). An example of applying metadata 
and perform associated checks can be found at the [pharmaverse E2E example](https://examples.pharmaverse.org/data/adsl/).

# Example Script 

ADaM | Sourcing Command
---- | --------------
ADSL | `use_ad_template("ADSL")`
