---
title: "Creating ADSL"
output: 
  rmarkdown::html_vignette:
    toc: false
vignette: >
  %\VignetteIndexEntry{Creating ADSL}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

This article describes creating an ADSL ADaM. Examples are currently 
presented and tested using `dm`, `ex` , `ae`, `lb` and `ds` SDTM domains. However, other domains 
could be used.

**Note:** *All examples assume CDISC SDTM and/or ADaM format as input unless 
otherwise specified.*

# Programming Flow

* [Read in Data](#readdata)
* [Derive Treatment Variables (`TRT0xP`, `TRT0xA`)](#treatmentvar)
* [Derive/Impute Numeric Treatment Date/Time and Duration (`TRTSDT`, `TRTEDT`, `TRTDURD`)](#trtdatetime)
* [Derive Disposition Variables](#disposition)
  * [Disposition Dates (e.g. `EOSDT`)](#disposition_date)
  * [Disposition Status (e.g. `EOSTT`)](#disposition_status)
  * [Disposition Reason(s) (e.g. `DCSREAS`, `DCSREASP`)](#disposition_reason)
* [Derive Death Variables](#death)
  * [Death Date (`DTHDT`)](#death_date)
  * [Cause of Death (`DTHCAUS`)](#death_cause)
  * [Duration Relative to Death](#death_other)
* [Derive Last Known Date Alive (`LSTALVDT`)](#lstalvdt)
* [Derive Groupings and Populations](#groupings)
  * [Grouping (e.g. `AGEGR1`)](#groupings_ex)
  * [Population Flags (e.g. `SAFFL`)](#popflag)
  * [Other Variables Required for Analysis](#other)
* [Example Script](#example)

## Read in Data {#readdata}

To start, all data frames needed for the creation of ADSL should be read into
the environment. This will be a company specific process. Some of the 
data frames needed may be `dm`, `ex`, `ds`, `ae`, and `lb`.

For example purpose, the CDISC Pilot SDTM datasets---which are included in `{admiral}`---are used.

```{r}
library(dplyr)
library(admiral)
library(lubridate)
library(stringr)

data("dm")
data("ds")
data("ex")
data("ae")
data("lb")
```

The `dm` domain is used as the basis for `ADSL`:

```{r eval=TRUE}
adsl <- dm %>% 
  mutate(DOMAIN = NULL)
head(adsl)
```

## Derive Treatment Variables (`TRT0xP`, `TRT0xA`) {#treatmentvar}

The mapping of the treatment variables is left to the ADaM programmer. An example mapping may be:

```{r eval=TRUE}
adsl <- dm %>%
  mutate(TRT01P = ARM, TRT01A = ACTARM)
```

## Derive/Impute Numeric Treatment Date/Time and Duration (`TRTSDTM`, `TRTEDTM`, `TRTDURD`) {#trtdatetime}

The functions `derive_var_trtsdtm()`, `derive_var_trtedtm()` can be used to derive the treatment 
start and end date/times using the `ex` domain.

Example calls:

```{r eval=TRUE}
adsl <- adsl %>%
  derive_var_trtsdtm(dataset_ex = ex) %>%
  derive_var_trtedtm(dataset_ex = ex)
```

This call returns the original data frame with the column `TRTSDTM` and `TRTEDTM` added.
The datetime variables returned can be converted to dates using the `date()` function from the `{lubridate}` package.

```{r eval=TRUE}
adsl <- adsl %>%
  mutate(
    TRTSDT = date(TRTSDTM),
    TRTEDT = date(TRTEDTM)
  )
```


Now, that `TRTSDT` and `TRTEDT` are derived, the function `derive_var_trtdurd()`
can be used to calculate the Treatment duration (`TRTDURD`).

```{r eval=TRUE}
adsl <- adsl %>%
  derive_var_trtdurd()

head(adsl %>% select(USUBJID, starts_with("TRT"), -starts_with("TRT0")))
```


## Derive Disposition Variables {#disposition}

### Disposition Dates (e.g. `EOSDT`) {#disposition_date}

The function `derive_disposition_dt()` can be used to derive a disposition date. 
The relevant disposition date (`ds.DSSTDTC`) is selected by adjusting the filter parameter. 

To derive the End of Study date (`EOSDT`), a call could be:

```{r eval=TRUE}
adsl <- adsl %>%
  derive_disposition_dt(
    dataset_ds = ds,
    new_var = EOSDT,
    dtc = DSSTDTC,
    filter = DSCAT == "DISPOSITION EVENT" & DSDECOD != "SCREEN FAILURE",
    date_imputation = NULL
  )
head(adsl %>% select(USUBJID, starts_with("EOS")))
```

This call would return the input dataset with the column `EOSDT` added. 
This function allows the user to impute partial dates as well. If imputation is needed and the date 
is to be imputed to the first of the month, then set `date_imputation = "FIRST"`.

### Disposition Status (e.g. `EOSTT`) {#disposition_status}

The function `derive_disposition_status()` can be used to derive a disposition status at a specific
timepoint. The relevant disposition variable (`ds.DSDECOD`) is selected by adjusting the filter 
parameter and used to derive `EOSSTT`.

To derive the End of Study status (`EOSSTT`), a call could be:

```{r eval=TRUE}
adsl <- adsl %>%
  derive_disposition_status(
    dataset_ds = ds,
    new_var = EOSSTT,
    status_var = DSDECOD,
    filter = DSCAT == "DISPOSITION EVENT"
  )
head(adsl %>% select(USUBJID, starts_with("EOS")))
```

This call would return the input dataset with the column `EOSSTT` added. 

By default, the function will derive `EOSSTT` as 

- `"COMPLETED"` if `DSDECOD == "COMPLETED"`
- `"DISCONTINUED"` if `DSDECOD` is not `"COMPLETED"` or `NA`
- `"ONGOING"` otherwise

If the default derivation must be changed, the user can create his/her own function and pass it to 
the `format_new_var` argument of the function (`format_new_var = new_mapping`) to map `DSDECOD` 
to a suitable `EOSSTT` value.

Example function `format_eosstt()`:

```{r eval=TRUE}
format_eosstt <- function(x) {
  case_when(
    x %in% c("COMPLETED") ~ "COMPLETED",
    x %in% c("SCREEN FAILURE") ~ NA_character_,
    !is.na(x) ~ "DISCONTINUED",
    TRUE ~ "ONGOING"
  )
}
```

The customized mapping function `format_eosstt()` can now be passed to the main function:

```{r eval=FALSE}

adsl <- adsl %>%
  derive_disposition_status(
    dataset_ds = ds,
    new_var = EOSSTT,
    status_var = DSDECOD,
    format_new_var = format_eosstt,
    filter = DSCAT == "DISPOSITION EVENT"
  )
```

This call would return the input dataset with the column `EOSSTT` added. 

### Disposition Reason(s) (e.g. `DCSREAS`, `DCSREASP`) {#disposition_reason}

The main reason for discontinuation is usually stored in `DSDECOD` while `DSTERM` provides 
additional details regarding subject’s discontinuation (e.g., description of `"OTHER"`).

The function `derive_disposition_reason()` can be used to derive a disposition reason (along with 
the details, if  required) at a specific timepoint. 
The relevant disposition variable(s) (`ds.DSDECOD`, `ds.DSTERM`) are selected by adjusting the 
filter parameter and used to derive the main reason (and details).

To derive the End of Study reason(s) (`DCSREAS` and `DCSREASP`), the call would be:

```{r eval=TRUE}
adsl <- adsl %>%
  derive_disposition_reason(
    dataset_ds = ds,
    new_var = DCSREAS,
    reason_var = DSDECOD,
    new_var_spe = DCSREASP,
    reason_var_spe = DSTERM,
    filter = DSCAT == "DISPOSITION EVENT" & DSDECOD != "SCREEN FAILURE"
  )
head(adsl %>% select(USUBJID, starts_with("EOS"), starts_with("DCSR")))
```

This call would return the input dataset with the column `DCSREAS` and `DCSREASP` added. 

By default, the function will map 

- `DCSREAS` as `DSDECOD` if `DSDECOD` is not `"COMPLETED"` or `NA`, `NA` otherwise
- `DCSREASP` as `DSTERM` if `DSDECOD` is not `"COMPLETED"` or `NA`, `NA` otherwise


If the default derivation must be changed, the user can create his/her own function and pass it to 
the `format_new_var` argument of the function (`format_new_var = new_mapping`) to map `DSDECOD` 
and  `DSTERM` to a suitable `DCSREAS`/`DCSREASP` value.

Example function `format_dcsreas()`:

```{r eval=TRUE}
format_dcsreas <- function(dsdecod, dsterm = NULL) {
  out <- if (is.null(dsterm)) dsdecod else dsterm
  case_when(
    dsdecod %notin% c("COMPLETED", "SCREEN FAILURE") & !is.na(dsdecod) ~ out,
    TRUE ~ NA_character_
  )
}
```

The customized mapping function `format_dcsreas()` can now be passed to the main function:

```{r eval=FALSE}
adsl <- adsl %>%
  derive_disposition_reason(
    dataset_ds = ds,
    new_var = DCSREAS,
    reason_var = DSDECOD,
    new_var_spe = DCSREASP,
    reason_var_spe = DSTERM,
    format_new_vars = format_dcsreas,
    filter_ds = DSCAT == "DISPOSITION EVENT"
  )
```

## Derive Death Variables {#death}

### Death Date (`DTHDT`) {#death_date}

The function `derive_vars_dt()` can be used to derive `DTHDT`. This function allows 
the user to impute the date as well.

Example calls:

```{r eval=TRUE}
adsl <- adsl %>%
  derive_vars_dt(
    new_vars_prefix = "DTH",
    dtc = DTHDTC
  )
adsl %>% select(USUBJID, starts_with("DTH")) %>% filter(DTHFL == "Y")
```

This call would return the input dataset with the columns `DTHDT` added and, by default, the 
associated date imputation flag (`DTHDTF`) populated with the controlled terminology outlined in 
the ADaM IG for date imputations. 
If the imputation flag is not required, the user must set the argument `flag_imputation` to `FALSE`.

If imputation is needed and the date is to be imputed to the first day of the month/year 
the call would be:

```{r eval=FALSE}
adsl <- adsl %>%
  derive_vars_dt(
    new_vars_prefix = "DTH",
    dtc = DTHDTC,
    date_imputation = "FIRST"
  )
```

See also [Date and Time Imputation](imputation.html).

### Cause of Death  (`DTHCAUS`) {#death_cause}

The cause of death `DTHCAUS` can be derived using the function `derive_var_dthcaus()`.

Since the cause of death could be collected/mapped in different domains (e.g. `DS`, `AE`, `DD`), it 
is important the user specifies the right source(s) to derive the cause of death from.

For example, if the date of death is collected in the AE form when the AE is Fatal, the cause of 
death would be set to the preferred term (`AEDECOD`) of that Fatal AE, while if the date of death is 
collected in  the `DS` form, the cause of death would be set to the disposition term (`DSTERM`).
To achieve this, the `dthcaus_source()` objects must be specified and defined such as it fits the 
study requirement. 

`dthcause_source()` specifications:

- `dataset`: the dataset to search for death information,
- `filter`: the condition to define death,
- `date_var`: the date of death,
- `mode`: `first` or `last` to select the first/last date of death if multiple dates are collected,
- `dthcaus`: variable  or text used to populate `DTHCAUS`.
- `traceability_vars`: whether the traceability variables need to be added (e.g source domain,
sequence, variable)

An example call to define the sources would be: 

```{r eval=TRUE}
src_ae <- dthcaus_source(
  dataset = ae,
  filter = AEOUT == "FATAL",
  date_var = AESTDTC,
  mode = "first",
  dthcaus = AEDECOD
)
src_ds <- dthcaus_source(
  dataset = ds,
  filter = DSDECOD == "DEATH" & grepl("DEATH DUE TO", DSTERM),
  date_var = DSSTDTC,
  mode = "first",
  dthcaus = "Death in DS"
)
```

Once the sources are defined, the function `derive_var_dthcaus()` can be used to derive `DTHCAUS`:

```{r eval=TRUE}
adsl <- adsl %>%
  derive_var_dthcaus(src_ae, src_ds)
adsl %>% select(USUBJID, starts_with("DTH")) %>% filter(DTHFL == "Y")
```

The function also offers the option to add some traceability variables (e.g. `DTHDOM` would 
store the domain where the date of death is collected, and `DTHSEQ` would store the `xxSEQ` value of 
that domain). To add them, the `traceabilty_vars()` argument must be added to the `dthcaus_source()`
arguments:

```{r eval=FALSE}
src_ae <- dthcaus_source(
  dataset = ae,
  filter = AEOUT == "FATAL",
  date_var = AESTDTC,
  mode = "first",
  dthcaus = AEDECOD,
  traceabilty_vars = vars(DTHDOM = "AE", DTHSEQ = AESEQ)
)

src_ds <- dthcaus_source(
  dataset = ds,
  filter = DSDECOD == "DEATH" & grepl("DEATH DUE TO", DSTERM),
  date = DSSTDTC,
  mode = "first",
  dthcaus = DSTERM,
  traceabilty_vars = vars(DTHDOM = "DS", DTHSEQ = DSSEQ)
)
adsl <- adsl %>%
  select(-DTHCAUS) %>% # remove it before deriving it again
  derive_var_dthcaus(src_ae, src_ds)

adsl %>% select(USUBJID, starts_with("DTH")) %>% filter(DTHFL == "Y")
```

### Duration Relative to Death {#death_other}

The function `derive_duration()` can be used to derive duration relative to death like the Relative 
Day of Death (`DTHADY`) or the numbers of days from last dose to death (`LDDTHELD`).

Example calls: 

- Relative Day of Death
```{r eval=TRUE}
adsl <- adsl %>%
  derive_duration(
    new_var = DTHADY,
    start_date = TRTSDT,
    end_date = DTHDT
  )
```

- Elapsed Days from Last Dose to Death
```{r eval=TRUE}
adsl <- adsl %>%
  derive_duration(
    new_var = LDDTHELD,
    start_date = TRTEDT,
    end_date = DTHDT,
    add_one = FALSE
  )

adsl %>% select(USUBJID, starts_with("DTH"), LDDTHELD) %>% filter(DTHFL == "Y")
```

## Derive the Last Date Known Alive (`LSTALVDT`) {#lstalvdt}

Similarly as for the cause of death (`DTHCAUS`), the last known alive date (`LSTALVDT`) can be
derived from multiples sources and the user must ensure the sources (`lstalvdt_source`) are 
correctly defined.

`lstalvdt_source()` specifications:

- `dataset`: the dataset to search for date information,
- `filter`: the filter to apply on the datasets,
- `date_var`: the date of interest,
- `date_imputation`: whether and how to impute partial dates,
- `traceability_vars`: whether the traceability variables need to be added (e.g source domain,
sequence, variable)

An example could be :

```{r eval=TRUE}
ae_src1 <- lstalvdt_source(
  dataset = ae,
  date_var = AESTDTC,
  date_imputation = "FIRST"
)
ae_src2 <- lstalvdt_source(
  dataset = ae,
  date_var = AEENDTC,
  date_imputation = "LAST"
)
lb_src <- lstalvdt_source(
  dataset = lb,
  date_var = LBDTC,
  filter = str_length(LBDTC) >= 10
)
adsl_src <- lstalvdt_source(
  dataset = adsl,
  date_var = TRTEDT
)
```

Once the sources are defined, the function `derive_var_lstalvdt()` can be used to derive `LSTALVDT`:

```{r eval=TRUE}
adsl <- adsl %>%
  derive_var_lstalvdt(ae_src1, ae_src2, lb_src, adsl_src)

head(adsl %>% select(USUBJID, TRTEDT, EOSDT, LSTALVDT))
```

Similarly to `dthcaus_source()`, the traceability variables can be added by specifying the 
`traceability_vars` argument in `lstalvdt_source()`.

```{r eval=TRUE}
ae_src1 <- lstalvdt_source(
  dataset = ae,
  date_var = AESTDTC,
  date_imputation = "FIRST",
  traceability_vars = vars(LALVDOM = "AE", LALVSEQ = AESEQ, LALVVAR = "AESTDTC")
)
ae_src2 <- lstalvdt_source(
  dataset = ae,
  date_var = AEENDTC,
  date_imputation = "LAST",
  traceability_vars = vars(LALVDOM = "AE", LALVSEQ = AESEQ, LALVVAR = "AEENDTC")
)
lb_src <- lstalvdt_source(
  dataset = lb,
  date_var = LBDTC,
  filter = str_length(LBDTC) >= 10,
  traceability_vars = vars(LALVDOM = "LB", LALVSEQ = LBSEQ, LALVVAR = "LBDTC")
)
adsl_src <- lstalvdt_source(
  dataset = adsl,
  date_var = TRTEDTM,
  traceability_vars = vars(LALVDOM = "ADSL", LALVSEQ = NA_integer_, LALVVAR = "TRTEDTM")
)

adsl <- adsl %>%
  select(-LSTALVDT) %>% # created in the previous call
  derive_var_lstalvdt(ae_src1, ae_src2, lb_src, adsl_src)

head(adsl %>% select(USUBJID, TRTEDT, EOSDT, LSTALVDT, starts_with("LAL")))
```
 

## Derive Groupings and Populations {#groupings}

### Grouping (e.g. `AGEGR1`) {#groupings_ex}

Numeric and categorical variables (`AGE`, `RACE`, `COUNTRY`, etc.) may need to be grouped to perform
the required analysis.
admiral does not **currently** have functionality to assist with the grouping. Instead, the user can 
create his/her own function to meet his/her study requirement.

For example, if 

- `AGEGR1` would categorize `AGE` in `< 18 `, `18-65`, `>= 65`, 
- `REGION1` would categorize `COUNTRY` in `North America`, `Rest of the World`, 

the user defined function(s) would be like:


```{r eval=TRUE}
format_agegr1 <- function(x) {
  case_when(
    !is.na(x) & x < 18 ~ "< 18",
    x >= 18 & x < 65 ~ "18 - 65",
    x >= 65 ~ ">= 65",
    TRUE ~ NA_character_
  )
}

format_region1 <- function(x) {
  case_when(
    x %in% c("CAN", "USA") ~ "North America",
    !is.na(x) ~ "Rest of the World",
    TRUE ~ "Missing"
  )
}
```

These functions are then used in a `mutate()` statement to derive the required grouping variables:

```{r eval=TRUE}
adsl <- adsl %>%
  mutate(
    AGEGR1 = format_agegr1(AGE),
    REGION1 = format_region1(COUNTRY)
  )
head(adsl %>% select(USUBJID, AGE, SEX, COUNTRY, AGEGR1, REGION1))
```

### Population Flags (e.g. `SAFFL`) {#popflag}

Since the populations flags are mainly company/study specific and that it can easily be derived
using an `if_else()` statement, admiral did not implement a functionality to assist with 
populations flags.

An example of a simple implementation could be:

```{r eval=TRUE}
adsl <- adsl %>%
  mutate(
    SAFFL = if_else(!is.na(TRTSDT), "Y", NA_character_)
  )
head(adsl %>% select(USUBJID, TRTSDT, SAFFL))
```


## Derive Other Variables  {#other}
The users can add specific code to cover their need for the analysis.

## Add Labels and Attributes {#attributes}
admiral does not **currently** support metadata integration, but the team is working on it!

## Example Script {#example}
ADaM | Sample Code
---- | --------------
ADSL | [ad_adsl.R](https://github.com/Roche-GSK/admiral/blob/master/inst/example_scripts/ad_adsl.R){target="_blank"} 

