---
title: "Creating a BDS Exposure ADaM"
output: 
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating a BDS Exposure ADaM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(admiraldev)
```

# Introduction

This article describes creating an Exposure ADaM using the BDS structure. 
Examples are currently presented using an underlying `EX` domain where the `EX` 
domain represents data as collected on the CRF and the `ADEX` ADaM is output. 
However, the examples can be applied to situations where an `EC` domain is
used as input and/or `ADEC` or another exposure ADaM is created.

There are many different approaches to modeling exposure data. This vignette 
gives examples of creating `PARAMCD` and `AVAL` combinations using exposure data.
This vignette is not meant to be a guide or standard for the structure of 
exposure analysis datasets.

**Note**: *All examples assume CDISC SDTM and/or ADaM format as input unless 
otherwise specified.*

# Programming Workflow

* [Read in Data](#readdata)
* [Derive/Impute Numeric Date/Time and Analysis Day (`ADT`, `ADTM`, `ADY`, `ADTF`, `ATMF`)](#datetime)
* [Compute Duration for a Record](#duration)
* [Create 1:1 Mapping Records](#onetoone)
* [Create Summary Records](#summaryrec)
* [Assign `PARAMCD`, `PARAMN`, etc. from Reference Tables](#paramcd)
* [Derive Categorization Variables (`AVALCATx`)](#cat)
* [Assign `ASEQ`](#aseq)
* [Add ADSL variables `ASEQ`](#adsl_vars)
* [Add Labels and Attributes](#attributes)

## Read in Data {#readdata}

To start, all data frames needed for the creation of `ADEX` should be read into
the environment.  This will be a company specific process.  Some of the 
data frames needed may be `EX` and `ADSL`.

For example purpose, the CDISC Pilot SDTM and ADaM datasets---which are included in `{admiral.test}`---are used.

```{r message=FALSE}
library(admiral)
library(dplyr, warn.conflicts = FALSE)
library(admiral.test)
library(lubridate)
library(stringr)
library(tibble)

data("admiral_adsl")
data("admiral_ex")

adsl <- admiral_adsl
ex <- admiral_ex
```
```{r echo=FALSE}
ex <- filter(ex, USUBJID %in% c("01-701-1015", "01-701-1023", "01-703-1086", "01-703-1096", "01-707-1037", "01-716-1024"))
```

At this step, it may be useful to join `ADSL` to your `EX` domain as well. Only the `ADSL` variables used for derivations are selected at this step. The rest of the relevant `ADSL` variables would be added later.


```{r eval=TRUE}

adsl_vars <- vars(TRTSDT, TRTSDTM, TRTEDT, TRTEDTM)

adex <- derive_vars_merged(
  ex,
  dataset_add = adsl,
  new_vars = adsl_vars,
  by_vars = vars(STUDYID, USUBJID)
)
```

```{r, eval=TRUE, echo=FALSE}
dataset_vignette(
  adex,
  display_vars = vars(
    USUBJID, EXTRT, EXDOSE, EXDOSFRQ,
    VISIT, EXSTDTC, EXENDTC,
    TRTSDTM, TRTEDTM
  )
)
```

The CDISC pilot `EX` domain data does not contain a dose adjustment flag or 
the planned dose information. For demonstration purposes, this will be added to 
the data.

```{r eval=TRUE, echo=TRUE}
adex <- adex %>%
  mutate(
    EXADJ = case_when(
      USUBJID == "01-701-1028" & VISIT %in% c("WEEK 2") ~ "ADVERSE EVENT",
      USUBJID == "01-701-1148" & VISIT %in% c("WEEK 2", "WEEK 24") ~ "MEDICATION ERROR",
      TRUE ~ NA_character_
    ),
    EXDOSE = case_when(
      USUBJID == "01-701-1028" & VISIT %in% c("WEEK 2") ~ 0,
      USUBJID == "01-701-1148" & VISIT %in% c("WEEK 2", "WEEK 24") ~ 0,
      TRUE ~ EXDOSE
    )
  ) %>%
  mutate(EXPLDOS = if_else(EXTRT == "PLACEBO", 0, 54))

distinct(adex, EXTRT, EXPLDOS)
count(adex, EXADJ)
```

## Derive/Impute Numeric Date/Time and Analysis Day (`ADT`, `ADTM`, `ADY`, `ADTF`, `ATMF`) {#datetime}

The function `derive_vars_dt()` can be used to derive `ADT`. This function allows
the user to impute the date as well.

Example calls:

```{r eval=TRUE, echo=TRUE}
adex <- derive_vars_dt(adex, new_vars_prefix = "AST", dtc = EXSTDTC)
adex <- derive_vars_dt(adex, new_vars_prefix = "AEN", dtc = EXENDTC)
```

```{r, eval=TRUE, echo=FALSE}
dataset_vignette(
  adex,
  display_vars = vars(USUBJID, VISIT, EXSTDTC, EXENDTC, ASTDT, AENDT)
)
```

The next examples demonstrates the datetime imputation features available
in the `derive_vars_dtm()` function, where the time is imputed as "00:00:00":

```{r eval=TRUE, echo=TRUE}
adex <- derive_vars_dtm(
  adex,
  dtc = EXSTDTC,
  highest_imputation = "M",
  new_vars_prefix = "AST"
)

adex <- derive_vars_dtm(
  adex,
  dtc = EXENDTC,
  highest_imputation = "M",
  date_imputation = "last",
  new_vars_prefix = "AEN"
)
```

```{r, eval=TRUE, echo=FALSE}
dataset_vignette(
  adex,
  display_vars = vars(USUBJID, VISIT, EXSTDTC, EXENDTC, ASTDTM, AENDTM)
)
```


The example above imputes the start date to the first first day of the month and
imputes the end date to the last day of the month.  

Please see the [Date and Time Imputation](imputation.Rmd) for additional examples on 
calculating and imputing analysis dates.

Next, the analysis study days can be derived:

```{r eval=TRUE, echo=TRUE}
adex <-
  derive_vars_dy(adex,
    reference_date = TRTSDT,
    source_vars = vars(ASTDT, AENDT)
  )
```

```{r, eval=TRUE, echo=FALSE}
dataset_vignette(
  adex,
  display_vars = vars(
    USUBJID,
    VISIT, ASTDT, ASTDY, AENDT, AENDY, TRTSDT
  )
)
```


## Compute duration for a record {#duration}

To compute the duration of treatment or exposure for a record, the
`derive_vars_duration()` function can be used.

```{r eval=TRUE, echo=TRUE}
adex <- adex %>%
  derive_vars_duration(
    new_var = EXDURD,
    start_date = ASTDT,
    end_date = AENDT
  )
```

```{r, eval=TRUE, echo=FALSE}
dataset_vignette(
  adex,
  display_vars = vars(
    USUBJID,
    VISIT, ASTDT, ASTDY, AENDT, AENDY, EXDURD
  )
)
```

The units of the calculated duration can also be changed. In this example, the
duration is output as years:

```{r eval=TRUE, echo=TRUE}
adex <- adex %>%
  derive_vars_duration(
    new_var = EXDURDY,
    out_unit = "years",
    start_date = ASTDT,
    end_date = AENDT
  )
```

```{r, eval=TRUE, echo=FALSE}
dataset_vignette(
  adex,
  display_vars = vars(
    USUBJID,
    VISIT, ASTDT, AENDT, EXDURD, EXDURDY
  )
)
```


Please refer to the `derive_vars_duration()` documentation for detailed information 
on the input parameters.

It may be necessary to calculate additional intermediate values.  For example, 
the cumulative doses received and cumulative planned doses may be calculated as:

```{r eval=TRUE, echo=TRUE}
adex <- adex %>%
  mutate(
    DOSEO = EXDOSE * EXDURD,
    PDOSEO = EXPLDOS * EXDURD
  )
```

```{r, eval=TRUE, echo=FALSE}
dataset_vignette(
  adex,
  display_vars = vars(USUBJID, EXDOSE, EXPLDOS, EXDURD, DOSEO, PDOSEO)
)
```

It may be of additional interest to turn a single record containing dosing
summary information into a set of multiple single records, each representing a
single dose over the interval specified by the summary record. This is another
approach to deriving a total dose parameter when `EXDOSFRQ != ONCE`.

The function `create_single_dose_dataset()` can 
be used to expand a record containing a start date, an end date, and a dosing
frequency to a corresponding set of records each representing one dose (i.e.
`EXDOSFRQ == "ONCE"`).

```{r eval=TRUE, echo=TRUE}
single_dose <- adex %>%
  filter(USUBJID == "01-701-1015" & EXSTDY == 1) %>%
  create_single_dose_dataset(keep_source_vars = vars(USUBJID, EXDOSE, EXPLDOS, EXDOSFRQ, ASTDT, AENDT))
```

```{r, eval=TRUE, echo=FALSE}
dataset_vignette(
  single_dose,
  display_vars = vars(USUBJID, EXDOSE, EXPLDOS, EXDOSFRQ, ASTDT, AENDT)
)
```

## Create 1:1 mapping records {#onetoone}

The first set of exposure records to create will be records mapped 1:1 to an
existing collected exposure record in SDTM.  For these records, the `AVAL` or
`AVALC` would be calculated using columns that exist on the data and no summarizing
of records would be necessary.

These records may be used for input into summary records or be used individually
for summarization in outputs.  Some examples may be exposure duration, dose
administered, dose adjusted, etc. based on one exposure record in SDTM.

These records can be derived using simple `dplyr::mutate` assignments and then combined:

```{r eval=TRUE, echo=TRUE}
adex_durd <- adex %>%
  mutate(
    PARAMCD = "DURD",
    AVAL = EXDURD
  )

adex_dose <- adex %>%
  mutate(
    PARAMCD = "DOSE",
    AVAL = DOSEO
  )

adex_pldos <- adex %>%
  mutate(
    PARAMCD = "PLDOSE",
    AVAL = PDOSEO
  )

adex_adj <- adex %>%
  mutate(
    PARAMCD = "ADJ",
    AVALC = if_else(!is.na(EXADJ), "Y", NA_character_)
  )

adex_adjae <- adex %>%
  mutate(
    PARAMCD = "ADJAE",
    AVALC = if_else(EXADJ == "ADVERSE EVENT", "Y", NA_character_)
  )

adex <- bind_rows(
  adex_durd,
  adex_dose,
  adex_pldos,
  adex_adj,
  adex_adjae
) %>%
  mutate(PARCAT1 = "INDIVIDUAL")

count(adex, PARAMCD)
```

```{r, eval=TRUE, echo=FALSE}
adex %>%
  arrange(USUBJID, VISIT, desc(PARAMCD), EXSTDTC, EXENDTC) %>%
  dataset_vignette(display_vars = vars(USUBJID, VISIT, ASTDT, AENDT, PARAMCD, AVAL, AVALC))
```

## Create Summary Records {#summaryrec}

Exposure is commonly analyzed by a timing interval (e.g. `APHASE`, `APERIOD`,
`AVISIT`, etc.). For these types of calculations, the `derive_param_exposure()`
function may be used. In addition to creating a summarized `AVAL`, the function
will also compute minimum and maximum dates for the record. 

For example, to calculate the total dose by subject and treatment,

```{r eval=TRUE, echo=TRUE}
adex <- derive_param_exposure(
  adex,
  by_vars = vars(STUDYID, USUBJID, !!!adsl_vars),
  input_code = "DOSE",
  analysis_var = AVAL,
  set_values_to = vars(PARAMCD = "TDOSE", PARCAT1 = "OVERALL"),
  summary_fun = function(x) sum(x, na.rm = TRUE)
)
```

```{r, eval=TRUE, echo=FALSE}
adex %>%
  arrange(USUBJID, PARAMCD, PARCAT1, VISIT, EXSTDTC, EXENDTC) %>%
  dataset_vignette(display_vars = vars(
    USUBJID, VISIT,
    PARCAT1, PARAMCD, AVAL, ASTDT, AENDT
  ))
```


```{r eval=TRUE, echo=FALSE}
adex <- filter(adex, PARAMCD != "TDOSE")
```

A record with `PARAMCD == "TDOSE"` is created with `PARCAT1` set to `"OVERALL"` using
the records in `ADEX` where `PARAMCD == "DOSE"` by summing `AVAL`. In addition,
the `ASTDT`, and `AENDT` are created as the minimum and
maximum date/times associated with each `by_vars` grouping. Note that, in addition
to `PARAMCD`, `PARCAT1`, `AVAL`, `ASTDT` and `AENDT`, only those variables specified in the `by_vars` argument will be populated in the new records.

Multiple parameters (records) may be created at one time using the 
`call_derivation()` function:

```{r eval=TRUE, echo=TRUE}
adex <- adex %>%
  call_derivation(
    derivation = derive_param_exposure,
    variable_params = list(
      params(
        set_values_to = vars(PARAMCD = "TDOSE", PARCAT1 = "OVERALL"),
        input_code = "DOSE",
        analysis_var = AVAL,
        summary_fun = function(x) sum(x, na.rm = TRUE)
      ),
      params(
        set_values_to = vars(PARAMCD = "TPDOSE", PARCAT1 = "OVERALL"),
        input_code = "PLDOSE",
        analysis_var = AVAL,
        summary_fun = function(x) sum(x, na.rm = TRUE)
      ),
      params(
        set_values_to = vars(PARAMCD = "TDURD", PARCAT1 = "OVERALL"),
        input_code = "DURD",
        analysis_var = AVAL,
        summary_fun = function(x) sum(x, na.rm = TRUE)
      ),
      params(
        set_values_to = vars(PARAMCD = "TADJ", PARCAT1 = "OVERALL"),
        input_code = "ADJ",
        analysis_var = AVALC,
        summary_fun = function(x) if_else(sum(!is.na(x)) > 0, "Y", NA_character_)
      ),
      params(
        set_values_to = vars(PARAMCD = "TADJAE", PARCAT1 = "OVERALL"),
        input_code = "ADJAE",
        analysis_var = AVALC,
        summary_fun = function(x) if_else(sum(!is.na(x)) > 0, "Y", NA_character_)
      )
    ),
    by_vars = vars(STUDYID, USUBJID, !!!adsl_vars)
  )

count(adex, PARAMCD, PARCAT1)
```


```{r, eval=TRUE, echo=FALSE}
adex %>%
  arrange(USUBJID, PARAMCD, PARCAT1, VISIT, EXSTDTC, EXENDTC) %>%
  dataset_vignette(display_vars = vars(
    USUBJID, VISIT,
    PARCAT1, PARAMCD, AVAL, AVALC, ASTDT, AENDT
  ))
```

Dose intensity can be calculated using the function `derive_param_doseint()`. The
planned dose and administered dose are passed into the function and a new record
is created with the dose intensity calculation. Again, only those variables 
specified in the `by_vars` argument will be populated in this new record.

```{r eval=TRUE, echo=TRUE}
adex <- adex %>%
  derive_param_doseint(
    by_vars = vars(STUDYID, USUBJID, !!!adsl_vars),
    set_values_to = vars(PARAMCD = "TNDOSINT"),
    tadm_code = "TDOSE",
    tpadm_code = "TPDOSE"
  )
```


```{r, eval=TRUE, echo=FALSE}
dataset_vignette(
  adex,
  display_vars = vars(
    USUBJID, VISIT, EXSTDTC, EXENDTC,
    PARCAT1, PARAMCD, AVAL, ASTDT, AENDT
  )
)
```


The default calculation for dose intensity is:
Administered Doses / Planned Doses * 100.

Please see the `derive_param_doseint()` documentation to see how planned doses of
0 or `NA` are handled.

## Assign `PARAMCD`, `PARAMN`, etc. from Reference tables {#paramcd}

To assign parameter level values such as `PARAM`, `PARAMN`, `PARCAT1`, etc., a
lookup can be created to join to the source data.

For example, when creating `ADEX`, a lookup based on the ADaM `PARAMCD` value
may be created:

PARAMCD | PARAM | PARAMN
------- | ----- | ------
DURD | Study drug duration during constant dosing interval (days) | 1
DOSE | Dose administered during constant dosing interval (mg) | 2
PLDOSE | Planned dose during constant dosing interval (mg) | 3
ADJ | Dose adjusted during constant dosing interval | 4
ADJAE | Dose adjusted  due to AE during constant dosing interval | 5
TDURD | Overall duration (days) | 6
TDOSE | Total dose administered (mg) | 7
TPDOSE | Total planned dose (mg) | 9
TADJ | Dose adjusted during study | 10
TADJAE | Dose adjusted during study due to AE | 11
TNDOSINT | Overall dose intensity (%) | 12

```{r eval=TRUE, include=FALSE, echo=FALSE}
param_lookup <- tribble(
  ~PARAMCD, ~PARAM, ~PARAMN,
  "DURD", "Study drug duration during constant dosing interval (days)", 1,
  "DOSE", "Dose administered during constant dosing interval (mg)", 2,
  "PLDOSE", "Planned dose during constant dosing interval (mg)", 3,
  "ADJ", "Dose adjusted during constant dosing interval", 4,
  "ADJAE", "Dose adjusted  due to AE during constant dosing interval", 5,
  "TDURD", "Overall duration (days)", 6,
  "TDOSE", "Total dose administered (mg)", 7,
  "TPDOSE", "Total planned dose (mg)", 9,
  "TADJ", "Dose adjusted during study", 10,
  "TADJAE", "Dose adjusted during study due to AE", 11,
  "TNDOSINT", "Overall dose intensity (%)", 12
)
```


```{r eval=TRUE, echo=TRUE}
adex <- derive_vars_merged(
  adex,
  dataset_add = param_lookup,
  by_vars = vars(PARAMCD)
)

count(adex, PARAMCD, PARAM, PARAMN)
```

Please note, this is an example only and additional columns may be needed for
the join depending on your lookup/metadata table.

## Derive Categorization Variables (`AVALCATx`) {#cat}

`{admiral}` does not currently have a generic function to aid in assigning `AVALCATx`/
`AVALCAxN` values. Below is a simple example of how these values may be
assigned using the `dplyr::mutate` function:

```{r eval=TRUE, echo=TRUE}
adex <- adex %>%
  mutate(
    AVALCAT1 = case_when(
      PARAMCD %in% c("TDURD") & AVAL < 30 ~ "< 30 days",
      PARAMCD %in% c("TDURD") & AVAL >= 30 & AVAL < 90 ~ ">= 30 and < 90 days",
      PARAMCD %in% c("TDURD") & AVAL >= 90 ~ ">=90 days",
      PARAMCD %in% c("TDOSE", "TPDOSE") & AVAL < 1000 ~ "< 1000 mg",
      PARAMCD %in% c("TDOSE", "TPDOSE") & AVAL >= 1000 ~ ">= 1000 mg",
      TRUE ~ NA_character_
    )
  )
```

```{r, eval=TRUE, echo=FALSE}
adex %>%
  arrange(USUBJID, AVALCAT1, PARCAT1, VISIT, EXSTDTC, EXENDTC) %>%
  dataset_vignette(display_vars = vars(USUBJID, VISIT, PARCAT1, PARAMCD, AVAL, AVALCAT1))
```

## Assign `ASEQ` {#aseq}

The `{admiral}` function `derive_var_obs_number()` can be used to derive `ASEQ`. An
example call is:

```{r eval=TRUE, echo=TRUE}
adex <- derive_var_obs_number(
  adex,
  new_var = ASEQ,
  by_vars = vars(STUDYID, USUBJID),
  order = vars(PARCAT1, ASTDT, VISIT, VISITNUM, EXSEQ, PARAMN),
  check_type = "error"
)
```

```{r, eval=TRUE, echo=FALSE}
dataset_vignette(
  adex,
  display_vars = vars(USUBJID, VISIT, PARCAT1, PARAMCD, AVAL, ASTDT, ASEQ)
)
```

## Add the `ADSL` variables  {#adsl_vars}

If needed, the other `ADSL` variables can now be added:

```{r eval=TRUE, echo=TRUE}
adex <- adex %>%
  derive_vars_merged(
    dataset_add = select(adsl, !!!negate_vars(adsl_vars)),
    by_vars = vars(STUDYID, USUBJID)
  )
```
## Add Labels and Attributes {#attributes}

Adding labels and attributes for SAS transport files is supported by the
following packages:

- [metacore](https://atorus-research.github.io/metacore/): establish a common
foundation for the use of metadata within an R session.

- [metatools](https://pharmaverse.github.io/metatools/): enable the use of
metacore objects. Metatools can be used to build datasets or enhance columns in
existing datasets as well as checking datasets against the metadata.

- [xportr](https://atorus-research.github.io/xportr/): functionality to
associate all metadata information to a local R data frame, perform data set
level validation checks and convert into a [transport v5
file(xpt)](https://documentation.sas.com/doc/en/pgmsascdc/9.4_3.5/movefile/n1xbwdre0giahfn11c99yjkpi2yb.htm).

NOTE: All these packages are in the experimental phase, but the vision is to
have them associated with an End to End pipeline under the umbrella of the
[pharmaverse](https://github.com/pharmaverse). An example of applying metadata 
and perform associated checks can be found at the [pharmaverse E2E example](https://examples.pharmaverse.org/data/adsl/).

# Example Scripts {#example}

ADaM | Sample Code
---- | --------------
ADEX | [ad_adex.R](https://github.com/pharmaverse/admiral/blob/main/inst/templates/ad_adex.R){target="_blank"}
