---
title: "Creating an Exposure (BDS) ADaM"
output: 
  rmarkdown::html_vignette:
    toc: false
vignette: >
  %\VignetteIndexEntry{Creating an Exposure (BDS) ADaM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

This article describes creating an Exposure ADaM using the BDS structure. 
Examples are currently presented using an underlying EX domain where the EX 
domain represents data as collected on the CRF and the ADEX ADaM is output. 
However, the examples can be applied to situations where an EC domain is
used as input and/or ADEC or another exposure ADaM is created.

There are many different approaches to modeling exposure data. This vignette 
gives examples of creating `PARAMCD` and `AVAL` combinations using exposure data.
This vignette is not meant to be a guide or standard for the structure of 
exposure analysis datasets.

**Note**: *All examples assume CDISC SDTM and/or ADaM format as input unless 
otherwise specified.*

## Programming Workflow

* [Read in Data](#readdata)
* [Derive/Impute Numeric Date/Time and Analysis Day (`ADT`, `ADTM`, `ADY`, `ADTF`, `ATMF`)](#datetime)
* [Compute Duration for a Record](#duration)
* [Create 1:1 Mapping Records](#onetoone)
* [Create Summary Records](#summaryrec)
* [Assign `PARAMCD`, `PARAMN`, etc. from Reference Tables](#paramcd)
* [Assign `ASEQ`](#aseq)
* [Derive Categorization Variables (`AVALCATx`)](#cat)
* [Example Scripts](#example)

### Read in Data {#readdata}

To start, all data frames needed for the creation of ADEX should be read into
the environment.  This will be a company specific process.  Some of the 
data frames needed may be `ex`, `suppex`, and `adsl`.

For example purpose, the CDISC Pilot SDTM and ADaM datasets---which are included in `{admiral}`---are used.

```{r message=FALSE}
library(admiral)
library(dplyr)
library(cdiscpilot)
library(lubridate)
library(stringr)
library(tibble)

data("adsl")
data("ex")
```

The `suppex` domain can be joined to the `ex` domain using the function `derive_vars_suppqual()`.

This function will transpose the supplemental SDTM domain (e.g. `suppex`) and
join the transposed data to the parent domain (e.g. `ex`) by `STUDYID`, 
`USUBJID` using the `IDVAR` and `IDVARVAL` as an additional join variable.

Example call:

```{r eval=FALSE}
ex <- derive_vars_suppqual(ex, suppex)
```

Please note, the CDISC pilot did not include a `suppex` dataset and therefore this
join is not demonstrated.

At this step, it may be useful to join `adsl` to your `ex` domain as well:

```{r eval=TRUE}
adex <- left_join(
  ex,
  select(adsl, STUDYID, USUBJID, starts_with("TRT")),
  by = c("STUDYID", "USUBJID")
)

head(adex)
```

The CDISC pilot `ex` domain data does not contain a dose adjustment flag or 
the planned dose information. For demonstration purposes, this will be added to 
the data.

```{r eval=TRUE, echo=TRUE}
adex <- adex %>%
  mutate(
    EXADJ = case_when(
      USUBJID == "01-701-1034" & VISIT %in% c("WEEK 2", "WEEK 24") ~ "ADVERSE EVENT",
      USUBJID == "01-701-1148" & VISIT %in% c("WEEK 24") ~ "MEDICATION ERROR",
      TRUE ~ NA_character_
    ),
    EXDOSE = case_when(
      USUBJID == "01-701-1034" & VISIT %in% c("WEEK 2", "WEEK 24") ~ 0,
      USUBJID == "01-701-1148" & VISIT %in% c("WEEK 24") ~ 0,
      TRUE ~ EXDOSE
    )
  ) %>%
  mutate(EXPLDOS = if_else(EXTRT == "PLACEBO", 0, 54))

distinct(adex, EXTRT, EXPLDOS)
count(adex, EXADJ)
```

### Derive/Impute Numeric Date/Time and Analysis Day (`ADT`, `ADTM`, `ADY`, `ADTF`, `ATMF`) {#datetime}

The function `derive_vars_dt()` can be used to derive `ADT`. This function allows
the user to impute the date as well.

Example calls:

```{r eval=TRUE, echo=TRUE}
adex <- derive_vars_dt(adex, new_vars_prefix = "AST", dtc = EXSTDTC)
adex <- derive_vars_dt(adex, new_vars_prefix = "AEN", dtc = EXENDTC)

head(select(adex, EXSTDTC, EXENDTC, ASTDT, AENDT))
```

```{r eval=TRUE, echo=TRUE}
adex <- select(adex, -ASTDTF, -AENDTF)
```


This next examples demonstrates the datetime imputation features available
in the `derive_vars_dtm` function where the time is imputed as "00:00:00":

```{r eval=TRUE, echo=TRUE}
adex <- derive_vars_dtm(
  adex,
  dtc = EXSTDTC,
  date_imputation = "first",
  new_vars_prefix = "AST"
)

adex <- derive_vars_dtm(
  adex,
  dtc = EXENDTC,
  date_imputation = "last",
  new_vars_prefix = "AEN"
)

head(select(adex, EXSTDTC, EXENDTC, ASTDT, ASTDTM, AENDT, AENDTM))
```

The example above imputes the start date to the first first day of the month and
imputes the end date to the last day of the month.  

Please see the [Date and Time Imputation](imputation.Rmd) for additional examples on 
calculating and imputing analysis dates.

Next, the analysis study days can be derived:

```{r eval=TRUE, echo=TRUE}
adex <- derive_var_astdy(adex, reference_date = TRTSDT, date = ASTDT)
adex <- derive_var_aendy(adex, reference_date = TRTSDT, date = AENDT)

head(distinct(adex, EXSTDTC, EXENDTC, ASTDT, ASTDY, AENDT, AENDY, TRTSDT))
```

### Compute duration for a record {#duration}

To compute the duration of treatment or exposure for a record, the
`derive_vars_duration()` function can be used.

```{r eval=TRUE, echo=TRUE}
adex <- adex %>%
  derive_vars_duration(
    new_var = EXDURD,
    start_date = ASTDT,
    end_date = AENDT
  )

head(distinct(adex, ASTDT, AENDT, EXDURD))
```

The units of the calculated duration can also be changed. In this example, the
duration is output as years:

```{r eval=TRUE, echo=TRUE}
adex <- adex %>%
  derive_vars_duration(
    new_var = EXDURDY,
    out_unit = "years",
    start_date = ASTDT,
    end_date = AENDT
  )

head(distinct(adex, ASTDT, AENDT, EXDURD, EXDURDY))
```

Please refer to the `derive_vars_duration()` documentation for detailed information on 
the input parameters.

It may be necessary to calculate additional intermediate values.  For example, 
the cumulative doses received and cumulative planned doses may be calculated as:

```{r eval=TRUE, echo=TRUE}
adex <- adex %>%
  mutate(
    DOSEO = EXDOSE * EXDURD,
    PDOSEO = EXPLDOS * EXDURD
  )

head(distinct(adex, EXDOSE, EXPLDOS, EXDURD, DOSEO, PDOSEO))
```


### Create 1:1 mapping records {#onetoone}

The first set of exposure records to create will be records mapped 1:1 to an
existing collected exposure record in SDTM.  For these records, the `AVAL` or
`AVALC` would be calculated using columns that exist on the data and no summarising
of records would be necessary.

These records may be used for input into summary records or be used individually
for summarization in outputs.  Some examples may be exposure duration, dose
administered, dose adjusted, etc. based on one exposure record in SDTM.

These records can be derived using simple `dplyr::mutate` assignments and then combined:

```{r eval=TRUE, echo=TRUE}
adex_durd <- adex %>%
  mutate(
    PARAMCD = "DURD",
    AVAL = EXDURD
  )

adex_dose <- adex %>%
  mutate(
    PARAMCD = "DOSE",
    AVAL = DOSEO
  )

adex_pldos <- adex %>%
  mutate(
    PARAMCD = "PLDOSE",
    AVAL = PDOSEO
  )

adex_adj <- adex %>%
  mutate(
    PARAMCD = "ADJ",
    AVALC = if_else(!is.na(EXADJ), "Y", NA_character_)
  )

adex_adjae <- adex %>%
  mutate(
    PARAMCD = "ADJAE",
    AVALC = if_else(EXADJ == "ADVERSE EVENT", "Y", NA_character_)
  )

adex <- bind_rows(
  adex_durd,
  adex_dose,
  adex_pldos,
  adex_adj,
  adex_adjae
) %>%
  mutate(PARCAT1 = "INDIVIDUAL")

count(adex, PARAMCD)
head(distinct(adex, PARAMCD, AVAL))
```

### Create Summary Records {#summaryrec}

Exposure is commonly analyzed by a timing interval (e.g. `APHASE`, `APERIOD`,
`AVISIT`, etc.). For these types of calculations, the `derive_params_exposure()`
function may be used. In addition to creating a summarized `AVAL`, the function
will also compute minimum and maximum dates for the record.

For example, to calculate the total dose by subject and treatment,

```{r eval=TRUE, echo=TRUE}
adex <- derive_params_exposure(
  adex,
  by_vars = vars(STUDYID, USUBJID, TRT01P, TRT01A, TRTSDTM, TRTEDTM, TRTSDT, TRTEDT),
  input_code = "DOSE",
  analysis_var = AVAL,
  set_values_to = vars(PARAMCD = "TDOSE", PARCAT1 = "OVERALL"),
  summary_fun = function(x) sum(x, na.rm = TRUE)
)

head(select(filter(adex, PARAMCD == "TDOSE"), PARAMCD, PARCAT1, AVAL, ASTDT, AENDT))
```

```{r eval=TRUE, echo=FALSE}
adex <- filter(adex, PARAMCD != "TDOSE")
```

A record with `PARAMCD == "TDOSE"` is created with `PARCAT1` set to `"OVERALL"` using
the records in `ADEX` where `PARAMCD == "DOSE"` by summing `AVAL`. In addition,
the `ASTDT`, and `AENDT` are created as the minimum and
maximum date/times associated with each `by_vars` grouping.

Multiple parameters (records) may be created at one time using the
`call_derivation()` function:

```{r eval=TRUE, echo=TRUE}
adex <- adex %>%
  call_derivation(
    derivation = derive_params_exposure,
    variable_params = list(
      params(
        set_values_to = vars(PARAMCD = "TDOSE", PARCAT1 = "OVERALL"),
        input_code = "DOSE",
        analysis_var = AVAL,
        summary_fun = function(x) sum(x, na.rm = TRUE)
      ),
      params(
        set_values_to = vars(PARAMCD = "TPDOSE", PARCAT1 = "OVERALL"),
        input_code = "PLDOSE",
        analysis_var = AVAL,
        summary_fun = function(x) sum(x, na.rm = TRUE)
      ),
      params(
        set_values_to = vars(PARAMCD = "TDURD", PARCAT1 = "OVERALL"),
        input_code = "DURD",
        analysis_var = AVAL,
        summary_fun = function(x) sum(x, na.rm = TRUE)
      ),
      params(
        set_values_to = vars(PARAMCD = "TADJ", PARCAT1 = "OVERALL"),
        input_code = "ADJ",
        analysis_var = AVALC,
        summary_fun = function(x) if_else(sum(!is.na(x)) > 0, "Y", NA_character_)
      ),
      params(
        set_values_to = vars(PARAMCD = "TADJAE", PARCAT1 = "OVERALL"),
        input_code = "ADJAE",
        analysis_var = AVALC,
        summary_fun = function(x) if_else(sum(!is.na(x)) > 0, "Y", NA_character_)
      )
    ),
    by_vars = vars(STUDYID, USUBJID, TRT01P, TRT01A, TRTSDTM, TRTEDTM, TRTSDT, TRTEDT)
  )

count(adex, PARAMCD, PARCAT1)
head(select(adex, USUBJID, PARAMCD, PARCAT1, AVALC, AVAL))
```

Dose intensity can be calculated using the function `derive_param_doseint()`. The
planned dose and administered dose are passed into the function and a new record
is created with the dose intensity calculation.

```{r eval=TRUE, echo=TRUE}
adex <- adex %>%
  derive_param_doseint(
    by_vars = vars(USUBJID),
    set_values_to = vars(PARAMCD = "TNDOSINT"),
    tadm_code = "TDOSE",
    tpadm_code = "TPDOSE"
  )

adex %>% 
  filter(PARAMCD %in% c("TNDOSINT", "TDOSE", "TPDOSE")) %>% 
  select(USUBJID, PARAMCD, AVAL) %>% 
  arrange(USUBJID) %>% 
  head()
```

The default calculation for dose intensity is:
Administered Doses / Planned Doses * 100.

Please see the `derive_param_doseint()` documentation to see how planned doses of
0 or `NA` are handled.

### Assign `PARAMCD`, `PARAMN`, etc. from Reference tables {#paramcd}

To assign parameter level values such as `PARAM`, `PARAMN`, `PARCAT1`, etc., a
lookup can be created to join to the source data.

For example, when creating `adex`, a lookup based on the ADaM `PARAMCD` value
may be created:

PARAMCD | PARAM | PARAMN
------- | ----- | ------
DURD | Study drug duration during constant dosing interval (days) | 1
DOSE | Dose administered during constant dosing interval (mg) | 2
PLDOSE | Planned dose during constant dosing interval (mg) | 3
ADJ | Dose adjusted during constant dosing interval | 4
ADJAE | Dose adjusted  due to AE during constant dosing interval | 5
TDURD | Overall duration (days) | 6
TDOSE | Total dose administered (mg) | 7
TPDOSE | Total planned dose (mg) | 9
TADJ | Dose adjusted during study | 10
TADJAE | Dose adjusted during study due to AE | 11
TNOSINT | Overall dose intensity (%) | 12

```{r eval=TRUE, include=FALSE}
param_lookup <- tribble(
  ~PARAMCD, ~PARAM, ~PARAMN,
  "DURD", "Study drug duration during constant dosing interval (days)", 1,
  "DOSE", "Dose administered during constant dosing interval (mg)", 2,
  "PLDOSE", "Planned dose during constant dosing interval (mg)", 3,
  "ADJ", "Dose adjusted during constant dosing interval", 4,
  "ADJAE", "Dose adjusted  due to AE during constant dosing interval", 5,
  "TDURD", "Overall duration (days)", 6,
  "TDOSE", "Total dose administered (mg)", 7,
  "TPDOSE", "Total planned dose (mg)", 9,
  "TADJ", "Dose adjusted during study", 10,
  "TADJAE", "Dose adjusted during study due to AE", 11,
  "TNDOSINT", "Overall dose intensity (%)", 12
)

head(param_lookup)
```

```{r eval=TRUE, echo=TRUE}
adex <- left_join(adex, param_lookup, by = "PARAMCD")

count(adex, PARAMCD, PARAM, PARAMN)
```

Please note, this is an example only and additional columns may be needed for
the join depending on your lookup/metadata table.

### Assign `ASEQ` {#aseq}

The admiral function `derive_obs_number()` can be used to derive `ASEQ`. An
example call is:

```{r eval=TRUE, echo=TRUE}
adex <- derive_obs_number(
  adex,
  new_var = ASEQ,
  by_vars = vars(STUDYID, USUBJID),
  order = vars(PARAMCD, ASTDT),
  check_type = "error"
)

head(select(adex, USUBJID, PARAMCD, ASTDT, ASEQ))
```

### Derive Categorization Variables (`AVALCATx`) {#cat}

Admiral does not currently have a generic function to aid in assigning `AVALCATx`/
`AVALCAxN` values. Below is a simple example of how these values may be
assigned using the `dplyr::mutate` function:

```{r eval=TRUE, echo=TRUE}
adex <- adex %>%
  mutate(
    AVALCAT = case_when(
      PARAMCD %in% c("TDURD", "PDURD") & AVAL < 30 ~ "< 30 days",
      PARAMCD %in% c("TDURD", "PDURD") & AVAL >= 30 & AVAL < 90 ~ ">= 30 and < 90 days",
      PARAMCD %in% c("TDURD", "PDURD") & AVAL >= 90 ~ ">=90 days",
      PARAMCD %in% c("TDOSE", "PDOSE") & AVAL < 100 ~ "< 100 mg",
      PARAMCD %in% c("TDOSE", "PDOSE") & AVAL >= 100 ~ ">= 100 mg",
      TRUE ~ NA_character_
    )
  )

distinct(filter(adex, PARAMCD %in% c("TDURD", "PDURD")), PARAMCD, AVAL, AVALCAT)
```

### Example Scripts {#example}

ADaM | Sample Code
---- | --------------
ADEX | [ad_adex.R](https://github.com/Roche-GSK/admiral/blob/master/inst/templates/ad_adex.R){target="_blank"}
