---
title: "Creating a BDS Finding ADaM"
output: 
  rmarkdown::html_vignette:
    toc: false
vignette: >
  %\VignetteIndexEntry{Creating a BDS Finding ADaM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

This article describes creating a BDS finding ADaM. Examples are currently 
presented and tested in the context of ADVS.  However, the examples could be
applied to other BDS Finding ADaMs such as ADEG, ADLB, etc. where a single 
result is captured in an SDTM Finding domain on a single date and/or time.

**Note**: *All examples assume CDISC SDTM and/or ADaM format as input unless 
otherwise specified.*

## Programming Workflow

* [Read in Data](#readdata)
* [Derive/Impute Numeric Date/Time and Analysis Day (`ADT`, `ADTM`, `ADY`, `ADTF`, `ATMF`)](#datetime)
* [Assign `PARAMCD`, `PARAM`, `PARAMN`, `PARCATy`](#paramcd)
* [Derive Results (`AVAL`, `AVALC`, `AVALU`)](#aval)
* [Derive Additional Parameters (e.g. `BMI` or `MAP` for ADVS)](#derive_param)
* [Derive Timing Variables (e.g. `APHASE`, `AVISIT`, `APERIOD`)](#timing)
* [Timing Flag Variables (e.g. `ONTRTFL`)](#timingflag)
* [Assign Reference Range Indicator (`ANRIND`)](#referencerange)
* [Derive Baseline (`BASETYPE`, `ABLFL`, `BASE`, `BASEC`, `BNRIND`)](#baseline)
* [Derive Change from Baseline (`CHG`, `PCHG`)](#bchange)
* [Derive Analysis Flags (e.g. `ANL01FL`)](#analysisrec)
* [Assign Treatment (`TRTA`, `TRTP`)](#treatment)
* [Assign `ASEQ`](#aseq)
* [Derive New Rows](#additional)
* [Derive Categorization Variables (`AVALCATx`)](#cat)
* [Example Scripts](#example)

### Read in Data {#readdata}

To start, all data frames needed for the creation of ADVS should be read into
the environment.  This will be a company specific process.  Some of the 
data frames needed may be `vs`, `suppvs`, and `adsl`.

For example purpose, the CDISC Pilot SDTM and ADaM datasets---which are included in `{admiral}`---are used.

```{r message=FALSE}
library(dplyr)
library(admiral)
library(lubridate)
library(stringr)
library(tibble)

data("adsl")
data("vs")
```

The `suppvs` domain can be joined to the `vs` domain using the function `derive_vars_suppqual()`.

This function will transpose the supplemental SDTM domain (e.g. `suppvs`) and
join the transposed data to the parent domain (e.g. `vs`) by `STUDYID`, 
`USUBJID` using the `IDVAR` and `IDVARVAL` as an additional join variable.

Example call:

```{r eval=FALSE}
vs <- derive_vars_suppqual(vs, suppvs)
```

Please note, the CDISC pilot did not include a `suppvs` dataset and therefore this
join is not demonstrated.

At this step, it may be useful to join ADSL to your VS domain as well:

```{r eval=TRUE}
advs <- left_join(vs, adsl, by = c("STUDYID", "USUBJID"))

head(advs)
```

### Derive/Impute Numeric Date/Time and Analysis Day (`ADT`, `ADTM`, `ADY`, `ADTF`, `ATMF`) {#datetime}

The function `derive_vars_dt()` can be used to derive `ADT`. This function allows 
the user to impute the date as well.

Example calls:

```{r eval=TRUE}
advs <- derive_vars_dt(advs, new_vars_prefix = "A", dtc = VSDTC)

head(select(advs, USUBJID, VSDTC, ADT, ADTF))
```

If imputation is needed and the date is to be imputed to the first of the month, 
the call would be:

```{r eval=TRUE, include=FALSE}
advs_old <- advs

advs <- advs %>% 
  mutate(
    VSDTC = if_else(
      USUBJID == "01-716-1024" & VISIT == "SCREENING 1",
      "2012-07",
      VSDTC
    )
  ) %>% 
  select(-ADT, -ADTF)
```

```{r eval=TRUE}
advs <- derive_vars_dt(
  advs, 
  new_vars_prefix = "A", 
  dtc = VSDTC, 
  date_imputation = "FIRST"
)

head(select(filter(advs, USUBJID == "01-716-1024"), USUBJID, VISIT, VSDTC, ADT, ADTF))
```

```{r eval=TRUE, echo=FALSE}
advs <- advs_old
```

Similarly, `ADTM` may be created using the function `derive_vars_dtm()`. 
Imputation may be done on both the date and time components of `ADTM`.

```{r eval=FALSE}
# CDISC Pilot data does not contain times and the output of the derivation
# ADTM is not presented.
advs <- derive_vars_dtm(
  advs, 
  new_vars_prefix = "A", 
  dtc = VSDTC, 
  date_imputation = "FIRST"
)
```

By default, the variable `ADTF` for `derive_vars_dt()` or `ADTF` and `ATMF` for 
`derive_vars_dtm()` will be created and populated with the controlled
terminology outlined in the ADaM IG for date imputations. 

See also [Date and Time Imputation](imputation.html).

Once `ADT` is derived, the function `derive_var_ady()` can be used to derive `ADY`.
This example assumes both `ADT` and `TRTSDT` exist on the data frame.

```{r eval=TRUE}
advs <- derive_var_ady(advs, reference_date = TRTSDT, date = ADT)

head(select(advs, USUBJID, TRTSDT, ADT, ADY))
```

### Assign `PARAMCD`, `PARAM`, `PARAMN`, `PARCAT1` {#paramcd}

To assign parameter level values such as `PARAMCD`, `PARAM`, `PARAMN`, `PARCAT1`,
etc., a lookup can be created to join to the source data.

For example, when creating `advs`, a lookup based on the SDTM `--TESTCD` value 
may be created:

VSTESTCD | PARAMCD | PARAM | PARAMN | PARCAT1 | PARCAT1N
--------- | --------- | -------- | ------- | --------- | ----------
HEIGHT | HEIGHT | Height (cm) | 1 | Subject Characteristic | 1
WEIGHT | WEIGHT | Weight (kg) | 2 | Subject Characteristic | 1
DIABP | DIABP | Diastolic Blood Pressure (mmHg) | 3 | Vital Sign | 2
MAP | MAP | Mean Arterial Pressure | 4 | Vital Sign | 2
PULSE | PULSE | Pulse Rate (beats/min) | 5 | Vital Sign | 2
SYSBP | SYSBP | Systolic Blood Pressure (mmHg) | 6 | Vital Sign | 2
TEMP | TEMP | Temperature (C) | 7 | Vital Sign | 2

This lookup may now be joined to the source data:

```{r eval=TRUE, include=FALSE}
param_lookup <- tribble(
  ~VSTESTCD, ~PARAMCD, ~PARAM, ~PARAMN, ~PARCAT1, ~PARCAT1N,
  "HEIGHT", "HEIGHT", "Height (cm)", 1, "Subject Characteristic", 1,
  "WEIGHT", "WEIGHT", "Weight (kg)", 2, "Subject Characteristic", 1,
  "DIABP", "DIABP", "Diastolic Blood Pressure (mmHg)", 3, "Vital Sign", 2,
  "MAP", "MAP", "Mean Arterial Pressure (mmHg)", 4, "Vital Sign", 2,
  "BSA", "BSA", "Body Surface Area (m^2)", 5, "Vital Sign", 2,
  "PULSE", "PULSE", "Pulse Rate (beats/min)", 6, "Vital Sign", 2,
  "SYSBP", "SYSBP", "Systolic Blood Pressure (mmHg)", 7, "Vital Sign",2,
  "TEMP", "TEMP", "Temperature (C)", 8, "Vital Sign", 2
)
attr(param_lookup$VSTESTCD, "label") <- "Vital Signs Test Short Name"

head(param_lookup)
```

```{r eval=TRUE}
advs <-  left_join(advs, param_lookup, by = "VSTESTCD")

count(advs, VSTESTCD, PARAMCD, PARAM, PARAMN, PARCAT1, PARCAT1N)

head(select(advs, USUBJID, VSTESTCD, PARAMCD, PARAM, PARAMN, PARCAT1, PARCAT1N))
```

Please note, it may be necessary to include other variables in the join. For
example, perhaps the `PARAMCD` is based on `VSTESTCD` and `VSPOS`, it may be 
necessary to expand this lookup or create a separate look up for `PARAMCD`.

### Derive Results (`AVAL`, `AVALC`, `AVALU`) {#aval}

The mapping of `AVAL`, `AVALC`, and `AVALU` is left to the ADaM programmer. An 
example mapping may be:

```{r eval=TRUE}
advs <- mutate(
  advs, 
  AVAL = VSSTRESN,
  AVALC = VSSTRESC,
  AVALU = VSSTRESU
)

head(select(advs, USUBJID, VSSTRESN, VSSTRESC, VSSTRESU, VSTPT, AVAL, AVALC, AVALU))
```
### Derive Additional Parameters (e.g. `BSA` or `MAP` for `ADVS`)] {#derive_param}

Optionally derive new parameters creating `PARAMCD`, `PARAM`, `AVAL` (and any 
other variables). Below is an example of creating `Mean Arterial Pressure` for `advs`, see also Example 3 in section below [Derive New Rows](#additional) for alternative way of creating new parameters.


```{r eval=TRUE}
advs <- derive_param_map(
  advs,
  by_vars = vars(USUBJID, VISIT, ADT, VSTPTNUM),
  set_values_to = vars(PARAMCD = "MAP", AVALU = "mmHg"),
  get_unit_expr = AVALU
)
 
head(select(filter(advs, PARAMCD == "MAP"), USUBJID, VISIT, VSTPTNUM, PARAMCD, AVAL, AVALU))
```

Likewise, function call below, to create parameter `Body Surface Area' for 
'advs' domain.

```{r eval=TRUE}
advs <- derive_param_bsa(
  advs,
  unit_var = NULL,
  by_vars = vars(USUBJID, VISIT, ADT),
  set_values_to = vars(PARAMCD = "BSA", AVALU = "m^2"),
  get_unit_expr = AVALU
  )
 
head(select(filter(advs, PARAMCD == "BSA"), USUBJID, VISIT, PARAMCD, AVAL, AVALU))
```

You can either create `PARAM`, `PARCAT1`, etc in function calls above in the 
`set_values_to` parameter or derive from look-up table described earlier.

``` {r eval=TRUE}
advs <- select(filter(advs, PARAMCD %in% c("MAP", "BSA")), 
               -PARAM, -PARAMN, -PARCAT1, -PARCAT1N) %>% 
  left_join(select(param_lookup, -VSTESTCD), by = "PARAMCD") %>%
  union_all(filter(advs, !(PARAMCD %in% c("MAP", "BSA"))))

advs %>%
  filter(PARAMCD %in% c("BSA", "MAP") & USUBJID == "01-701-1015" & 
           VISIT == "SCREENING 1") %>%
  select(PARAMCD, PARAM, PARAMN, PARCAT1, PARCAT1N) %>%
  head()

```

Similarly, for 'adeg' domain, the parameters 'QTCBF' 'QTCBS' and 'QTCL' can be
created with a function call. See example below for `PARAMCD` = `QTCF'.

```{r eval=FALSE}
adeg <- derive_param_qtcf(
  adeg,
  unit_var = NULL,
  by_vars = vars(USUBJID, VISIT, ADT),
  set_values_to = vars(PARAMCD = "QTCFR", AVALU = 'msec')
  )
 
```

```{r eval=TRUE, echo=FALSE}
advs <- filter(advs, !(PARAMCD %in% c("MAP", "BSA")))
```

### Derive Timing Variables (e.g. `APHASE`, `AVISIT`, `APERIOD`) {#timing}

Categorical timing variables are protocol and analysis dependent.  Below is a 
simple example to calculate `APHASE`. 

```{r eval=TRUE}
advs <- advs %>% 
  mutate(
    APHASE = case_when(
      is.na(ADT) ~ NA_character_,
      ADT < RFSTDTC ~ "Screening",
      RFSTDTC <= ADT & ADT <= RFENDTC ~ "Treatment",
      RFENDTC < ADT ~ "Follow-up",
      TRUE ~ "UNASSIGNED"
    ),
    AVISIT = case_when(
      str_detect(VISIT, "SCREEN") ~ NA_character_,
      str_detect(VISIT, "UNSCHED") ~ NA_character_,
      str_detect(VISIT, "RETRIEVAL") ~ NA_character_,
      str_detect(VISIT, "AMBUL") ~ NA_character_,
      !is.na(VISIT) ~ str_to_title(VISIT)
    ),
    AVISITN = as.numeric(case_when(
      VISIT == "BASELINE" ~ "0",
      str_detect(VISIT, "WEEK") ~ str_trim(str_replace(VISIT, "WEEK", ""))
    )),
    ATPT = VSTPT,
    ATPTN = VSTPTNUM
  )

head(select(advs, USUBJID, ADT, RFSTDTC, RFENDTC, APHASE))

count(advs, VISITNUM, VISIT, AVISITN, AVISIT)

count(advs, VSTPTNUM, VSTPT, ATPTN, ATPT)
```

### Timing Flag Variables (e.g. `ONTRTFL`) {#timingflag}

In some analyses, it may be necessary to flag an observation as on-treatment.
The admiral function `derive_var_ontrtfl()` can be used.

For example, if on-treatment is defined as any observation between treatment
start and treatment end, the flag may be derived as:

```{r eval=TRUE}
advs <- derive_var_ontrtfl(
  advs, 
  start_date = ADT, 
  ref_start_date = TRTSDT, 
  ref_end_date = TRTEDT
)

head(select(advs, USUBJID, PARAMCD, ADT, TRTSDT, TRTEDT, ONTRTFL))
```

This function returns the original data frame with the column `ONTRTFL` added.
Additionally, this function does have functionality to handle a window on the
`ref_end_date`.  For example, if on-treatment is defined as between treatment 
start and treatment end plus 60 days, the call would be:

```{r include=FALSE}
advs <- select(advs, -ONTRTFL)
```

```{r eval=TRUE}
advs <- derive_var_ontrtfl(
  advs, 
  start_date = ADT, 
  ref_start_date = TRTSDT, 
  ref_end_date = TRTEDT,
  ref_end_window = 60
)
```

Lastly, the function does allow you to filter out pre-treatment observations
that occurred on the start date.  For example, if observations with `VSTPT == PRE`
should not be considered on-treatment when the observation date falls between 
the treatment start and end date, the user may specify this using the 
`filter_pre_timepoint` parameter:

```{r include=FALSE}
advs <- select(advs, -ONTRTFL)
```

```{r eval=TRUE}
advs <- derive_var_ontrtfl(
  advs,
  start_date = ADT,
  ref_start_date = TRTSDT,
  ref_end_date = TRTEDT,
  filter_pre_timepoint = ATPT == "AFTER LYING DOWN FOR 5 MINUTES"
)
```

### Assign Reference Range Indicator (`ANRIND`) {#referencerange}

The admiral function `derive_var_anrind()` may be used to derive the reference
range indicator `ANRIND`.

This function requires the reference range boundaries to exist on the data frame 
(`ANRLO`, `ANRHI`) and also accommodates the additional boundaries `A1LO` and `A1HI`.

```{r include=FALSE}
range_lookup <- tibble::tribble(
  ~PARAMCD, ~ANRLO, ~ANRHI, ~A1LO, ~A1HI,
  "SYSBP", 90, 130, 70, 140,
  "DIABP", 60, 80, 40, 90,
  "PULSE", 60, 100, 40, 110,
  "TEMP", 36.5, 37.5, 35, 38
)

advs <- left_join(advs, range_lookup, by = "PARAMCD")
```

The function is called as:

```{r eval=TRUE}
advs <- derive_var_anrind(advs)

head(select(advs, USUBJID, PARAMCD, AVAL, ANRLO, ANRHI, A1LO, A1HI, ANRIND))
```

### Derive Baseline (`BASETYPE`, `ABLFL`, `BASE`, `BASEC`, `BNRIND`) {#baseline}

The `BASETYPE` should be derived using the function `derive_var_basetype()`.
The parameter basetypes of this function requires a named list of expression 
detailing how the BASETYPE should be assigned.  Note, if a record falls into
multiple expressions within the basetypes expression, a row will be produced for 
each `BASETYPE`.

```{r eval=TRUE}
advs <- derive_var_basetype(
  dataset = advs,
  basetypes = exprs(
    "AFTER LYING DOWN FOR 5 MINUTES" = ATPTN == 815,
    "AFTER STANDING FOR 1 MINUTE" = ATPTN == 816,
    "AFTER STANDING FOR 3 MINUTES" = ATPTN == 817
  )
)

count(advs, ATPT, ATPTN, BASETYPE)
```

It is important to derive `BASETYPE` first so that it can be utilized in 
subsequent derivations. This will be important if the data frame contains
multiple values for `BASETYPE`.

Next, the analysis baseline flag (`ABLFL`) can be derived using the admiral
function `derive_extreme_flag()`.  For example, if baseline is defined as the last 
non-missing `AVAL` prior or on `TRTSDT`, the function call for `ABLFL` would be:

```{r eval=TRUE}
advs <- derive_extreme_flag(
  advs,
  by_vars = vars(STUDYID, USUBJID, BASETYPE, PARAMCD),
  order = vars(ADT, ATPTN, VISITNUM),
  new_var = ABLFL,
  mode = "last",
  filter = (!is.na(AVAL) & ADT <= TRTSDT & !is.na(BASETYPE))
)

head(select(advs, USUBJID, BASETYPE, PARAMCD, ADT, TRTSDT, ATPTN, TRTSDT, ABLFL))
```

Note: Additional examples of the `derive_extreme_flag()` function can be 
found [above.](#analysisrec)

Lastly, the `BASE` and `BASEC` columns can be derived using the admiral function
`derive_var_base()` and `derive_var_basec()`.  In addition, `BNRIND` can be derived
using the low-level function `derive_baseline()`.  Example calls are:

```{r eval=TRUE}
advs <- derive_var_base(
  advs,
  by_vars = vars(STUDYID, USUBJID, PARAMCD, BASETYPE)
)

advs <- derive_var_basec(
  advs,
  by_vars = vars(STUDYID, USUBJID, PARAMCD, BASETYPE)
)

advs <- derive_baseline(
  advs,
  by_vars = vars(STUDYID, USUBJID, PARAMCD, BASETYPE),
  source_var = ANRIND, 
  new_var = BNRIND
)

head(select(advs, USUBJID, BASETYPE, PARAMCD, ABLFL, BASE, BASEC, ANRIND, BNRIND))
```

### Derive Change from Baseline (`CHG`, `PCHG`) {#bchange}

Change and percent change from baseline can be derived using the admiral 
functions `derive_var_chg()` and `derive_var_pchg()`.  These functions expect `AVAL` 
and `BASE` to exist in the data frame.  The `CHG` is simply `AVAl - BASE` and the
`PCHG` is `(AVAL - BASE) / absolute value (BASE) * 100`.  Examples calls are:

```{r eval=TRUE}
advs <- derive_var_chg(advs)

advs <- derive_var_pchg(advs)

head(select(advs, USUBJID, BASE, AVAL, CHG, PCHG))
```

### Derive Analysis Flags (e.g. `ANL01FL`) {#analysisrec}

In most finding ADaMs, an analysis flag is derived to identify the appropriate 
observation(s) to use for a particular analysis when a subject has multiple
observations within a particular timing period.

In this situation, an analysis flag (e.g. `ANLxxFL`) may be used to choose the
appropriate record for analysis.

This flag may be derived using the admiral function `derive_extreme_flag()`. 
For this example, we will assume we would like to choose the latest and 
highest value by `USUBJID`, `PARAMCD`, `AVISIT`, and `ATPT`.

```{r eval=TRUE}
advs <- derive_extreme_flag(
  advs,
  by_vars = vars(USUBJID, PARAMCD, AVISIT),
  order = vars(ADT, ATPTN, AVAL),
  new_var = ANL01FL,
  mode = "last",
  filter = !is.na(AVISITN)
)

head(select(advs, USUBJID, PARAMCD, AVISIT, ATPTN, ADT, AVAL, ANL01FL))
```

Another common example would be flagging the worst value for a subject,
parameter, and visit. For this example, we will assume we have 3 `PARAMCD` 
values (SYSBP, DIABP, and RESP).  We will also assume high is worst for SYSBP 
and DIABP and low is worst for RESP.

```{r eval=TRUE}
advs <- derive_extreme_flag(
  advs,
  by_vars = vars(USUBJID, PARAMCD, AVISIT),
  order = vars(AVAL, ADT, ATPTN),
  new_var = WORST_HIGH,
  mode = "last",
  filter = !is.na(AVISIT) & !is.na(AVAL)
) 

advs <- derive_extreme_flag(
  advs,
  by_vars = vars(USUBJID, PARAMCD, AVISIT),
  order = vars(AVAL, ADT, ATPTN),
  new_var = WORST_LOW,
  mode = "first",
  filter = !is.na(AVISIT) & !is.na(AVAL)
)

advs <- mutate(
  advs, 
  WORSTFL = case_when(
    PARAMCD %in% c("SYSBP", "DIABP") ~ WORST_HIGH,
    PARAMCD == "RESP" ~ WORST_LOW
  )
)

head(select(advs, USUBJID, PARAMCD, AVISIT, AVAL, ADT, ATPTN, WORSTFL))
```

### Assign Treatment (`TRTA`, `TRTP`) {#treatment}

`TRTA` and `TRTP` must correlate to treatment `TRTxxP` and/or `TRTxxA` in ADSL. The 
derivation of `TRTA` and `TRTP` for a record are protocol and analysis specific.  
admiral does not currently have functionality to assist with `TRTA` and `TRTP` 
assignment.

However, an example of a simple implementation could be:

```{r eval=TRUE}
advs <- mutate(advs, TRTP = TRT01P, TRTA = TRT01A)

count(advs, TRTP, TRTA, TRT01P, TRT01A)

head(select(advs, USUBJID, TRT01P, TRT01A, TRTP, TRTA))
```

### Assign `ASEQ` {#aseq}

The admiral function `derive_obs_number()` can be used to derive `ASEQ`. An 
example call is:

```{r eval=TRUE}
advs <- derive_obs_number(
  advs,
  new_var = ASEQ,
  by_vars = vars(STUDYID, USUBJID),
  order = vars(PARAMCD, ADT, AVISITN, ATPTN, VISIT, ADT),
  check_type = "warning"
)

head(select(advs, USUBJID, PARAMCD, ADT, AVISITN, ATPTN, VISIT, ADT, ASEQ))
```

### Derive New Rows {#additional}

When deriving new rows for a data frame, it is essential the programmer takes 
time to insert this derivation in the correct location of the code. The location
will vary depending on what previous computations should be retained on the new 
record and what computations must be done with the new records.

Sample Code | Description 
----------- | ------------------------------------------------------------------
ad_advs.R | Create End of Treatment based on a subset of rows, for each USUBJID, PARAMCD
ad_advs.R | Derive an Average Record (e.g. Triplicates) for each USUBJID, PARAMCD, and timepoint.
ad_advs.R | Derive a new PARAMCD based on existing PARAMCDs (MAP, based on Systolic and Diastolic BP collected at a particular timepoint.)

Example 1 (Creating a new record):

```{r eval=TRUE}
advs <- advs %>%
  derive_extreme_flag(
    by_vars = vars(STUDYID, USUBJID, PARAMCD),
    order = vars(ADT, AVISITN, ATPTN, AVAL),
    new_var = EOTFL,
    mode = "last",
    filter = (4 < AVISITN & AVISITN <= 12 & ANL01FL == "Y")
  ) %>%
  filter(EOTFL == "Y") %>%
  mutate(
    AVISIT = "End of Treatment",
    AVISITN = 99
  ) %>%
  union_all(advs) %>%
  select(-EOTFL)

advs %>% 
  select(USUBJID, PARAMCD, ADT, AVISITN, AVISIT, ATPTN, AVAL, ANL01FL) %>% 
  filter(ANL01FL == "Y") %>% 
  arrange(USUBJID, PARAMCD, ADT, AVISITN, ATPTN) %>% 
  head(n = 10)
```

Example 2 (Deriving a Summary record):

```{r eval=TRUE}
advs <- derive_summary_records(
  advs,
  by_vars = vars(STUDYID, USUBJID, PARAMCD, VISITNUM, ADT),
  analysis_var = AVAL,
  summary_fun = mean,
  set_values_to = vars(DTYPE = "AVERAGE")
)

head(select(advs, USUBJID, PARAMCD, VISITNUM, ADT, AVAL, DTYPE))
```

Example 3 (Derive a new PARAMCD):

```{r eval=TRUE}
sysbp <- advs %>%
  filter(VSTESTCD == "SYSBP") %>%
  select(
    -VSSEQ, -VSTESTCD, -VSTEST, -VSORRES, -VSORRESU, -VSSTRESC, -VSSTRESN,
    -VSSTRESU, -VSLOC, -VSBLFL, -VSSTAT, -AVALC
  )

diabp <- advs %>%
  filter(VSTESTCD == "DIABP") %>%
  select(STUDYID, USUBJID, VISITNUM, VSDTC, VSTPT, AVAL) %>%
  rename(DBPAVAL = AVAL)

advs <- sysbp %>%
  left_join(
    diabp,
    by = c("STUDYID", "USUBJID", "VISITNUM", "VSDTC", "VSTPT")
  ) %>%
  mutate(
    AVAL = (2 * DBPAVAL + AVAL) / 3,
    PARAMCD = "MAP"
  ) %>%
  filter(!is.na(AVAL)) %>%
  left_join(select(param_lookup, -VSTESTCD), by = "PARAMCD") %>%
  union_all(advs) %>%
  select(-DBPAVAL)

advs %>% 
  filter(PARAMCD %in% c("SYSBP", "DIABP", "MAP")) %>% 
  arrange(USUBJID, PARAMCD, VISITNUM, VSDTC, VSTPT, AVAL) %>% 
  head()
```

### Derive Categorization Variables (`AVALCATx`) {#cat}

Admiral does not currently have a generic function to aid in assigning `AVALCATx`/
`AVALCAxN` values. Below is a simple example of how these values may be
assigned:

```{r eval=TRUE}
avalcat_lookup <- tibble::tribble(
  ~PARAMCD, ~AVALCA1N, ~AVALCAT1,
  "HEIGHT", 1, ">100 cm",
  "HEIGHT", 2, "<= 100 cm"
)

advs <- advs %>% 
  mutate(
    AVALCA1N = case_when(
      PARAMCD == "HEIGHT" & AVAL > 100 ~ 1,
      PARAMCD == "HEIGHT" & AVAL <= 100 ~ 2
    )
  ) %>%
  left_join(avalcat_lookup, by = c("PARAMCD", "AVALCA1N"))

advs %>% 
  select(USUBJID, PARAMCD, AVAL, AVALCA1N, AVALCAT1) %>% 
  filter(PARAMCD == "HEIGHT") %>% 
  head()
```

### Example Scripts {#example}

ADaM | Sample Code 
---- | --------------
ADVS | [ad_advs.R](https://github.com/Roche-GSK/admiral/blob/master/inst/example_scripts/ad_advs.R){target="_blank"}
