---
title: "Creating a BDS Time-to-Event ADaM"
output:
  rmarkdown::html_vignette:
    toc: false

vignette: >
  %\VignetteIndexEntry{Creating a BDS Time-to-Event ADaM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

This article describes creating a BDS time-to-event ADaM.

The main part in programming a time-to-event dataset is the definition of the
events and censoring times. admiral supports single events like death or
composite events like disease progression or death. More than one source dataset
can be used for the definition of the event and censoring times.

**Note**: *All examples assume CDISC SDTM and/or ADaM format as input unless
otherwise specified.*

## Required Packages

The examples of this vignette require the following packages.

```{r, warning=FALSE, message=FALSE}
library(admiral)
library(dplyr)
```

```{r, warning=FALSE, message=FALSE, include=FALSE}
library(lubridate)
```

# Programming Workflow

* [Read in Data](#readdata)
* [Derive Parameters (`CNSR`, `ADT`, `STARTDT`)](#parameters)
* [Derive Analysis Value (`AVAL`)](#aval)
* [Derive Analysis Sequence Number (`ASEQ`)](#aseq)
* [Add ADSL Variables](#adslvars)

## Read in Data {#readdata}

To start, all datasets needed for the creation of the time-to-event dataset
should be read into the environment. This will be a company specific process.

For example purpose, the CDISC Pilot SDTM and ADaM datasets---which are included
in `{admiral}`---are used.

```{r}
data("adsl")
```

## Derive Parameters (`CNSR`, `ADT`, `STARTDT`) {#parameters}

The events and censorings are defined by the `event_source()` and the
`censor_source()` class respectively. It defines

- which observations (`filter` parameter) of a source dataset (`dataset_name`
parameter) are potential events or censorings,
- the value of the `CNSR` variable (`censor` parameter), and
- which variable provides the date (`date` parameter).

The date can be provided as date (`*DT` variable), datetime (`*DTM` variable),
or character ISO-8601 date (`*DTC` variable).

The `dataset_name` parameter expects a character value which is used as an
identifier. The actual data which is used for the derivation of the parameter is
provided via the `source_datasets` parameter of `derive_param_tte()`. It expects
a named list of datasets. The names correspond to the identifiers specified for
the `dataset_name` parameter. This allows to define events and censoring
independent of the data.

### First Example

For example the following call defines a death event based on ADSL variables.
The `censor` parameter is not available for events because `CNSR = 0` is used
for events following the CDISC recommendation.

```{r}
death <- event_source(
  dataset_name = "adsl",
  filter = DTHFL == "Y",
  date = DTHDT)
```

A corresponding censoring based on the last known alive date can be defined by
the following call. For censorings the value of the `CNSR` variable must be a
positive integer. It is defaulted to `1`.
```{r}
lstalv <- censor_source(
  dataset_name = "adsl",
  date = LSTALVDT)
```

The definitions can be passed to `derive_param_tte()` to create a new time-to-event parameter.
```{r}
adtte <- derive_param_tte(
  dataset_adsl = adsl,
  source_datasets = list(adsl = adsl),
  event_conditions = list(death),
  censor_conditions = list(lstalv),
  set_values_to = vars(PARAMCD = "OS",
                       PARAM = "Overall Survival"))
```
```{r, echo=FALSE}
dataset_vignette(
  adtte,
  display_vars = vars(USUBJID, PARAMCD, PARAM, STARTDT, ADT, CNSR))
```

### Add Additional Information for Events and Censoring (`EVNTDESC`, `SRCVAR`, ...)
To add additional information like event or censoring description (`EVNTDESC`)
or source variable (`SRCVAR`) the `set_values_to` parameter can be specified in
the event/censoring definition.
```{r}
# define death event #
death <- event_source(
  dataset_name = "adsl",
  filter = DTHFL == "Y",
  date = DTHDT,
  set_values_to =vars(
    EVNTDESC = "DEATH",
    SRCDOM = "ADSL",
    SRCVAR = "DTHDT"))

# define censoring at last known alive date #
lstalv <- censor_source(
  dataset_name = "adsl",
  date = LSTALVDT,
  set_values_to = vars(
    EVNTDESC = "LAST KNOWN ALIVE DATE",
    SRCDOM = "ADSL",
    SRCVAR = "LSTALVDT"))

# derive time-to-event parameter #
adtte <- derive_param_tte(
  dataset_adsl = adsl,
  source_datasets = list(adsl = adsl),
  event_conditions = list(death),
  censor_conditions = list(lstalv),
  set_values_to = vars(PARAMCD = "OS",
                       PARAM = "Overall Survival"))
```
```{r, echo=FALSE}
dataset_vignette(
  adtte,
  display_vars = vars(USUBJID, EVNTDESC, SRCDOM, SRCVAR, CNSR, ADT))
# save adtte and adsl for next section
adtte_bak <- adtte
adsl_bak <- adsl
```

### Handling Subjects Without Assessments

If a subject has no valid date in any of the event or censoring source datasets,
it will not be included in the output dataset. Therefore the `start` censoring
is defined below to achieve that subjects without data in `adrs` are censored at
the start date.

The ADaM IG requires that computed date must be accompanied by imputation flags.
If the start date is imputed, the date imputation flag can be specified by the
`start_date_imputation_flag` parameter. If a date variable from one of the event
or censoring source datasets is imputed, the imputation flag can be specified
for the `set_values_to` parameter in `event_source()` or `censor_source()` (see
definition of the `start` censoring below).
```{r, eval=FALSE}
View(adsl)
```
```{r, echo=FALSE}
adsl <- tibble::tribble(
    ~USUBJID, ~DTHFL, ~DTHDT,            ~TRTSDT,           ~TRTSDTF,
    "01",     "Y",    ymd("2021-06-12"), ymd("2021-01-01"), "M",
    "02",     "N",    NA,                ymd("2021-02-03"), NA,
    "03",     "Y",    ymd("2021-08-21"), ymd("2021-08-10"), NA,
    "04",     "N",    NA,                ymd("2021-02-03"), NA,
    "05",     "N",    NA,                ymd("2021-04-01"), "D") %>%
    mutate(STUDYID = "AB42")
dataset_vignette(
  adsl,
  display_vars = vars(USUBJID, DTHFL, DTHDT, TRTSDT, TRTSDTF))
```
```{r, eval=FALSE}
View(adrs)
```
```{r, echo=FALSE}
  adrs <- tibble::tribble(
    ~USUBJID, ~AVALC, ~ADT,              ~ASEQ,
    "01",     "SD",   ymd("2021-01-03"), 1,
    "01",     "PR",   ymd("2021-03-04"), 2,
    "01",     "PD",   ymd("2021-05-05"), 3,
    "02",     "PD",   ymd("2021-02-03"), 1,
    "04",     "SD",   ymd("2021-02-13"), 1,
    "04",     "PR",   ymd("2021-04-14"), 2,
    "04",     "CR",   ymd("2021-05-15"), 3) %>%
    mutate(STUDYID = "AB42",
           PARAMCD = "OVR",
           PARAM = "Overall Response") %>% 
    select(STUDYID, USUBJID, PARAMCD, PARAM, ADT, ASEQ, AVALC)

dataset_vignette(
  adrs,
  display_vars = vars(USUBJID, AVALC, ADT, ASEQ, PARAMCD, PARAM))
```

```{r}
# progressive disease event #
pd <- event_source(
  dataset_name = "adrs",
  filter = AVALC == "PD",
  date = ADT,
  set_values_to = vars(
    EVNTDESC = "PD",
    SRCDOM = "ADRS",
    SRCVAR = "ADT",
    SRCSEQ = ASEQ
  )
)

# death event #
death <- event_source(
  dataset_name = "adsl",
  filter = DTHFL == "Y",
  date = DTHDT,
  set_values_to = vars(
    EVNTDESC = "DEATH",
    SRCDOM = "ADSL",
    SRCVAR = "DTHDT"
  )
)

# last tumor assessment censoring (CNSR = 1 by default) #
lastvisit <- censor_source(
  dataset_name = "adrs",
  date = ADT,
  set_values_to = vars(
    EVNTDESC = "LAST TUMOR ASSESSMENT",
    SRCDOM = "ADRS",
    SRCVAR = "ADT"
  )
)

# start date censoring (for patients without tumor assessment) (CNSR = 2) #
start <- censor_source(
  dataset_name = "adsl",
  date = TRTSDT,
  censor = 2,
  set_values_to = vars(
    EVNTDESC = "TREATMENT START",
    SRCDOM = "ADSL",
    SRCVAR = "TRTSDT",
    ADTF = TRTSDTF
  )
)

# derive time-to-event parameter #
adtte <- derive_param_tte(
  dataset_adsl = adsl,
  source_datasets = list(adsl = adsl, adrs = adrs),
  start_date = TRTSDT,
  start_date_imputation_flag = TRTSDTF,
  event_conditions = list(pd, death),
  censor_conditions = list(lastvisit, start),
  set_values_to = vars(PARAMCD = "PFS",
                       PARAM = "Progression Free Survival")
)
```
```{r, echo=FALSE}
dataset_vignette(
  adtte %>%
  select(STUDYID, USUBJID, PARAMCD, PARAM, STARTDT, STARTDTF, ADT, ADTF, CNSR,
    EVNTDESC, SRCDOM, SRCVAR
  ),
  display_vars = vars(USUBJID, PARAMCD, STARTDT, STARTDTF, ADT, ADTF, CNSR))
```

### Deriving a Series of Time-to-Event Parameters
If several similar time-to-event parameters need to be derived the
`call_derivation()` function is useful.

In the following example parameters for time to first AE, time to first serious
AE, and time to first related AE are derived. The censoring is the same for all
three. Only the definition of the event differs.
```{r echo=FALSE}
adtte <- adtte_bak
adsl <- adsl_bak
```

```{r}
# add end of AE observation period to ADSL #
adsl <- adsl %>% mutate(AEOBSEDT = pmin(TRTEDT + 30, EOSDT))
```
```{r, echo=FALSE}
dataset_vignette(
  adsl,
  display_vars = vars(USUBJID, TRTEDT, EOSDT, AEOBSEDT))
```

```{r}
# define censoring #
observation_end <- censor_source(
  dataset_name = "adsl",
  date = AEOBSEDT,
  censor = 1,
  set_values_to = vars(
    EVNTDESC = "OBSERVATION END",
    SRCDOM = "ADSL",
    SRCVAR = "AEOBSEDT"
  )
)

# define time to first AE #
tt_ae <- event_source(
  dataset_name = "ae",
  date = AESTDTC,
  set_values_to = vars(
    EVNTDESC = "ADVERSE EVENT",
    SRCDOM = "AE",
    SRCVAR = "AESTDTC"
  )
)

# define time to first serious AE #
tt_ser_ae <- event_source(
  dataset_name = "ae",
  filter = AESER == "Y",
  date = AESTDTC,
  set_values_to = vars(
    EVNTDESC = "SERIOUS ADVERSE EVENT",
    SRCDOM = "AE",
    SRCVAR = "AESTDTC"
  )
)

# define time to first related AE #
tt_rel_ae <- event_source(
  dataset_name = "ae",
  filter = AEREL %in% c("PROBABLE", "POSSIBLE", "REMOTE"),
  date = AESTDTC,
  set_values_to = vars(
    EVNTDESC = "RELATED ADVERSE EVENT",
    SRCDOM = "AE",
    SRCVAR = "AESTDTC"
  )
)

# derive all three time to event parameters #
adaette <- call_derivation(
  derivation = derive_param_tte,
  variable_params = list(
    params(event_conditions = list(tt_ae),
           set_values_to = vars(PARAMCD = "TTAE")),
    params(event_conditions = list(tt_ser_ae),
           set_values_to = vars(PARAMCD = "TTSERAE")),
    params(event_conditions = list(tt_rel_ae),
           set_values_to = vars(PARAMCD = "TTRELAE"))),
  dataset_adsl = adsl,
  source_datasets = list(adsl = adsl, ae = ae),
  censor_conditions = list(observation_end)
  )
```
```{r, echo=FALSE}
dataset_vignette(
  adaette %>% select(STUDYID, USUBJID, PARAMCD, STARTDT, ADT, CNSR, EVNTDESC, SRCDOM, SRCVAR)
  %>% arrange(USUBJID, PARAMCD),
  display_vars = vars(USUBJID, PARAMCD, STARTDT, ADT, CNSR, EVNTDESC, SRCDOM, SRCVAR))
```

### Deriving Time-to-Event Parameters Using By Groups

If time-to-event parameters need to be derived for each by group of a source
dataset, the `by_vars` parameter can be specified. Then a time-to-event
parameter is derived for each by group.

Please note that CDISC requires separate parameters (`PARAMCD`, `PARAM`) for the
by groups. Therefore the variables specified for the `by_vars` parameter are not
included in the output dataset. The `PARAMCD` variable should be specified for
the `set_value_to` parameter using an expression on the right hand side which
results in a unique value for each by group. If the values of the by variables
should be included in the output dataset, they can be stored in `PARCATn`
variables.

In the following example a time-to-event parameter for each preferred term in
the AE dataset is derived.
```
View(adsl)
```
```{r, echo=FALSE}
adsl <- tibble::tribble(
  ~USUBJID, ~TRTSDT,           ~EOSDT,
  "01",     ymd("2020-12-06"), ymd("2021-03-06"),
  "02",     ymd("2021-01-16"), ymd("2021-02-03")) %>%
  mutate(STUDYID = "AB42")

dataset_vignette(adsl)
```
```
View(ae)
```
```{r, echo=FALSE}
ae <- tibble::tribble(
  ~USUBJID, ~AESTDTC,           ~AESEQ, ~AEDECOD,
  "01",     "2021-01-03T10:56", 1,      "Flu",
  "01",     "2021-03-04",       2,      "Cough",
  "01",     "2021",             3,      "Flu") %>%
  mutate(STUDYID = "AB42")

dataset_vignette(ae)
```
```{r}
# define time to first adverse event event #
ttae <- event_source(
  dataset_name = "ae",
  date = AESTDTC,
  set_values_to = vars(
    EVNTDESC = "AE",
    SRCDOM = "AE",
    SRCVAR = "AESTDTC",
    SRCSEQ = AESEQ))

# define censoring at end of study #
eos <- censor_source(
  dataset_name = "adsl",
  date = EOSDT,
  set_values_to = vars(
    EVNTDESC = "END OF STUDY",
    SRCDOM = "ADSL",
    SRCVAR = "EOSDT"))

# derive time-to-event parameter #
adtte <- derive_param_tte(
  dataset_adsl = adsl,
  by_vars = vars(AEDECOD),
  start_date = TRTSDT,
  event_conditions = list(ttae),
  censor_conditions = list(eos),
  source_datasets = list(adsl = adsl, ae = ae),
  set_values_to = vars(
    PARAMCD = paste0("TTAE", as.numeric(as.factor(AEDECOD))),
    PARAM = paste("Time to First", AEDECOD, "Adverse Event"),
    PARCAT1 = "TTAE",
    PARCAT2 = AEDECOD
  )
)
```
```{r, echo=FALSE}
dataset_vignette(
  adtte %>%
    select(
      USUBJID, STARTDT, PARAMCD, PARAM, PARCAT1, PARCAT2, ADT, CNSR,
      EVNTDESC, SRCDOM, SRCVAR, SRCSEQ
    ),
  display_vars = vars(USUBJID, STARTDT, PARAMCD, PARAM, ADT, CNSR, SRCSEQ)
)
```

## Derive Analysis Value (`AVAL`) {#aval}

The analysis value (`AVAL`) can be derived by calling `derive_vars_duration()`.

This example derives the time to event in days. Other units can be requested by
the specifying the `out_unit` parameter.
```{r echo=FALSE}
adtte <- adtte_bak
adsl <- adsl_bak
```
```{r eval=TRUE}
adtte <- derive_vars_duration(
  adtte,
  new_var = AVAL,
  start_date = STARTDT,
  end_date = ADT
)
```
```{r, echo=FALSE}
dataset_vignette(
  adtte)
```

## Derive Analysis Sequence Number (`ASEQ`) {#aseq}

The admiral function `derive_obs_number()` can be used to derive `ASEQ`:

```{r eval=TRUE}
adtte <- derive_obs_number(
  adtte,
  by_vars = vars(STUDYID, USUBJID),
  order = vars(PARAMCD),
  check_type = "error"
)
```
```{r, echo=FALSE}
dataset_vignette(
  adtte)
```

## Add ADSL Variables {#adslvars}

Variables from ADSL which are required for time-to-event analyses, e.g.,
treatment variables or covariates can be added using `left_join()`.

```{r eval=TRUE}
adtte <- left_join(adtte,
                   adsl %>% select(STUDYID, USUBJID, ARMCD, ARM, ACTARMCD, ACTARM, AGE, SEX),
                   by = c("STUDYID", "USUBJID"))
```
```{r, echo=FALSE}
dataset_vignette(
  adtte)
```
