---
title: "Admiral Programming Concepts and Conventions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Admiral Programming Concepts and Conventions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(admiraldev)
```

# Introduction

This vignette aims to discuss some of the common programming concepts and conventions that have been adopted within the
`{admiral}` packages. This includes how missing values are handled in `{admiral}`, as well as the way in which users are
expected to pass inputs to `{admiral}` functions, and how those inputs should look.

# Handling of Missing Values

When using the `{haven}` package to read SAS datasets into R, SAS-style character missing values, i.e. `""`, are *not* 
converted into proper R `NA` values. Rather they are kept as is. This is problematic for any downstream data processing 
as R handles `""` just as any other string. Thus, before any data manipulation is being performed SAS blanks should be 
converted to R `NA`s using `{admiral}`'s `convert_blanks_to_na()` function, e.g.

```r
dm <- haven::read_sas("dm.sas7bdat") %>% 
  convert_blanks_to_na()
```

Note that any logical operator being applied to an `NA` value *always* returns `NA` rather than `TRUE` or `FALSE`.

```{r}
visits <- c("Baseline", NA, "Screening", "Week 1 Day 7")
visits != "Baseline"
```

The only exception is `is.na()` which returns `TRUE` if the input is `NA`.

```{r}
is.na(visits)
```

Thus, to filter all visits which are not `"Baseline"` the following condition would need to be used.

```{r}
visits != "Baseline" | is.na(visits)
```

Also note that most aggregation functions, like `mean()` or `max()`, also return `NA` if any element of the input vector is missing.

```{r}
mean(c(1, NA, 2))
```

To avoid this behavior one has to explicitly set `na.rm = TRUE`.

```{r}
mean(c(1, NA, 2), na.rm = TRUE)
```

This is very important to keep in mind when using `{admiral}`'s aggregation functions such as `derive_summary_records()`.

For handling of `NA`s in sorting variables see [Sort Order](generic.html#sort_order).

# Argument Conventions

Within the `{admiral}` package, any arguments which expect variable names or expressions of variable names, symbols or 
expressions must be specified rather than strings.

- For arguments which expect a single variable name, the name can be specified without quotes and quotation, e.g.
`new_var = TEMPBL`

- For arguments which expect one or more variable names, a list of symbols is expected, e.g. `by_vars = exprs(PARAMCD, AVISIT)` 

- For arguments which expect a single expression, the expression needs to be passed "as is", e.g. `filter = PARAMCD == "TEMP"`

- For arguments which expect one or more expressions, a list of expressions is expected, e.g. `order = exprs(AVISIT, desc(AESEV))`

If you are new to expressions, read the [corresponding section](#exprs) to learn more.

# `exprs()` {#exprs}

## What is `exprs()`?

`exprs()` is a function from the `{rlang}` package. It is used to create a list of expressions. The expressions are not evaluated 
- rather, they are passed on to the derivation function which evaluates them in its own environment. This allows the derivation 
function to evaluate the expressions in the context of the input dataset. Specifically, the `exprs()` function allows users to pass
variable names of datasets to the function without wrapping them in quotation marks. 

The expressions/`exprs()` framework is powerful because users are able to intuitively "inject code" into `admiral` functions
(through the function parameters) using very similar syntax as if they were writing open code, with the exception possibly being an
outer `exprs()` wrapper.  For instance, in the `derive_vars_merged()` call below, the user is merging `adsl` with `ex` and is able
to filter `ex` prior to the merge using an expression passed to the `filter_add` parameter. Because `filter_add` accepts expressions, 
the user has full power to filter their dataset as they please. In the same vein, the user is able to create any new variables they 
wish after the merge using the `new_vars` argument, to which they pass a list of expressions containing standard R code.

``` {r, eval = FALSE}
derive_vars_merged(
  adsl,
  dataset_add = ex,
  filter_add = !is.na(EXENDTM),
  by_vars = exprs(STUDYID, USUBJID),
  new_vars = exprs(
    TRTEDTM = EXENDTM, 
    TRTETMF = EXENTMF, 
    COMPTRT = ifelse(!is.na(EXENDTM, "Y", "N"))
  ),
  order = exprs(EXENDTM),
  mode = "last"
)
```

In summary, although expressions/`exprs()` may seem slightly clunky and un-intuitive to begin with, they allow for such power
and flexibility that they are a key part of the `{admiral}` package. 

## Common pitfalls 

# See also

- [Programming Strategy](https://pharmaverse.github.io/admiraldev/articles/programming_strategy.html)
