---
title: "Generic Functions"
output: 
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Generic Functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(admiraldev)
```

# Introduction

This vignette explains some of the `{admiral}` generic functions, which we believe
are able to be re-used to sometimes save users having to create so many of their
own functions for study-specific analysis derivations. These re-usable functions give
the users great flexibility, but the challenge is knowing which to use when. So
this vignette is intended to help understand the differences between each, thus
improving the findability of the function needed for each task.

The focus will be on the following different sets of generic functions:

* [_merged_ functions](#merged)
* [_joined_ functions](#joined)
* [_extreme_ functions](#extreme)

In each section below, we will show examples of different functions that belong
to each of these sets, but we don't list out each and every function offered.
Users should refer to the Reference page and search for example "merged" to see
the full list of functions offered for this respective set, with more examples
shown within each function page.

## Required Packages

The examples in this vignette require the following packages.

For example purpose, the SDTM datasets from `{admiral.test}` are used.

```{r, warning=FALSE, message=FALSE}
library(admiral)
library(admiral.test)
library(dplyr, warn.conflicts = FALSE)
library(stringr)
library(tibble)

data("admiral_dm")
data("admiral_ds")
data("admiral_ex")
data("admiral_ae")
dm <- convert_blanks_to_na(admiral_dm)
ds <- convert_blanks_to_na(admiral_ds)
ex <- convert_blanks_to_na(admiral_ex)
ae <- convert_blanks_to_na(admiral_ae)
```
```{r echo=FALSE}
# Filter test patients and make more realistic and interesting for the examples
dm <- filter(dm, USUBJID %in% c("01-701-1111", "01-701-1047", "01-701-1057"))
ds <- filter(ds, USUBJID %in% c("01-701-1111", "01-701-1047", "01-701-1057")) %>%
  mutate(DSSTDTC = case_when(
    USUBJID == "01-701-1111" & DSDECOD == "RANDOMIZED" ~ "2012-08-01",
    TRUE ~ DSSTDTC
  ))
ex <- filter(ex, USUBJID %in% c("01-701-1111", "01-701-1047", "01-701-1057"))
ae <- filter(ae, USUBJID %in% c("01-701-1111", "01-701-1047")) %>%
  mutate(AESTDTC = case_when(
    USUBJID == "01-701-1111" & AESTDY == "-61" ~ "2012-09-14",
    TRUE ~ AESTDTC
  )) %>%
  mutate(AESTDY = case_when(
    USUBJID == "01-701-1111" & AESTDY == "-61" ~ 8,
    TRUE ~ AESTDY
  ))
```

# Difference between _merged_ and _joined_ functions

Firstly, we need to address the question all new users will have... why do we have
"merged" and "joined" functions and are they not the same thing??

Essentially in both sets of functions we are trying to take information from
an additional dataframe (passed into the `dataset_add` argument) and join it to
our original dataframe (passed into the `dataset` argument). The subtle difference
is that "merged" functions select the relevant information from the additional
dataframe and then adds it to the original dataframe; whilst "joined" functions
allow an extra filter condition after the join to further select the relevant
information.

So in `derive_vars_joined()` for example, the filter conditions can depend on variables
from both dataframes; whilst in `derive_vars_merged()` only variables from the
additional dataframe can be used. Examples will be shown in the below sections to
further explain the difference.

In principle though, we actually could achieve every "merged" function result using
the equivalent "joined" function. However, the "joined" functions require much more
resources (time and memory), hence why we include both options for users.

# _merged_ functions {#merged}

## Description

As explained above, this simpler set of functions for joins selects relevant
information from the additional dataframe and joins it to the original dataframe.

The additional dataframe can be filtered based on a user-defined condition passed
into the `filter_add` argument and/or (where relevant) by selecting the first or
last observation for each by group (`order` and `mode` arguments) and then joined with
the input dataframe by user-specified variables passed into `by_vars` argument.
Variables from the additional dataframe can be joined to the input dataframe as
they are or can be renamed (`new_vars` argument).

The dataframe returned from the function call contains all observations and variables
from the original dataframe (with or without filtering applied) and additionally
the variable(s) specified for `new_var`/`new_vars` from the additional dataframe.
This is achieved via an underlying [dplyr::left_join](https://dplyr.tidyverse.org/reference/mutate-joins.html).
For observations without a match in the additional dataframe the new variable(s)
are set to `NA`. Observations in the additional dataframe which have no match in
the original dataframe are ignored.

## Examples

A simple call to derive `ADSL` randomization date (`ADSL.RANDDT`) from an
intermediate additional dataframe would use `derive_vars_merged()` as follows.

```{r eval=TRUE}
# Use DM domain as basis to build ADSL
adsl_01 <- dm %>%
  select(-DOMAIN)

# Convert disposition character date to numeric date without imputation
ds_ext <- derive_vars_dt(
  dataset = ds,
  dtc = DSSTDTC,
  new_vars_prefix = "DSST"
)

# Join randomization date to ADSL
adsl_02 <- adsl_01 %>%
  derive_vars_merged(
    dataset_add = ds_ext,
    filter_add = DSDECOD == "RANDOMIZED",
    by_vars = vars(STUDYID, USUBJID),
    new_vars = vars(RANDDT = DSSTDT)
  )
```

This call would return the input dataframe with the variable `RANDDT` added.

Now, an example to add first treatment datetime (`ADSL.TRTSDT`), where we
need to make use of `derive_vars_merged()` with the `mode` and `order` arguments
to select the required observations from the additional dataframe. Note: the
`filter_add` argument here shows a possible method for checking only for valid doses.

```{r eval=TRUE}
# Convert exposure start date to numeric date without imputation
ex_ext <- derive_vars_dt(
  dataset = ex,
  dtc = EXSTDTC,
  new_vars_prefix = "EXST"
)

# Determine first exposure datetime and add to ADSL
adsl_03 <- adsl_02 %>%
  derive_vars_merged(
    dataset_add = ex_ext,
    filter_add = (EXDOSE > 0 | (EXDOSE == 0 & str_detect(EXTRT, "PLACEBO"))) & !is.na(EXSTDT),
    new_vars = vars(TRTSDT = EXSTDT),
    order = vars(EXSTDT, EXSEQ),
    mode = "first",
    by_vars = vars(STUDYID, USUBJID)
  )
```

This call would return the input dataframe with `TRTSDT` added.

Beyond `derive_vars_merged()`, other "merged" functions offer specific common
analysis needs. For example, below shows an example of using `derive_var_merged_exist_flag()`
for adding safety population flag (`ADSL.SAFFL`).

In this case we create a new flag variable (`new_var` argument) based on information we
take from the additional dataframe to tell us if a certain `condition` is met.
For all by groups (`by_vars` argument) where the check evaluates as `TRUE` at least
once the flag is set to `"Y"` (or whatever user passes in `true_value` argument).

Note: we have extra arguments available here for patients that do not meet the `condition`.
In the below example, we set patients in `EX` with no valid dose to `"N"` (`false_value`)
and the same for patients with no observations at all in `EX` (`missing_value`).

```{r eval=TRUE}
# Add safety population flag to ADSL
adsl_04 <- adsl_03 %>%
  derive_var_merged_exist_flag(
    dataset_add = ex,
    by_vars = vars(STUDYID, USUBJID),
    new_var = SAFFL,
    condition = (EXDOSE > 0 | (EXDOSE == 0 & str_detect(EXTRT, "PLACEBO"))),
    false_value = "N",
    missing_value = "N"
  )
```

Here's how all these above derived variables then look in the dataset.

```{r, eval=TRUE, echo=FALSE}
dataset_vignette(
  dataset = adsl_04,
  display_vars = vars(USUBJID, RANDDT, TRTSDT, SAFFL)
)
```

Some further examples of "merged" functions are `derive_vars_merged_lookup()` to
join a user-defined lookup table as commonly used in BDS ADaMs,
`derive_var_merged_cat()` to create a categorization variable from the
information joined from the additional dataframe, or
`derive_var_merged_summary()` to merge summarized values from the additional
dataframe.

# _joined_ functions {#joined}

## Description

For any "joined" function remember from above that the main difference is that
given the additional option to filter after the join, it means that selection of
the observations to add from the additional dataframe can depend on variables from
both this and the original dataframe.

So, as with the "merged" functions the additional dataframe can first be filtered
based on a user-defined condition passed into the `filter_add` argument. Then it
is joined with the input dataframe by user-specified variables passed into `by_vars`
argument. The joined dataframe can then be further restricted by the `filter_join`
condition, before optionally selecting the first or last observation for each by
group (`order` and `mode` arguments). Finally the joined variables from the additional
dataframe can be renamed (`new_vars` argument).

To illustrate this, imagine that in the above randomization date example you only
wanted the date populated for patients randomized within 30 days of first treatment.
You would now need to check both the additional `DS` and the original `ADSL` dataframes,
so this would not be possible in one step using the "merged" functions.

## Examples

The above mentioned randomization date variable (let's call it `RAND30DT` here)
would use `derive_vars_joined()` as follows.

```{r eval=TRUE}
# Join randomization date to ADSL only for safety population patients
adsl_05 <- adsl_04 %>%
  derive_vars_joined(
    dataset_add = ds_ext,
    filter_add = DSDECOD == "RANDOMIZED",
    by_vars = vars(STUDYID, USUBJID),
    new_vars = vars(RAND30DT = DSSTDT),
    filter_join = DSSTDT >= TRTSDT - 30
  )
```

```{r, eval=TRUE, echo=FALSE}
dataset_vignette(
  dataset = adsl_05,
  display_vars = vars(USUBJID, RANDDT, TRTSDT, RAND30DT)
)
```

Now, let's consider another example of `derive_vars_joined()`. For this we'll need to
additionally use the `join_vars` argument, which is where the user needs to call out
any variables from the additional dataframe being used in `filter_join`. Note: If a
specified variable exists in both dataframes, then the suffix ".join" is added to the
variable from the additional dataframe.

In this case we want to add a datacut flag to `AE`, only for events occurring up
to and including this day. So in this case `AE` is the original dataframe and the
datacut source is the additional dataframe that needs to be joined.

```{r eval=TRUE}
# Create a unique datacut day for each patient
datacut <- tribble(
  ~USUBJID,      ~DCUTDY, ~DCUTFL,
  "01-701-1047",      25, "Y",
  "01-701-1111",       5, "Y"
)

# Join datacut flag to AE only for events up to and including this date
ae_01 <- ae %>%
  derive_vars_joined(
    dataset_add = datacut,
    by_vars = vars(USUBJID),
    new_vars = vars(DCUTFL),
    join_vars = vars(DCUTDY),
    filter_join = AESTDY <= DCUTDY
  )
```

```{r, eval=TRUE, echo=FALSE}
ae_01 %>%
  select(USUBJID, AEDECOD, AESTDY, DCUTFL) %>%
  arrange(USUBJID, AESTDY) %>%
  dataset_vignette(display_vars = vars(USUBJID, AEDECOD, AESTDY, DCUTFL))
```

The `derive_vars_joined()` function could also be used to join the original dataframe
back with itself, which may be necessary if you're comparing across different observations.
For example, the below shows how a "nadir" calculation would be achieved which checks
for the worst value prior up to that observation.

Here is how you would derive the highest severity AE the patient has occurred
post-baseline up to and excluding the current AE day.

```{r eval=TRUE}
# Add a numeric version of severity for sorting with severe=1, moderate=2, mild=3
ae_ext <- ae_01 %>%
  mutate(TEMP_SEVN = as.integer(factor(AESEV, levels = c("SEVERE", "MODERATE", "MILD"))))

# Derive nadir severity (AENADSEV)
ae_02 <- ae_ext %>%
  derive_vars_joined(
    dataset_add = ae_ext,
    filter_add = AESTDY > 0,
    by_vars = vars(USUBJID),
    order = vars(TEMP_SEVN),
    new_vars = vars(AENADSEV = AESEV),
    join_vars = vars(AESTDY),
    filter_join = AESTDY.join < AESTDY,
    mode = "first",
    check_type = "none"
  )
```

```{r, eval=TRUE, echo=FALSE}
ae_02 %>%
  select(USUBJID, AEDECOD, AESTDY, AESEV, AENADSEV) %>%
  arrange(USUBJID, AESTDY) %>%
  dataset_vignette(display_vars = vars(USUBJID, AEDECOD, AESTDY, AESEV, AENADSEV))
```

There exists a further function `derive_var_confirmation_flag()`, which follows
a similar principle to the "joined" set as explained here. This can be used for whenever
one set of observations "confirms" another set in a dataframe, i.e. the flag is only
set for your observation if both the observation and some other observation meet both
the user-specified criteria.

# _extreme_ functions {#extreme}

## Description

The "extreme" functions are used for finding the first or last observation
(from `order` and `mode` arguments) within each by group (`by_vars` argument).

These can then be output as a flag variable, a date, or as new records dependent
on which respective function is used between `derive_var_extreme_flag()`,
`derive_var_extreme_dt()` or `derive_extreme_records()` for example.

## Example

Building on the above case, let's use `derive_var_extreme_flag()` to now flag
the earliest observation with the highest severity per patient.

```{r eval=TRUE}
# Highest severity flag (AEHSEVFL)
ae_03 <- ae_02 %>%
  derive_var_extreme_flag(
    new_var = AEHSEVFL,
    by_vars = vars(USUBJID),
    order = vars(TEMP_SEVN, AESTDY, AESEQ),
    mode = "first"
  )
```

```{r, eval=TRUE, echo=FALSE}
ae_03 %>%
  select(USUBJID, AESTDY, AESEQ, AESEV, AEHSEVFL) %>%
  arrange(USUBJID, AESTDY, AESEQ) %>%
  dataset_vignette(display_vars = vars(USUBJID, AESTDY, AESEQ, AESEV, AEHSEVFL))
```

Common further examples of usage of this set of functions could be using
`derive_var_extreme_dt()` for last known alive date calculation or `derive_extreme_records()`
for creating derived minimum or maximum value records.
