---
title: "Higher Order Functions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Higher Order Functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(admiraldev)
```

# Introduction

This vignette explains some of the more advanced options of `{admiral}` related
to higher order functions. A higher order function is a function that takes another
function as input. By introducing these higher order functions, we intend to give
the user greater power over our derivation functions, whilst trying to negate the
need for adding additional `{admiral}` functions or arguments, or the user needing
many separate steps.

The functions covered here are:

* `call_derivation()`: Call a single derivation multiple times with some
arguments being fixed across iterations and others varying
* `restrict_derivation()`: Execute a single derivation on a subset of the input dataset
* `slice_derivation()`: The input dataset is split into slices (subsets) and for
each slice a single derivation is called separately. Some or all arguments of the
derivation may vary depending on the slice.

## Required Packages

The examples of this vignette require the following packages.

For example purpose, the ADSL dataset---which is included
in `{admiral}`---and the SDTM datasets from `{admiral.test}` are used.

```{r, warning=FALSE, message=FALSE}
library(admiral)
library(admiral.test)
library(dplyr, warn.conflicts = FALSE)

data("admiral_adsl")
data("admiral_ae")
data("admiral_vs")
adsl <- admiral_adsl
ae <- convert_blanks_to_na(admiral_ae)
vs <- convert_blanks_to_na(admiral_vs)
```
```{r echo=FALSE}
adsl <- filter(adsl, USUBJID %in% c("01-701-1111", "01-705-1393"))
ae <- filter(ae, USUBJID %in% c("01-701-1111", "01-705-1393"))
vs <- filter(vs, USUBJID %in% c("01-701-1015"))
```

The following code creates a minimally viable ADAE dataset to be used where
needed in the following examples.

```{r}
adae <- ae %>%
  left_join(adsl, by = c("STUDYID", "USUBJID")) %>%
  derive_vars_dt(
    new_vars_prefix = "AST",
    dtc = AESTDTC,
    highest_imputation = "M"
  ) %>%
  mutate(TRTEMFL = if_else(ASTDT >= TRTSDT, "Y", NA_character_))
```

# Call Derivation

This function exists purely for convenience to save the user repeating numerous
similar derivation function calls. It is best used when multiple derived variables
have very similar specifications with only slight variations.

As an example, imagine the case where all the parameters in a BDS ADaM required
both a highest value flag and a lowest value flag.

Here is an example of how to achieve this **without** using `call_derivation()`: 

```{r}
vs_without <- vs %>%
  derive_var_extreme_flag(
    by_vars = vars(USUBJID, VSTESTCD),
    order = vars(VSORRES, VSSEQ),
    new_var = AHIFL,
    mode = "last"
  ) %>%
  derive_var_extreme_flag(
    by_vars = vars(USUBJID, VSTESTCD),
    order = vars(VSORRES, VSSEQ),
    new_var = ALOFL,
    mode = "first"
  )
```

```{r, eval=TRUE, echo=FALSE}
vs_without %>%
  arrange(USUBJID, VSTESTCD, VSDY, VSSEQ) %>%
  dataset_vignette(
    display_vars = vars(USUBJID, VSTESTCD, VSORRES, ALOFL, AHIFL),
    filter = VSTESTCD %in% c("TEMP", "WEIGHT")
  )
```

Here is an example of how to achieve the same **with** using `call_derivation()`,
where any different arguments are passed using `params()`: 

```{r}
vs_with <- vs %>%
  call_derivation(
    derivation = derive_var_extreme_flag,
    variable_params = list(
      params(new_var = AHIFL, mode = "last"),
      params(new_var = ALOFL, mode = "first")
    ),
    by_vars = vars(USUBJID, VSTESTCD),
    order = vars(VSORRES, VSSEQ)
  )
```

```{r, eval=TRUE, echo=FALSE}
vs_with %>%
  arrange(USUBJID, VSTESTCD, VSDY, VSSEQ) %>%
  dataset_vignette(
    display_vars = vars(USUBJID, VSTESTCD, VSORRES, ALOFL, AHIFL),
    filter = VSTESTCD %in% c("TEMP", "WEIGHT")
  )
```

In the example, you can see how in these higher order functions, `derivation`
is where the user supplies the name of the derivation function to apply, with
no trailing parentheses required. Then `variable_params` is used to pass a
list of the different arguments needed for each derived variable.

The advantage of this higher order function would be further highlighted with
examples where more than two variable derivations had similar needs, such as the
below case where multiple time to AE parameters are derived in one call. 
Note that this example relies on pre-defined `tte_source` objects, as explained
at [Creating a BDS Time-to-Event ADaM](bds_tte.html).

```{r}
adaette <- call_derivation(
  derivation = derive_param_tte,
  variable_params = list(
    params(
      event_conditions = list(ae_event),
      set_values_to = vars(PARAMCD = "TTAE")
    ),
    params(
      event_conditions = list(ae_ser_event),
      set_values_to = vars(PARAMCD = "TTSERAE")
    ),
    params(
      event_conditions = list(ae_sev_event),
      set_values_to = vars(PARAMCD = "TTSEVAE")
    ),
    params(
      event_conditions = list(ae_wd_event),
      set_values_to = vars(PARAMCD = "TTWDAE")
    )
  ),
  dataset_adsl = adsl,
  source_datasets = list(adsl = adsl, adae = adae),
  censor_conditions = list(lastalive_censor)
)
```

```{r, eval=TRUE, echo=FALSE}
adaette %>%
  select(USUBJID, PARAMCD, STARTDT, ADT, CNSR, EVNTDESC, SRCDOM, SRCVAR) %>%
  arrange(USUBJID, PARAMCD) %>%
  dataset_vignette(display_vars = vars(USUBJID, PARAMCD, STARTDT, ADT, CNSR, EVNTDESC, SRCDOM, SRCVAR))
```

Developing your ADaM scripts this way using `call_derivation()` could give the
following benefits:

* code becomes more efficient and readable
* maintenance would be eased in case of specification changes
* downstream quality checking would require less effort

# Restrict Derivation

The idea behind this function is that sometimes you want to apply a
derivation only for certain records from the input dataset. Introducing
`restrict_derivation()` therefore gives the users the ability to achieve
this across any function, without each function needing to have such an
argument to allow for this.

An example would be if you wanted to flag the first occurring AE with the
highest severity for each patient, but you only wanted to do this for
records occurring on or after study day 1.

Here is how you could achieve this using `restrict_derivation()`,
where the function arguments are passed using `params()` and the restriction
criteria is given using `filter`:

```{r}
ae <- ae %>%
  mutate(TEMP_AESEVN = as.integer(factor(AESEV, levels = c("SEVERE", "MODERATE", "MILD")))) %>%
  restrict_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      new_var = AHSEVFL,
      by_vars = vars(USUBJID),
      order = vars(TEMP_AESEVN, AESTDY, AESEQ),
      mode = "first"
    ),
    filter = AESTDY >= 1
  )
```

```{r, eval=TRUE, echo=FALSE}
ae %>%
  arrange(USUBJID, AESTDY, AESEQ, desc(TEMP_AESEVN)) %>%
  dataset_vignette(
    display_vars = vars(USUBJID, AEDECOD, AESTDY, AESEQ, AESEV, AHSEVFL)
  )
```

# Slice Derivation

This function in a way combines the features of the above two. It allows a single
derivation to be applied with different arguments for different slices (subsets)
of records from the input dataset. You could do this with separate `restrict_derivation()`
calls for each different set of records, but `slice_derivation()` allows
to achieve this in one call.

An example would be if you wanted to achieve the same derivation as above
for records occurring on or after study day 1, but for pre-treatment AEs
you wanted to flag only the last occurring AE.

Here is how you could achieve this using `slice_derivation()`,
where the function arguments are passed using `params()` and via the different
slices controlled by `filter`:

```{r}
ae <- ae %>%
  slice_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      new_var = AHSEV2FL,
      by_vars = vars(USUBJID)
    ),
    derivation_slice(
      filter = AESTDY >= 1,
      args = params(order = vars(TEMP_AESEVN, AESTDY, AESEQ), mode = "first")
    ),
    derivation_slice(
      filter = TRUE,
      args = params(order = vars(AESTDY, AESEQ), mode = "last")
    )
  )
```

```{r, eval=TRUE, echo=FALSE}
ae %>%
  arrange(USUBJID, AESTDY, AESEQ, desc(TEMP_AESEVN)) %>%
  dataset_vignette(
    display_vars = vars(USUBJID, AEDECOD, AESTDY, AESEQ, AESEV, AHSEV2FL)
  )
```

As you can see in the example, the `derivation_slice` ordering is important.
Here we addressed all the AEs on or after study day 1 first, and then we used
`filter = TRUE` option to catch all remaining records (in this case
pre-treatment AEs).

The ordering is perhaps shown even more when we look at the below example where
three slices are taken. Remember that observations that match with more than one
slice are only considered for the first matching slice. So in this case we're
creating a flag for each patient for the record with the first severe AE, and
then the first moderate AE, and finally flagging the last occurring AE where
not severe or moderate.

```{r}
ae <- ae %>%
  slice_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      new_var = AHSEV3FL,
      by_vars = vars(USUBJID)
    ),
    derivation_slice(
      filter = AESEV == "SEVERE",
      args = params(order = vars(AESTDY, AESEQ), mode = "first")
    ),
    derivation_slice(
      filter = AESEV == "MODERATE",
      args = params(order = vars(AESTDY, AESEQ), mode = "first")
    ),
    derivation_slice(
      filter = TRUE,
      args = params(order = vars(AESTDY, AESEQ), mode = "last")
    )
  )
```

```{r, eval=TRUE, echo=FALSE}
ae %>%
  arrange(USUBJID, AESTDY, AESEQ) %>%
  dataset_vignette(
    display_vars = vars(USUBJID, AEDECOD, AESTDY, AESEQ, AESEV, AHSEV3FL)
  )
```

The order is only important when the slices are not mutually exclusive, so
in the above case the moderate AE slice could have been above the severe AE
slice, for example, and there would have been no difference to the result.
However the third slice had to come last to check all remaining (i.e. not severe
or moderate) records only.
