---
title: "Date and Time Imputation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Date and Time Imputation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(admiraldev)
```

# Introduction

Date and time is collected in SDTM as character values using the extended [ISO
8601](https://en.wikipedia.org/wiki/ISO_8601) format. For example,
`"2019-10-9T13:42:00"`. It allows that some parts of the date or time are
missing, e.g., `"2019-10"` if the day and the time is unknown.

The ADaM timing variables like `ADTM` (Analysis Datetime) or `ADY` (Analysis
Relative Day) are numeric variables. They can be derived only if the date or
datetime is complete. Therefore `{admiral}` provides imputation functions which fill
in missing date or time parts according to certain imputation rules.

In {admiral} we use only two functions `derive_vars_dt()` and
`derive_vars_dtm()` for date and datetime imputations respectively. In all other
functions where dates can be passed as an argument, we expect full dates or
datetimes (unless otherwise specified), so if any possibility of partials then
these functions should be used as a first step to make the required imputation.

## Required Packages

The examples of this vignette require the following packages.

```{r, warning=FALSE, message=FALSE}
library(admiral)
library(lubridate)
library(tibble)
library(dplyr, warn.conflicts = FALSE)
```

# Imputation Rules

In {admiral} we don't allow users to pick any single part of the date/time to
impute, we only enable to impute up to a highest level, i.e. you couldn't choose
to say impute months, but not days.

The simplest imputation rule is to set the missing parts to a fixed value. For
example

```{r}
impute_dtc_dtm(
  "2019-10",
  highest_imputation = "M",
  date_imputation = "01-01",
  time_imputation = "00:00:00"
)
```

Sometimes this does not work as it would result in invalid dates, e.g.,

```{r}
impute_dtc_dtm(
  "2019-02",
  highest_imputation = "M",
  date_imputation = "02-31",
  time_imputation = "00:00:00"
)
```

Therefore the keywords `"first"` or `"last"` can be specified to request that
missing parts are replaced by the first or last possible value:

```{r}
impute_dtc_dtm(
  "2019-02",
  highest_imputation = "M",
  date_imputation = "last",
  time_imputation = "00:00:00"
)
```

For dates, there is the additional option to use keyword `"mid"` to impute
missing day to `15` or missing day and month to `06-30`, but note the
different behaviour below depending on `preserve` argument for case when month
only is missing:

```{r}
dates <- c(
  "2019-02",
  "2019",
  "2019---01"
)
impute_dtc_dtm(
  dates,
  highest_imputation = "M",
  date_imputation = "mid",
  time_imputation = "00:00:00",
  preserve = FALSE
)
impute_dtc_dtm(
  dates,
  highest_imputation = "M",
  date_imputation = "mid",
  time_imputation = "00:00:00",
  preserve = TRUE
)
```

If you wanted to achieve a similar result by replacing any missing part of the
date with a fixed value `06-15`, this is also possible, but note the difference
in days for cases when month is missing:

```{r}
dates <- c(
  "2019-02",
  "2019",
  "2019---01"
)
impute_dtc_dtm(
  dates,
  highest_imputation = "M",
  date_imputation = "06-15",
  time_imputation = "00:00:00"
)
```
## Imputation Level

The imputation level, i.e., which components are imputed if they are missing, is
controlled by the `highest_imputation` argument. All components up to the
specified level are imputed.

```{r}
dates <- c(
  "2019-02-03T12:30:15",
  "2019-02-03T12:30",
  "2019-02-03",
  "2019-02",
  "2019"
)

# Do not impute
impute_dtc_dtm(
  dates,
  highest_imputation = "n"
)

# Impute seconds only
impute_dtc_dtm(
  dates,
  highest_imputation = "s"
)

# Impute time (hours, minutes, seconds) only
impute_dtc_dtm(
  dates,
  highest_imputation = "h"
)

# Impute days and time
impute_dtc_dtm(
  dates,
  highest_imputation = "D"
)

# Impute date (months and days) and time
impute_dtc_dtm(
  dates,
  highest_imputation = "M"
)
```

For imputation of years (`highest_imputation = "Y"`) see next section.

## Minimum/Maximum Dates

In some scenarios the imputed date should not be before or after certain dates.
For example an imputed date after data cut off date or death date is not
desirable. The `{admiral}` imputation functions provide the `min_dates` and
`max_dates` argument to specify those dates. For example:

```{r}
impute_dtc_dtm(
  "2019-02",
  highest_imputation = "M",
  date_imputation = "last",
  time_imputation = "last",
  max_dates = list(ymd("2019-01-14"), ymd("2019-02-25"))
)
```

It is ensured that the imputed date is not after any of the specified dates.
Only dates which are in the range of possible dates of the dtc value are
considered. The possible dates are defined by the missing parts of the dtc date,
i.e., for "2019-02" the possible dates range from "2019-02-01" to "2019-02-28".
Thus "2019-01-14" is ignored. This ensures that the non-missing parts of the dtc
date are not changed.

If the `min_dates` or `max_dates` argument is specified, it is also possible to
impute completely missing dates. For `date_imputation = "first"` the `min_dates`
argument must be specified and for `date_imputation = "last"` the `max_dates`
argument. For other imputation rules imputing the year is not possible.

```{r}
# Impute year to first
impute_dtc_dtm(
  c("2019-02", NA),
  highest_imputation = "Y",
  min_dates = list(ymd("2019-01-14"), ymd("2019-02-25"))
)

# Impute year to last
impute_dtc_dtm(
  c("2019-02", NA),
  highest_imputation = "Y",
  date_imputation = "last",
  time_imputation = "last",
  max_dates = list(ymd("2019-01-14"), ymd("2019-02-25"))
)
```

# Imputation Flags

ADaM requires that date or datetime variables for which imputation was used are
accompanied by date and/or time imputation flag variables (`*DTF` and `*TMF`,
e.g., `ADTF` and `ATMF` for `ADTM`). These variables indicate the highest level
that was imputed, e.g., if minutes and seconds were imputed, the imputation flag
is set to `"M"`. The `{admiral}` functions which derive imputed variables are also
adding the corresponding imputation flag variables.

Note: The `{admiral}` datetime imputation function provides the `ignore_seconds_flag`
argument which can be set to `TRUE` in cases where seconds were never collected.
This is due to the following from ADaM IG: For a given SDTM DTC variable, if only
hours and minutes are ever collected, and seconds are imputed in `*DTM` as `00`,
then it is not necessary to set `*TMF` to `"S"`.

# Imputation Functions

`{admiral}` provides the following functions for imputation:

- `derive_vars_dt()`: Adds a date variable and a date imputation flag variable
(optional) based on a --DTC variable and imputation rules.
- `derive_vars_dtm()`: Adds a datetime variable, a date imputation flag variable,
and a time imputation flag variable (both optional) based on a --DTC variable
and imputation rules.
- `impute_dtc_dtm()`: Returns a complete ISO 8601 datetime or `NA` based on a
partial ISO 8601 datetime and imputation rules.
- `impute_dtc_dt()`: Returns a complete ISO 8601 date (without time) or `NA`
based on a partial ISO 8601 date(time) and imputation rules.
- `convert_dtc_to_dt()`: Returns a date if the input ISO 8601 date is complete.
Otherwise, `NA` is returned.
- `convert_dtc_to_dtm()`: Returns a datetime if the input ISO 8601 date is complete 
(with missing time replaced by `"00:00:00"` as default). Otherwise, NA is returned.
- `compute_dtf()`: Returns the date imputation flag.
- `compute_tmf()`: Returns the time imputation flag.

# Examples

## Create an Imputed Datetime and Date Variable and Imputation Flag Variables

The `derive_vars_dtm()` function derives an imputed datetime variable and the
corresponding date and time imputation flags. The imputed date variable can be
derived by using the `derive_vars_dtm_to_dt()` function. It is not necessary and
advisable to perform the imputation for the date variable if it was already done
for the datetime variable. CDISC considers the datetime and the date variable as
two representations of the same date. Thus the imputation must be the same and the
imputation flags are valid for both the datetime and the date variable.

```{r}
ae <- tribble(
  ~AESTDTC,
  "2019-08-09T12:34:56",
  "2019-04-12",
  "2010-09",
  NA_character_
) %>%
  derive_vars_dtm(
    dtc = AESTDTC,
    new_vars_prefix = "AST",
    highest_imputation = "M",
    date_imputation = "first",
    time_imputation = "first"
  ) %>%
  derive_vars_dtm_to_dt(vars(ASTDTM))
```
```{r, echo=FALSE}
dataset_vignette(ae)
```

## Create an Imputed Date Variable and Imputation Flag Variable

If an imputed date variable without a corresponding datetime variable is
required, it can be derived by the `derive_vars_dt()` function.

```{r}
ae <- tribble(
  ~AESTDTC,
  "2019-08-09T12:34:56",
  "2019-04-12",
  "2010-09",
  NA_character_
) %>%
  derive_vars_dt(
    dtc = AESTDTC,
    new_vars_prefix = "AST",
    highest_imputation = "M",
    date_imputation = "first"
  )
```
```{r, echo=FALSE}
dataset_vignette(ae)
```

## Impute Time without Imputing Date

If the time should be imputed but not the date, the `highest_imputation` argument
should be set to `"h"`. This results in `NA` if the date is partial. As
no date is imputed the date imputation flag is not created.

```{r}
ae <- tribble(
  ~AESTDTC,
  "2019-08-09T12:34:56",
  "2019-04-12",
  "2010-09",
  NA_character_
) %>%
  derive_vars_dtm(
    dtc = AESTDTC,
    new_vars_prefix = "AST",
    highest_imputation = "h",
    time_imputation = "first"
  )
```
```{r, echo=FALSE}
dataset_vignette(ae)
```

## Avoid Imputed Dates Before a Particular Date
Usually the adverse event start date is imputed as the earliest date of all
possible dates when filling the missing parts. The result may be a date before
treatment start date. This is not desirable because the adverse event would not
be considered as treatment emergent and excluded from the adverse event
summaries. This can be avoided by specifying the treatment start date variable
(`TRTSDTM`) for the `min_dates` argument.

Please note that `TRTSDTM` is used as imputed date only if the non missing date
and time parts of `AESTDTC` coincide with those of `TRTSDTM`. Therefore
`2019-10` is not imputed as `2019-11-11 12:34:56`. This ensures that collected
information is not changed by the imputation.

```{r}
ae <- tribble(
  ~AESTDTC,              ~TRTSDTM,
  "2019-08-09T12:34:56", ymd_hms("2019-11-11T12:34:56"),
  "2019-10",             ymd_hms("2019-11-11T12:34:56"),
  "2019-11",             ymd_hms("2019-11-11T12:34:56"),
  "2019-12-04",          ymd_hms("2019-11-11T12:34:56")
) %>%
  derive_vars_dtm(
    dtc = AESTDTC,
    new_vars_prefix = "AST",
    highest_imputation = "M",
    date_imputation = "first",
    time_imputation = "first",
    min_dates = vars(TRTSDTM)
  )
```
```{r, echo=FALSE}
dataset_vignette(ae)
```

## Avoid Imputed Dates After a Particular Date

If a date is imputed as the latest date of all possible dates when filling the
missing parts, it should not result in dates after data cut off or death. This
can be achieved by specifying the dates for the `max_dates` argument.

Please note that non missing date parts are not changed. Thus `2019-12-04` is
imputed as `2019-12-04 23:59:59` although it is after the data cut off date. It
may make sense to replace it by the data cut off date but this is not part of
the imputation. It should be done in a separate data cleaning or data cut off
step.
```{r}
ae <- tribble(
  ~AEENDTC,              ~DTHDT,            ~DCUTDT,
  "2019-08-09T12:34:56", ymd("2019-11-11"), ymd("2019-12-02"),
  "2019-11",             ymd("2019-11-11"), ymd("2019-12-02"),
  "2019-12",             NA,                ymd("2019-12-02"),
  "2019-12-04",          NA,                ymd("2019-12-02")
) %>%
  derive_vars_dtm(
    dtc = AEENDTC,
    new_vars_prefix = "AEN",
    highest_imputation = "M",
    date_imputation = "last",
    time_imputation = "last",
    max_dates = vars(DTHDT, DCUTDT)
  )
```
```{r, echo=FALSE}
dataset_vignette(ae)
```

## Imputation Without Creating a New Variable

If imputation is required without creating a new variable the
`convert_dtc_to_dt()` function can be called to obtain a vector of imputed
dates. It can be used for example in conditions:

```{r}
mh <- tribble(
  ~MHSTDTC,     ~TRTSDT,
  "2019-04",    ymd("2019-04-15"),
  "2019-04-01", ymd("2019-04-15"),
  "2019-05",    ymd("2019-04-15"),
  "2019-06-21", ymd("2019-04-15")
) %>%
  filter(
    convert_dtc_to_dt(
      MHSTDTC,
      highest_imputation = "M",
      date_imputation = "first"
    ) < TRTSDT
  )
```
```{r, echo=FALSE}
dataset_vignette(mh)
```

## Using More Than One Imputation Rule for a Variable

Using different imputation rules depending on the observation can be done by
using `slice_derivation()`.

```{r}
vs <- tribble(
  ~VSDTC,                ~VSTPT,
  "2019-08-09T12:34:56", NA,
  "2019-10-12",          "PRE-DOSE",
  "2019-11-10",          NA,
  "2019-12-04",          NA
) %>%
  slice_derivation(
    derivation = derive_vars_dtm,
    args = params(
      dtc = VSDTC,
      new_vars_prefix = "A"
    ),
    derivation_slice(
      filter = VSTPT == "PRE-DOSE",
      args = params(time_imputation = "first")
    ),
    derivation_slice(
      filter = TRUE,
      args = params(time_imputation = "last")
    )
  )
```
```{r, echo=FALSE}
dataset_vignette(vs)
```
