---
title: "Creating an OCCDS ADaM"
output: 
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating an OCCDS ADaM}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

This article describes creating an OCCDS ADaM. Examples are currently 
presented and tested in the context of ADAE.
However, the examples could be applied to other OCCDS ADaMs such as 
ADCM, ADMH, ADDV, etc.

**Note**: *All examples assume CDISC SDTM and/or ADaM format as input unless 
otherwise specified.*

# Programming Workflow

* [Read in Data](#readdata)
* [Derive/Impute End and Start Analysis Date/time and Relative Day](#datetime)
* [Derive Durations](#duration)
* [Derive ATC variables](#atc)
* [Derive Planned and Actual Treatment](#trtpa)
* [Derive Date/Date-time of Last Dose](#last_dose)
* [Derive Severity, Causality, and Toxicity Grade](#severity)
* [Derive Treatment Emergent Flag](#trtflag)
* [Derive Occurrence Flags](#occflag)
* [Derive Query Variables](#query)
* [Add ADSL variables](#adsl_vars)

## Read in Data {#readdata}

To start, all data frames needed for the creation of `ADAE` should be read into
the environment. This will be a company specific process. Some of the data
frames needed may be `AE`, `ADSL`, `SUPPAE`, 'SUPPDM`.

For example purpose, the CDISC Pilot SDTM and ADaM datasets---which are included in `{admiral}`---are used.

```{r, message=FALSE, warning=FALSE}
library(admiral)
library(dplyr)
library(admiral.test)
library(lubridate)

data("ae")
data("suppae")
data("adsl")
```

The `SUPPAE` domain can be joined to the 'AE` domain using the function `derive_vars_suppqual()`.

This function will transpose the supplemental SDTM domain (e.g. `SUPPAE`) and
join the transposed data to the parent domain (e.g. `ae`) by `STUDYID`, 
`USUBJID` using the `IDVAR` and `IDVARVAL` as an additional join variable.

Example call:

To derive Supplemental Qualifiers, `derive_vars_suppqual()` can be used. 

```{r eval=TRUE}
ae <- derive_vars_suppqual(ae, suppae)
```

```{r, eval=TRUE, echo=FALSE}
dataset_vignette(ae, 
                 display_vars = vars(USUBJID, AETERM, AEDECOD,AESTDTC, AETRTEM),
                 filter = USUBJID == "01-701-1015")
```

At this step, it may be useful to join `ADSL` to your `AE` domain. Only the 
`ADSL` variables used for derivations are selected at this step. The rest of the
relevant `ADSL` would be added later.

```{r eval=TRUE}

adsl_vars <- vars(TRTSDT, TRTEDT, TRT01A, TRT01P, DTHDT, EOSDT)

adae <- left_join(
  ae,
  select(adsl, STUDYID, USUBJID, !!!adsl_vars),
  by = c("STUDYID", "USUBJID")
)
```

```{r, eval=TRUE, echo=FALSE}
dataset_vignette(adae, 
                 display_vars = vars(USUBJID, AESEQ, AETERM, AESTDTC, TRTSDT,
                                     TRTEDT, TRT01A, TRT01P, DTHDT, EOSDT))
```
## Derive/Impute End and Start Analysis Date/time and Relative Day {#datetime}

This part derives `ASTDTM`, `ASTDT`, `ASTDY`, `AENDTM`, `AENDT`, and `AENDY`. 
The function `derive_vars_dtm()` can be used to derive `ASTDTM` and `AENDTM` 
where `ASTDTM` could be company-specific. 
`ASTDT` and `AENDT` can be derived from `ASTDTM` and `AENDTM`, respectively using
function `derive_vars_dtm_to_dt`. 
`derive_var_astdy()` and `derive_var_aendy()` can be used to create `ASTDY` and 
`AENDY`, respectively.

```{r eval=TRUE}
adae <- adae %>% 
  derive_vars_dtm(
    dtc = AESTDTC,
    new_vars_prefix = "AST",
    date_imputation = "first",
    time_imputation = "first",
    min_dates = vars(TRTSDT)
  ) %>% 
  derive_vars_dtm(
    dtc = AEENDTC,
    new_vars_prefix = "AEN",
    date_imputation = "last",
    time_imputation = "last",
    max_dates = vars(DTHDT, EOSDT)
  ) %>%
  derive_vars_dtm_to_dt(vars(ASTDTM, AENDTM)
  ) %>% 
  derive_var_astdy(
    reference_date = TRTSDT, 
    date = ASTDT
  ) %>%
  derive_var_aendy(
    reference_date = TRTSDT, 
    date = AENDT
  )
```

```{r, eval=TRUE, echo=FALSE}
dataset_vignette(adae, 
                 display_vars = vars(USUBJID, AESTDTC, AEENDTC, ASTDTM, ASTDT,
                                     ASTDY, AENDTM, AENDT, AENDY))
```

See also [Date and Time Imputation](imputation.html).

## Derive Durations {#duration}

The function `derive_vars_duration()` can be used to create the variables 
`ADURN` and `ADURU`. 

```{r eval=TRUE}
adae <- adae %>%
  derive_vars_duration(
    new_var = ADURN,
    new_var_unit = ADURU,
    start_date = ASTDT,
    end_date = AENDT
  )
```

```{r, eval=TRUE, echo=FALSE}
dataset_vignette(adae, 
                 display_vars = vars(USUBJID, AESTDTC, AEENDTC, ASTDT, AENDT, 
                                     ADURN, ADURU))
```

## Derive ATC variables {#atc}

The function `derive_vars_atc()` can be used to derive 
ATC Class Variables.

It helps to add Anatomical Therapeutic Chemical class variables from `FACM` to `ADCM`.

The expected result is the input dataset with ATC variables added.


```{r eval=TRUE}
cm <- tibble::tribble(
         ~USUBJID, ~CMGRPID,  ~CMREFID,            ~CMDECOD,
   "BP40257-1001",     "14", "1192056",       "PARACETAMOL",
   "BP40257-1001",     "18", "2007001",        "SOLUMEDROL",
   "BP40257-1002",     "19", "2791596",    "SPIRONOLACTONE"
 )
 facm <- tibble::tribble(
         ~USUBJID, ~FAGRPID,  ~FAREFID,   ~FATESTCD, ~FASTRESC,
   "BP40257-1001",      "1", "1192056",  "CMATC1CD",       "N",
   "BP40257-1001",      "1", "1192056",  "CMATC2CD",     "N02",
   "BP40257-1001",      "1", "1192056",  "CMATC3CD",    "N02B",
   "BP40257-1001",      "1", "1192056",  "CMATC4CD",   "N02BE",

   "BP40257-1001",      "1", "2007001",  "CMATC1CD",       "D",
   "BP40257-1001",      "1", "2007001",  "CMATC2CD",     "D10",
   "BP40257-1001",      "1", "2007001",  "CMATC3CD",    "D10A",
   "BP40257-1001",      "1", "2007001",  "CMATC4CD",   "D10AA",
   "BP40257-1001",      "2", "2007001",  "CMATC1CD",       "D",
   "BP40257-1001",      "2", "2007001",  "CMATC2CD",     "D07",
   "BP40257-1001",      "2", "2007001",  "CMATC3CD",    "D07A",
   "BP40257-1001",      "2", "2007001",  "CMATC4CD",   "D07AA",
   "BP40257-1001",      "3", "2007001",  "CMATC1CD",       "H",
   "BP40257-1001",      "3", "2007001",  "CMATC2CD",     "H02",
   "BP40257-1001",      "3", "2007001",  "CMATC3CD",    "H02A",
   "BP40257-1001",      "3", "2007001",  "CMATC4CD",   "H02AB",

   "BP40257-1002",      "1", "2791596",  "CMATC1CD",       "C",
   "BP40257-1002",      "1", "2791596",  "CMATC2CD",     "C03",
   "BP40257-1002",      "1", "2791596",  "CMATC3CD",    "C03D",
   "BP40257-1002",      "1", "2791596",  "CMATC4CD",   "C03DA"
 )

derive_vars_atc(cm, facm)
```

## Derive Planned and Actual Treatment {#trtpa}

`TRTA` and `TRTP` must correlate to treatment `TRTxxP` and/or `TRTxxA` in ADSL.
The derivation of `TRTA` and `TRTP` for a record are protocol and analysis 
specific.  admiral does not currently have functionality to assist with `TRTA` 
and `TRTP` assignment.

However, an example of a simple implementation could be:

```{r eval=TRUE}
adae <- mutate(adae, TRTP = TRT01P, TRTA = TRT01A)

count(adae, TRTP, TRTA, TRT01P, TRT01A)

```



## Derive Date/Date-time of Last Dose {#last_dose}


The function `derive_var_last_dose_date()` can be used to derive the last dose date before the start of the event. 

Additionally, this function can also provide the traceability variables 
(e.g. `LDOSEDOM`, `LDOSESEQ`) using the `traceability_vars` argument.


```{r eval=TRUE}
data(ex_single)
adae <- adae %>% 
  derive_var_last_dose_date(
    ex_single,
    filter_ex = (EXDOSE > 0 | (EXDOSE == 0 & grepl("PLACEBO", EXTRT))) &
      nchar(EXENDTC) >= 10,
    dose_date = EXSTDTC,
    analysis_date = ASTDT,
    single_dose_condition = (EXSTDTC == EXENDTC),
    new_var = LDOSEDTM,
    output_datetime = TRUE
  )
```

```{r, eval=TRUE, echo=FALSE}
dataset_vignette(adae, 
                 display_vars = vars(USUBJID, AEDECOD, AESEQ, AESTDTC, AEENDTC, 
                                     ASTDT, AENDT, LDOSEDTM
                                     ))
```

## Derive Severity, Causality, and Toxicity Grade {#severity}

The variables `ASEV`, `AREL`, and `ATOXGR` can be added by simply
`mutate()` if no imputation is required.

```{r eval=TRUE}
adae <- adae %>%
  mutate(
    ASEV = AESEV, 
    AREL = AEREL
  )
```

## Derive Treatment Emergent Flag {#trtflag}

To derive the treatment emergent flag `TRTEMFL`, one can use simple
`dplyr::mutate()`. In the example below, we use 30 days in the flag derivation.

```{r eval=TRUE}
adae <- adae %>% 
  mutate(
    TRTEMFL = ifelse(ASTDT >= TRTSDT & ASTDT <= TRTEDT + days(30), "Y", NA_character_)
  )
```

```{r, eval=TRUE, echo=FALSE}
dataset_vignette(adae, 
                 display_vars = vars(USUBJID, TRTSDT, TRTEDT, AESTDTC, ASTDT,
                                     TRTEMFL))
```

To derive on-treatment flag (`ONTRTFL`) in an ADaM dataset with a single assessment date, we
use `derive_var_ontrtfl()`.

The expected result is the input dataset with an additional column named `ONTRTFL` with a value of `"Y"` or `NA`.

```{r eval=TRUE}
bds1 <- tibble::tribble(
  ~USUBJID, ~ADT,              ~TRTSDT,           ~TRTEDT,
  "P01",    ymd("2020-02-24"), ymd("2020-01-01"), ymd("2020-03-01"),
  "P02",    ymd("2020-01-01"), ymd("2020-01-01"), ymd("2020-03-01"),
  "P03",    ymd("2019-12-31"), ymd("2020-01-01"), ymd("2020-03-01")
)
derive_var_ontrtfl(
  bds1,
  start_date = ADT,
  ref_start_date = TRTSDT,
  ref_end_date = TRTEDT
)

bds2 <- tibble::tribble(
  ~USUBJID, ~ADT,              ~TRTSDT,           ~TRTEDT,
  "P01",    ymd("2020-07-01"), ymd("2020-01-01"), ymd("2020-03-01"),
  "P02",    ymd("2020-04-30"), ymd("2020-01-01"), ymd("2020-03-01"),
  "P03",    ymd("2020-03-15"), ymd("2020-01-01"), ymd("2020-03-01")
)
derive_var_ontrtfl(
  bds2,
  start_date = ADT,
  ref_start_date = TRTSDT,
  ref_end_date = TRTEDT,
  ref_end_window = 60
)

bds3 <- tibble::tribble(
  ~ADTM,              ~TRTSDTM,           ~TRTEDTM,           ~TPT,
  "2020-01-02T12:00", "2020-01-01T12:00", "2020-03-01T12:00", NA,
  "2020-01-01T12:00", "2020-01-01T12:00", "2020-03-01T12:00", "PRE",
  "2019-12-31T12:00", "2020-01-01T12:00", "2020-03-01T12:00", NA
) %>%
 mutate(
  ADTM = ymd_hm(ADTM),
  TRTSDTM = ymd_hm(TRTSDTM),
  TRTEDTM = ymd_hm(TRTEDTM)
 )
derive_var_ontrtfl(
  bds3,
  start_date = ADTM,
  ref_start_date = TRTSDTM,
  ref_end_date = TRTEDTM,
  filter_pre_timepoint = TPT == "PRE"
)
```

## Derive Occurrence Flags {#occflag}

The function `derive_var_extreme_flag()` can help derive variables such as `AOCCIFL`, 
`AOCCPIFL`, `AOCCSIFL`, `AOCXIFL`, `AOCXPIFL`, and `AOCXSIFL`.

If grades were collected, the following can be used to flag first occurrence of
maximum toxicity grade.

```{r, eval=FALSE}
adae <- adae %>% 
  derive_var_extreme_flag(
    by_vars = vars(USUBJID),
    order = vars(desc(ATOXGR), ASTDTM, AESEQ),
    new_var = AOCCIFL,
    filter = TRTEMFL == "Y", 
    mode = "first"
  )
```

Similarly, `ASEV` can also be used to derive the occurrence flags if severity is
collected. In this case, the variable may need to be firstly recorded into a
numeric one. Flag first occurence of most severe adverse event:

```{r, eval=TRUE}
adae <- adae %>%
  mutate(
    ASEVN = as.integer(factor(ASEV, levels = c("MILD", "MODERATE", "SEVERE", "DEATH THREATENING")))
  ) %>% 
  derive_var_extreme_flag(
    by_vars = vars(USUBJID),
    order = vars(desc(ASEVN), ASTDTM, AESEQ),
    new_var = AOCCIFL,
    filter = TRTEMFL == "Y", 
    mode = "first"
  )
```

```{r, eval=TRUE, echo=FALSE}
dataset_vignette(adae, 
                 display_vars = vars(USUBJID, ASTDTM, ASEV, ASEVN, AESEQ, TRTEMFL,
                                     AOCCIFL))
```

## Derive Query Variables {#query}

For deriving query variables `SMQzzNAM`, `SMQzzCD`, `SMQzzSC`, `SMQzzSCN`, or
`CQzzNAM` the `derive_vars_query()` function can be used. As input it expects a
queries dataset, which provides the definition of the queries. See [Queries
dataset documentation](queries_dataset.html) for a detailed description ot the
queries dataset. The `create_query_data()` function can be used to create
queries datasets.

The following example shows how to derive query variables for Standardized MedDRA
Queries (SMQs) in ADAE.
```{r, eval=TRUE}
data("queries")
```

```{r, eval=TRUE, echo=FALSE}
dataset_vignette(queries)
```


```{r, eval=TRUE}
adae1 <- tibble::tribble(
  ~USUBJID, ~ASTDTM, ~AETERM, ~AESEQ, ~AEDECOD, ~AELLT, ~AELLTCD,
  "01", "2020-06-02 23:59:59", "ALANINE AMINOTRANSFERASE ABNORMAL",
    3, "Alanine aminotransferase abnormal", NA_character_, NA_integer_,
  "02", "2020-06-05 23:59:59", "BASEDOW'S DISEASE",
    5, "Basedow's disease", NA_character_, 1L,
  "03", "2020-06-07 23:59:59", "SOME TERM",
    2, "Some query", "Some term", NA_integer_,
  "05", "2020-06-09 23:59:59", "ALVEOLAR PROTEINOSIS",
    7, "Alveolar proteinosis", NA_character_,  NA_integer_
)
  
adae_query <- derive_vars_query(dataset = adae1 , dataset_queries = queries)
```

```{r, eval=TRUE, echo=FALSE}
dataset_vignette(adae_query)
```

Similarly to SMQ, the `derive_vars_query()` function can be used to derive
Standardized Drug Groupings (SDG).

```{r, eval=TRUE}
sdg <- tibble::tribble(
  ~VAR_PREFIX, ~QUERY_NAME,       ~SDG_ID, ~QUERY_SCOPE, ~QUERY_SCOPE_NUM, ~TERM_LEVEL, ~TERM_NAME,         ~TERM_ID,
  "SDG01",     "Diuretics",       11,      "BROAD",      1,                "CMDECOD",   "Diuretic 1",       NA,
  "SDG01",     "Diuretics",       11,      "BROAD",      2,                "CMDECOD",   "Diuretic 2",       NA,
  "SDG02",     "Costicosteroids", 12,      "BROAD",      1,                "CMDECOD",   "Costicosteroid 1", NA,
  "SDG02",     "Costicosteroids", 12,      "BROAD",      2,                "CMDECOD",   "Costicosteroid 2", NA,
  "SDG02",     "Costicosteroids", 12,      "BROAD",      2,                "CMDECOD",   "Costicosteroid 3", NA,
)          
adcm <- tibble::tribble(
  ~USUBJID, ~ASTDTM,               ~CMDECOD,
  "01",     "2020-06-02 23:59:59", "Diuretic 1",
  "02",     "2020-06-05 23:59:59", "Diuretic 1",
  "03",     "2020-06-07 23:59:59", "Costicosteroid 2",
  "05",     "2020-06-09 23:59:59", "Diuretic 2"
)
adcm_query <- derive_vars_query(adcm, sdg)
```

```{r, eval=TRUE, echo=FALSE}
dataset_vignette(adcm_query)
```

## Add the `ADSL` variables  {#adsl_vars}

If needed, the other `ADSL` variables can now be added:

```{r eval=TRUE, echo=TRUE}
adae <- adae %>%
  left_join(select(adsl, !!!admiral:::negate_vars(adsl_vars)),
            by = c("STUDYID", "USUBJID")
  )

```

```{r, eval=TRUE, echo=FALSE}
dataset_vignette(adae, 
                 display_vars = vars(USUBJID, AEDECOD, ASTDTM, DTHDT, RFSTDTC,
                                     RFENDTC, AGE, AGEU, SEX))
```

# Example Scripts

ADaM | Sample Code
---- | --------------
ADAE | [ad_adae.R](https://github.com/Roche-GSK/admiral/blob/main/inst/example_scripts/ad_adae.R){target="_blank"}
ADCM | [ad_adcm.R](https://github.com/Roche-GSK/admiral/blob/main/inst/example_scripts/ad_adcm.R){target="_blank"}
