---
title: "Best practices when writing vignette in admiral"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Best practices when writing vignette in admiral}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup0, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

This guidance ensures consistency across the vignettes in the admiral package in terms of content, structure and code execution. 
As a general rule, the vignette workflow defined in [r-pkgs.org](https://r-pkgs.org/vignettes.html) is followed.


# Metadata

Each vignette in admiral should start with the following metadata.

```
---
title: "<Title>"
output:  rmarkdown::html_vignette:
vignette: >
  %\VignetteIndexEntry{<Title>}
  %\VignetteEngine{knitr::rmarkdown}
---
```

The `<Title>` of the vignette should be meaningful. 

# Markdown

## Default options

    `r ''````{r setup, include=FALSE}
    knitr::opts_chunk$set(
      collapse = TRUE,
      comment = "#>"
      )
    ```

## Required Packages

    `r ''````{r, warning=FALSE, message=FALSE}
    library(admiral)
    #<all packages used in the vignette>
    ```
The `warning = FALSE` and `message = FALSE` options are there to prevent the usual messages:

<span style="color: red;font-family: textmate; font-size = 12pt;" >Attaching package: 'xxxx'</span>

<span style="color: red;font-family: textmate; font-size = 12pt;">The following objects are masked from 'package:yyyyy':</span>

<span style="color: red;font-family: textmate; font-size = 12pt;">  fun1, fun2</span>



## Format Sections

### Table of contents

Headings must start from Heading 1:

<span style="color: red;" > **Note: some vignettes starts the heading at H2** </span>

```
# Heading 1
## Heading 2

```

The first section (# Introduction) gives a brief introduction.

For vignettes describing an ADaM workflow, the second section will summarize the programming workflow.
The first sub-section within this workflow will describe the data read to demonstrate the use of the 
admiral functions, while the last sub-section should link to a template script. 

```
## Programming Workflow

* [Read in Data](#readdata)
* [Derive/Impute End and Start Analysis Date/time and Relative Day](#datetime)
* ...
* [Assign `ASEQ`](#aseq)
* [Example Script](#example)

```

### Conventions

+ Each sub-section within the programming workflow should be tagged (e.g. [Step1] (#step)), so that 
the user can go on the relevant section from the programming workflow (in addition to the Table of
contents). Don’t use a tag with a number but use a meaningful name (e.g. do not use `(#link1)`,
but use `(#this_action)`)

+ Any variable name, dataset name, function, argument name must be quoted with backticks: e.g. "`ADSL`, `dm`, `ADT`".

+ Functions must also end with `()`.

+ All the codes must be executed and the output result displayed once the code is executed. Use:
 

      `r ''````{r}
      #<code>
      ```


+ Any output created must clearly show what the function has derived. 
It should at least show the variables/records used as input to the function and the derived variables/records.

<span style="color: red;" > **Use of dataTable to show the result???** </span>


```{r, include=FALSE}
library(dplyr)
library(lubridate)
library(DT)
data(dm)
data(ds)
```

```{r}
adsl <- dm %>%
  derive_disposition_dt(
    dataset_ds = ds,
    new_var = EOSDT,
    dtc = DSSTDTC,
    filter = DSCAT == "DISPOSITION EVENT" & DSDECOD != "SCREEN FAILURE",
    date_imputation = NULL
  )

datatable(adsl %>% select(USUBJID, starts_with("EOS"),everything()),
          rownames = FALSE,
           filter = 'top',
           options = list(
             searchHighlight = TRUE,
             searching = FALSE,
             pageLength = 5,
             lengthMenu = c(5, 10, 15, 20, 50, 100)
  ))
#head(adsl %>% select(USUBJID, starts_with("EOS"), everything()))
```


