---
title: "Visit and Period Variables"
output:
  rmarkdown::html_vignette:
vignette: >
  %\VignetteIndexEntry{Visit and Period Variables}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

The derivation of visit variables like `AVISIT`, `AVISITN`, `AWLO`, `AWHI`, ...
or period, subperiod, or phase variables like `APERIOD`, `TRT01A`, `TRT02A`,
`ASPER`, `PHSDTM`, `PHEDTM`, ... is highly study-specific. Therefore admiral
cannot provide functions which derive these variables. However, for common
scenarios like visit assignments based on time windows or deriving BDS period
variables from ADSL period variables, functions are provided which support those
derivations.

## Required Packages

The examples of this vignette require the following packages.

```{r, warning=FALSE, message=FALSE}
library(admiral)
library(tibble)
library(dplyr, warn.conflicts = FALSE)
library(lubridate)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(admiraldev)
```

# Visit variables (`AVISIT`, `AVISITN`, `AWLO`, `AWHI`, ...) {#visits}

The most common ways of deriving `AVISIT` and `AVISITN` are:

- The variables are set to the collected visits (`VISIT` and `VISITNUM`).
- The variables are set based on time windows.

The former can be achieved simply by calling `mutate()`, like in the vignettes
and the template scripts.

For the latter a (study-specific) reference dataset needs to be created which
provides for each visit the start and end day (`AWLO` and `AWHI`) and the values
of other visit related variables (`AVISIT`, `AVISITN`, `AWTARGET`, ...).
```{r}
windows <- tribble(
  ~AVISIT,    ~AWLO, ~AWHI, ~AVISITN, ~AWTARGET,
  "BASELINE",   -30,     1,        0,         1,
  "WEEK 1",       2,     7,        1,         5,
  "WEEK 2",       8,    15,        2,        11,
  "WEEK 3",      16,    22,        3,        19,
  "WEEK 4",      23,    30,        4,        26
)
```
Then the visits can be assigned based on the analysis day (`ADY`) by calling
`derive_vars_joined()`:
```{r}
adbds <- tribble(
  ~USUBJID, ~ADY,
  "1",       -33,
  "1",        -2,
  "1",         3,
  "1",        24,
  "2",        NA,
)

derive_vars_joined(
  adbds,
  dataset_add = windows,
  filter_join = AWLO <= ADY & ADY <= AWHI
)
```

# Period, Subperiod, and Phase Variables

If periods, subperiods, or phases are used, the corresponding variables have to
be consistent across all datasets. This can be achieved by defining the periods,
subperiods, or phases once and then use this definition for all datasets. The
definition can be stored in ADSL or in a separate dataset. In the following
examples, this separate dataset is called period reference dataset.

## Period Reference Dataset

The period reference dataset contains one observation per subject and period,
subperiod, or phase. For example:

```{r echo=FALSE}
phase_ref <- tribble(
  ~USUBJID, ~APHASEN, ~PHSDT,       ~PHEDT,       ~APHASE,
  "1",             1, "2021-01-04", "2021-02-06", "TREATMENT",
  "1",             2, "2021-02-07", "2021-03-07", "FUP",
  "2",             1, "2021-02-02", "2021-03-02", "TREATMENT"
) %>%
  mutate(
    STUDYID = "xyz",
    APHASEN = as.integer(APHASEN),
    across(matches("PH[ES]DT"), ymd)
  ) %>%
  select(STUDYID, everything())

phase_ref
```

The admiral functions expect separate datasets for periods, subperiods, and
phases. For periods the numeric variable `APERIOD` is expected, for subperiods
the numeric variables `APERIOD` and `ASPER`, and for phases the numeric variable
`APHASEN`.

## Creating ADSL Period, Subperiod, or Phase Variables {#periods_adsl}

If a period reference dataset is available, the ADSL variables for periods,
subperiods, or phases can be created from this dataset by calling
`derive_vars_period()`.

For example the period reference dataset from the previous section can be used
to add the phase variables (`PHwSDT`, `PHwEDT`, and `APHASEw`) to ADSL:
```{r}
adsl <- tibble(STUDYID = "xyz", USUBJID = c("1", "2"))

adsl <- derive_vars_period(
  adsl,
  dataset_ref = phase_ref,
  new_vars = vars(PHwSDT = PHSDT, PHwEDT = PHEDT, APHASEw = APHASE)
) %>%
  select(STUDYID, USUBJID, PH1SDT, PH1EDT, PH2SDT, PH2EDT, APHASE1, APHASE2)

adsl
```

## Creating BDS and OCCDS Period, Subperiod, or Phase Variables {#periods_bds}

If a period reference dataset is available, BDS and OCCDS variables for periods,
subperiods, or phases can be created by calling `derive_vars_joined()`.

For example the variables `APHASEN`, `PHSDT`, `PHEDT`, `APHASE` can be derived
from the period reference dataset defined above.

```{r}
adae <- tribble(
  ~USUBJID, ~ASTDT,
  "1",      "2021-01-01",
  "1",      "2021-01-05",
  "1",      "2021-02-05",
  "1",      "2021-03-05",
  "1",      "2021-04-05",
  "2",      "2021-02-15",
) %>%
  mutate(ASTDT = ymd(ASTDT))

derive_vars_joined(
  adae,
  dataset_add = phase_ref,
  by_vars = vars(USUBJID),
  filter_join = PHSDT <= ASTDT & ASTDT <= PHEDT
)
```

If no period reference dataset is available but period variables are in ADSL,
the period reference dataset can be created from ADSL by calling
`create_period_dataset()`.

For example, a period reference dataset for phases can be created from the ADSL
dataset created above:
```{r}
create_period_dataset(
  adsl,
  new_vars = vars(PHSDT = PHwSDT, PHEDT = PHwEDT, APHASE = APHASEw)
)
```

# Treatment Variables (`TRTxxP`, `TRTxxA`, `TRTP`, `TRTA`, ...)

In studies with multiple periods the treatment can differ by period, e.g. for a
crossover trial. CDISC defines variables for planned and actual treatments in
ADSL (`TRTxxP`, `TRTxxA`, `TRxxPGy`, `TRxxAGy`, ...) and corresponding variables
in BDS and OCCDS datasets (`TRTP`, `TRTA`, `TRTPGy`, `TRTAGy`, ...). They can be
derived in the same way (and same step) as the period, subperiod, and phase
variables.

## Creating ADSL Treatment Variables {#treatment_adsl}

If the treatment information is included in the period reference dataset, the
treatment ADSL variables can be created by calling `derive_vars_period()`:

```{r}
# Add period variables to ADSL
period_ref <- tribble(
  ~USUBJID, ~APERIOD, ~APERSDT,     ~APEREDT,     ~TRTA,
  "1",             1, "2021-01-04", "2021-02-06", "DRUG A",
  "1",             2, "2021-02-07", "2021-03-07", "DRUG B",
  "2",             1, "2021-02-02", "2021-03-02", "DRUG B",
  "2",             2, "2021-03-03", "2021-04-01", "DRUG B"
) %>%
  mutate(
    STUDYID = "xyz",
    APERIOD = as.integer(APERIOD),
    across(ends_with("DT"), ymd)
  )

adsl <- derive_vars_period(
  adsl,
  dataset_ref = period_ref,
  new_vars = vars(
    APxxSDT = APERSDT,
    APxxEDT = APEREDT,
    TRTxxA = TRTA
  )
) %>%
  select(
    STUDYID, USUBJID,
    TRT01A, TRT02A,
    AP01SDT, AP01EDT, AP02SDT, AP02EDT
  )

adsl
```

## Creating BDS and OCCDS Treatment Variables {#treatment_bds}

If a period reference dataset is available, BDS and OCCDS variables for
treatment can be created by calling `derive_vars_joined()`.

For example the variables `APERIOD` and `TRTA` can be derived from the period
reference dataset defined above.

```{r}
adae <- tribble(
  ~USUBJID, ~ASTDT,
  "1",      "2021-01-05",
  "1",      "2021-02-05",
  "1",      "2021-03-05",
  "1",      "2021-04-05",
  "2",      "2021-02-15",
  "2",      "2021-03-10",
) %>%
  mutate(
    ASTDT = ymd(ASTDT),
    STUDYID = "xyz"
  )

derive_vars_joined(
  adae,
  dataset_add = period_ref,
  by_vars = vars(STUDYID, USUBJID),
  new_vars = vars(APERIOD, TRTA),
  join_vars = vars(APERSDT, APEREDT),
  filter_join = APERSDT <= ASTDT & ASTDT <= APEREDT
)
```

If no period reference dataset is available but period variables are in ADSL,
the period reference dataset can be created from ADSL by calling
`create_period_dataset()`.

For example, a period reference dataset for periods and treatments can be
created from the ADSL dataset created above:
```{r}
create_period_dataset(
  adsl,
  new_vars = vars(APERSDT = APxxSDT, APEREDT = APxxEDT, TRTA = TRTxxA)
)
```
