---
title: "Writing Vignettes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Writing Vignettes}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

This guidance ensures consistency across the vignettes in the `{admiral}` package in terms of content, structure and code execution. 
As a general rule, the vignette workflow defined in [r-pkgs.org](https://r-pkgs.org/vignettes.html) is followed.


# Metadata

Each vignette in `{admiral}` should start with the following metadata.

```
---
title: "<Title>"
output:  rmarkdown::html_vignette:
vignette: >
  %\VignetteIndexEntry{<Title>}
  %\VignetteEngine{knitr::rmarkdown}
---
```

The `<Title>` of the vignette should be meaningful. 

# Markdown

## Default Options

If any chunks are used within the vignette, the following options should be set after the metadata 
to ensure that the chunks are rendered consistently across all vignettes.

    `r ''````{r setup, include=FALSE}
    knitr::opts_chunk$set(
      collapse = TRUE,
      comment = "#>"
    )
    ```


## Format Sections

### Table of Contents

Headings must be title case and start from Heading 1:

```
# Heading 1
This is the description of my first section.

## Heading 1.1
This is the description of my first sub-section within my first section.

## Heading 1.2
This is the description of my second sub-section within my first section.

```

The first section gives a brief introduction of the vignette. The last sub-section 
of the introduction should describe the packages required to run the `{admiral}` functions. 
The `{admiral}` package should be described first.

```
# Introduction

This is the introduction of my vignette.

## Required Packages
```

    `r ''````{r, warning=FALSE, message=FALSE}
    library(admiral)
    # <all packages used in the vignette>
    ```

The `warning=FALSE` and `message=FALSE` options are there to prevent the usual messages:

<span style="color: red;font-family: textmate, monospace; font-size = 10pt;" >Attaching package: 'xxxx'</br>The following objects are masked from 'package:yyyyy'</br>  fun1, fun2</span>

### Conventions

#### General Conventions

+ Any new vignette must be added to the `_pkgdown.yml` file in the relevant section.

+ Any variable name, dataset name, function, argument name must be quoted with backticks: e.g.


```
  The `date` parameter of the `derive_my_var()` function expects a date variable, e.g., `ADT`.
```

+ Functions must also end with `()`.

+ Variables and datasets name are expected to be uppercase.

+ All the codes must be described, executed and the output result displayed once the code is executed.
Use:
 

      `r ''````{r}
      #<code>
      ```

+ Any output created must clearly show what the function has derived. 
It should at least show the variables/records used as input to the function and the derived 
variables/records. If a dataset must be displayed, it will be formatted using the `dataset_vignette()`
function so that it is displayed consistently across the vignettes.E.g.


  + Description and execution of the code used to derive the variable/record
  


```{r, include=FALSE}
library(lubridate)
library(dplyr)
library(admiral.test)
library(DT)

data(admiral_vs)
```

```{r, eval=FALSE, echo=TRUE}
vs1 <- admiral_vs %>%
  derive_vars_dt(
    new_vars_prefix = "A",
    dtc = VSDTC,
    date_imputation = "first"
  )
```


  + Output dataset formatted using `dataset_vignette()`...

#TODO - Fix this

```{r, eval=FALSE, echo=FALSE}
vs1 <- admiral_vs %>%
  derive_vars_dt(
    new_vars_prefix = "A",
    dtc = VSDTC,
    date_imputation = "first"
  )

dataset_vignette(
  vs1,
  display_vars = exprs(USUBJID, VSTESTCD, VISIT, VSDTC, ADT),
  filter = VSTESTCD == "WEIGHT"
)
```


Note: The call to get the formatted dataset would be: 
 

      ```{r, eval=FALSE, echo=TRUE}
dataset_vignette(
  vs1,
  display_vars = exprs(USUBJID, VSTESTCD, VISIT, VSDTC, ADT),
  filter = VSTESTCD == "WEIGHT"
)
      ```
      
Displaying many big datasets on a vignette, may require a long time to load
the page and may cause messages from the browser that the page is not
responsive. In this case the number of displayed observations should be
restricted either by restricting the source datasets at the beginning of the
vignette or in the call of `dataset_vignette()` if only some calls are
affected.

#### Conventions for ADaM Workflow

For vignettes describing an ADaM workflow, 

+ the second section will summarize the programming workflow. The first sub-section within this 
workflow will always describe the data read to demonstrate the use of the `{admiral}` functions,

+ Each sub-section within the programming workflow should be tagged (e.g. [Step1] (#step)), so that 
the user can go on the relevant section from the programming workflow (in addition to the Table of
contents). Don’t use a tag with a number but use a meaningful name (e.g. do not use `(#link1)`,
but use `(#this_action)`)

+ the last section should link to a template script. 


```
# Programming Workflow

* [Read in Data](#readdata)
* [Derive/Impute End and Start Analysis Date/time and Relative Day](#datetime)
* ...
* [Assign `ASEQ`](#aseq)

## Read in Data {#readdata}
## Derive/Impute End and Start Analysis Date/time and Relative Day {#datetime}
## ...
## Assign `ASEQ` {#aseq}

# Another Section

# Example Script
ADaM | Sample Code
---- | --------------
ADxx | [ad_adxx.R](https://github.com/pharmaverse/admiral/blob/main/inst/templates/ad_adxx.R){target="_blank"}

```

+ `ADSL` variables

  All `ADSL` variables which are required for any derivation are merged to the SDTM dataset before the
  first derivation. 
  These `ADSL` variables have to be added to the by-variables for derivations which add observations. 
  This ensures that the `ADSL` variables are populated `for` the new observations. 
  A `adsl_vars` variable should be created at the beginning of the script and added to the `by_vars` 
  parameter for derivations which add observations.

  `ADSL` variables which should be in the final dataset but are not required for any derivation are
  merged to the dataset after the last derivation.



